### 2014-10-15 paper_v1.Rnw
Adding first draft of paper.

\documentclass{sigchi}

% Use this command to override the default ACM copyright statement (e.g. for preprints). 
% Consult the conference website for the camera-ready copyright statement.

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
\toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
{\emph{L@S'15}}, March 14--15, 2015, Vancouver, Canada. \\
Copyright \copyright~2015 ACM ISBN/14/04...\$15.00. \\
DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

% Arabic page numbers for submission. 
% Remove this line to eliminate page numbers for the camera ready copy
\pagenumbering{arabic}

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage{url}      % llt: nicely formatted URLs

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{\def\UrlFont{\sf}}{\def\UrlFont{\small\bf\ttfamily}}}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, 
% to give it a fighting chance of not being over-written, 
% since its job is to redefine many LaTeX commands.
\usepackage[pdftex]{hyperref}
\hypersetup{
pdftitle={SIGCHI Conference Proceedings Format},
pdfauthor={LaTeX},
pdfkeywords={SIGCHI, proceedings, archival format},
bookmarksnumbered,
pdfstartview={FitH},
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black,
breaklinks=true,
}

% create a shortcut to typeset table headings
\newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses}

\numberofauthors{2}
\author{
  \alignauthor Omitted for blind review\\
  \affaddr{Institution}\\
  \affaddr{Address}\\
  \email{Email}\\
%   \alignauthor Sherif Halawa\\
%     \affaddr{Department of Electrical Engineering}\\
%     \affaddr{Stanford University}\\
%     \email{halawa@stanford.edu}\\
%   \alignauthor Ren\'{e} Kizilcec\\
%     \affaddr{Department of Communication}\\
%     \affaddr{Stanford University}\\
%     \email{kizilcec@stanford.edu}\\
}

\maketitle

\begin{abstract}
Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from [N] online learners in [M] courses. Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. Study 2 is a case study to develop a deeper understanding of attrition in MOOCs by conducting a case study. Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey. 
\end{abstract}

\keywords{Online learning, persistence, attrition, dropout, disengagement, massive open online courses, MOOC, psychological factors}

\category{H.5.m.}{Information Interfaces and Presentation (e.g. HCI)}{Miscellaneous}
\category{K.3.1.}{Computers and Education}{Computer Uses in Education}

\section{Introduction}

Educational environments has become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accomodating people with less flexibile schedules \cite{goldrick2010challenges}. Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (``correspondence education''), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of poeple invovled (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} (other?) resemble those on online media platforms, such as YouTube or tumblr. This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into sucesses and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs} (other?), motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one world is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other world adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition.

This paper presents a systematic investigation into attrition in MOOCs, based on self-report and behavioral data collected from [N] online learners in [M] courses. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, we sought to develop a deeper understanding of attrition in MOOCs by conducting a case study. Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey. 
TODO: complete paragraph based on what we actually do in Study 2.

\section{Related Work}

Research and theorizing on attrition in education has a rich history. This review is intended to serve as a foundation to build on with the current research. The focus of this breif review is on how ways of thinking about attrition have developed over the last decades.

\subsection{In-person Education}

The majority of early work on attrition centered around theoretical models of students' decision to persist or dopout of a traditional higher education setting. An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, adn norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}. Hence, motivation alone is necessary but not sufficient for persistence. Students may fail to sustain efforts in the absence of strong self-regulatory skills.

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model

\subsection{Distance Education and e-Learning}

\subsection{MOOCs}

\section{Study 1: Reasons for Disengaging}

\subsection{Methods}

TODO: Outline different courses and go through variabels in the course info table.

Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate

\subsection{Results}

Table: proportions for each dropout reason for each course

Table: correlations between reasons and one with cors between reasons and intentions

Plot: ecdf curves for proportion videos watched for each disengagement reason

\subsection{Discussion}

\section{Study 2: Understanding Disengagement}

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller but more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. Identifying learners who were likely to drop out was achieved with machine learning.

\subsection{Methods}

The particular MOOC under observation was an undergraduate level course on an advanced topic in computer science. It was offered in 2014 through Coursera. There were [N1] enrolled learners; [N2] watched more than one video, and [N3] attempted more than one assignment.

\subsubsection{Predicting Disengagement}

TODO: Details of the prediction algorithm.

\subsubsection{Feedback Survey}

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A very low response rate was expected, given that this subpopulation was defined by low engagement. 535 out of 6,050 learners started the survey (8.8\% response rate), and 499 completed it (8.2\% completion rate).

Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked. In addition, they reported intentions for engaging with each week's course content. Then, they were asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of. This question was deliberately asked prior to any survey questions that could suggest particular reasons for disengagement.

Two research assistants independently developed a codebook for the resulting 435 non-empty open responses. Their codebooks were consolidated and applied on Mechanical Turk, where 4 `classification experts' independently coded all 200 randomly selected responses. Specific updates to the initial codebook were informed by category frequency, category correlations, and inter-coder agreement. The updated codebook was then applied by 4 Mechanical Turk `classification experts' to the 235 remaining open responses.

Following the open resposne question, learners reported the difficulty of the course material and the extent to which their progress was hindred by a number of obstacles (see Table [XX]). The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.67, SD=0.72, \alpha=0.87$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=2.45, SD=0.90, \alpha=0.78$), and goal striving (5 items on motivation, perceived importance, committment, confidence, and distractions; $M=3.08, SD=0.83, \alpha=0.76$).

TODO: update numbers once finalized, explain use of imputation

\subsection{Results}

TODO: Add results of prediction. How well did the prediction work? And who took the survey; how sure was the model that those who took the survey would actually drop out, relative to those who didn't take the survey?

TODO: Write up results of diagnostic survey, keeping it simple for now.

\subsection{Discussion}

\section{General Discussion}

\section{Conclusion}

\section{Acknowledgments}
% We thank Elise Ogle and Ruth Bram for helping with the development of the codebook. 
Omitted for blind review.

% Balancing columns in a ref list is a bit of a pain because you
% either use a hack like flushend or balance, or manually insert
% a column break.  http://www.tex.ac.uk/cgi-bin/texfaq2html?label=balance
% multicols doesn't work because we're already in two-column mode,
% and flushend isn't awesome, so I choose balance.  See this
% for more info: http://cs.brown.edu/system/software/latex/doc/balance.pdf
%
% Note that in a perfect world balance wants to be in the first
% column of the last page.
%
% If balance doesn't work for you, you can remove that and
% hard-code a column break into the bbl file right before you
% submit:
%
% http://stackoverflow.com/questions/2149854/how-to-manually-equalize-columns-
% in-an-ieee-paper-if-using-bibtex
%
% Or, just remove \balance and give up on balancing the last page.
%
\balance

\bibliographystyle{acm-sigchi}
\bibliography{dropout.bib}
\end{document}

### 2014-10-16 paper_v1.Rnw
added some research questions and hypotheses

@@
 
+TODO: develop
+
+\begin{description}
+  \item[RQ1] Why do people who drop out report they drop out?
+  \item[RQ2] Who is most likely to drop out in terms of demographics, intentions/motivations, geo location, prior experience with online courses?
+  \item[RQ3] What reasons do people from different subgroups (demographic, intention, etc.) report for dropping out?
+  \item[RQ4] What proportion of people who dropped out were satisfied with what they got out of the course?
+  \item[RQ5] What behaviors do people with different dropout reasons exhibit?
+  \item[H1] prior experience -- technical difficulties
+  \item[H2] demographics (education) -- course difficulty
+\end{description}
+
 \subsection{Methods}
@@
 Identifying learners who were likely to drop out was achieved with machine learning.
+
+TODO: develop
+
+H1: People who report `not enough time' can be split into those who have external constraints and those who have low volition.
+
+H2: people who report external committments (family, school), are more likely to specify the reason than people who have `no time' because they have low volition
 

### 2014-10-17 paper_v1.Rnw
extended the lit. review

@@
 The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution.
-Tinto's \citeyear{tinto1975dropout} student integration model
+Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college.
+While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
+Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
+This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}.
+
+Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration.
+Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.
 
 \subsection{Distance Education and e-Learning}
+
+Dinstance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
+However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner.
+Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
+Persistence is thought to be a function of background characteristics (e.g., demographics), adademic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
+The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
+
+Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs.
+Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning.
+And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
+An exploratory study of reasons for attrition suggested the following eight constructs based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
 

### 2014-10-17 paper_v1.tex
Adding first draft of paper.
extended the lit. review

--- 2014-10-15 paper_v1.Rnw
@@
-\documentclass{sigchi}
+\documentclass{sigchi}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 % End of preamble. Here it comes the document.
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution.
-Tinto's \citeyear{tinto1975dropout} student integration model
+Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college.
+While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
+Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
+This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}.
+Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration.
+Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.
 
 \subsection{Distance Education and e-Learning}
+
+Dinstance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
+However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner.
+Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
+Persistence is thought to be a function of background characteristics (e.g., demographics), adademic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
+The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
+
+Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs.
+Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning.
+And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
+An exploratory study of reasons for attrition suggested the following eight constructs based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
 
@@
 \section{Study 1: Reasons for Disengaging}
+
+TODO: develop
+
+\begin{description}
+  \item[RQ1] Why do people who drop out report they drop out?
+  \item[RQ2] Who is most likely to drop out in terms of demographics, intentions/motivations, geo location, prior experience with online courses?
+  \item[RQ3] What reasons do people from different subgroups (demographic, intention, etc.) report for dropping out?
+  \item[RQ4] What proportion of people who dropped out were satisfied with what they got out of the course?
+  \item[RQ5] What behaviors do people with different dropout reasons exhibit?
+  \item[H1] prior experience -- technical difficulties
+  \item[H2] demographics (education) -- course difficulty
+\end{description}
 

### 2014-10-17 paper_v1.Rnw
more work on lit. review

@@
 And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
-An exploratory study of reasons for attrition suggested the following eight constructs based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
+An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
+An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.
 
 \subsection{MOOCs}
+
+A survey of over one hundred learners who dropped out of MOOC showed that most learners indicated too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}.
+A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}.
+A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that i.
+the likelihood of dropout was lower for those who actively participated in the first week of the course; ii.
+those who served as an authority figure in the community on the forum, and iii.
+those who did not engage in a particular subcommunity on the forum \cite{yang2013turn}.
+In addition, a number of machine learning approaches to predicting attrition yielded promising results (e.g.
+\cite{taylor2014likely,halawa2014dropout}), but fell short of offering insight into why a given learner left the course.
 

### 2014-10-17 paper_v1.tex
more work on lit. review

@@
 This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}.
+
 Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration.
@@
 And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
-An exploratory study of reasons for attrition suggested the following eight constructs based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
+An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
+An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.
 
 \subsection{MOOCs}
+
+A survey of over one hundred learners who dropped out of MOOC showed that most learners indicated too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}.
+A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}.
+Based on just over one thousand learners who participated in the discussion forum of a particular MOOC, the likelihood of dropout was lower for those who actively participated in the first week of the course, those who served as an authority figure in the community on the forum, and those who did not engage in a particular subcommunity on the forum \cite{rose2014social}.
+
+A number of machine learning approaches to predicting attrition yielded promising results, e.g.
+\cite{taylor2014likely,halawa2014dropout}, but fell short of offering insight into why a given learner left the course.
 

### 2014-10-17 paper_v1.Rnw
more work on lit. review

@@
 A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}.
-A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that i.
-the likelihood of dropout was lower for those who actively participated in the first week of the course; ii.
-those who served as an authority figure in the community on the forum, and iii.
-those who did not engage in a particular subcommunity on the forum \cite{yang2013turn}.
-In addition, a number of machine learning approaches to predicting attrition yielded promising results (e.g.
-\cite{taylor2014likely,halawa2014dropout}), but fell short of offering insight into why a given learner left the course.
+Based on just over one thousand learners who participated in the discussion forum of a particular MOOC, the likelihood of dropout was lower for those who actively participated in the first week of the course, those who served as an authority figure in the community on the forum, and those who did not engage in a particular subcommunity on the forum \cite{rose2014social}.
+
+A number of machine learning approaches to predicting attrition have been tested, incluging hidden Markov models \cite{balakrishnan2013predicting},
 

### 2014-10-17 paper_v1.tex
more work on lit. review

@@
 
-A number of machine learning approaches to predicting attrition yielded promising results, e.g.
-\cite{taylor2014likely,halawa2014dropout}, but fell short of offering insight into why a given learner left the course.
+A number of machine learning approaches to predicting attrition have been tested, incluging hidden Markov models \cite{balakrishnan2013predicting},
 

### 2014-10-17 paper_v1.Rnw
integrating RQs and Hs into lit. review

@@
 This review is intended to serve as a foundation to build on with the current research.
+We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work.
 The focus of this breif review is on how ways of thinking about attrition have developed over the last decades.
@@
 
+\begin{description}
+  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC?
+  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
+\end{description}
+
 The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution.
@@
 Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
-This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}.
+This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2007question}.
+
+\begin{description}
+  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
+\end{description}
 
@@
 Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.
+Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (i.e., instilling a growth mindset) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}.
+The results highlight the powerful influence of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.
+
+\begin{description}
+  \item[RQ2] What proportion of learners who dropped out were satisfied with their achievements in the course?
+  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
+\end{description}
 
@@
 
-A survey of over one hundred learners who dropped out of MOOC showed that most learners indicated too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}.
+A survey of over one hundred learners who dropped out of a MOOC showed that the majority indicated having too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}.
 A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}.
-Based on just over one thousand learners who participated in the discussion forum of a particular MOOC, the likelihood of dropout was lower for those who actively participated in the first week of the course, those who served as an authority figure in the community on the forum, and those who did not engage in a particular subcommunity on the forum \cite{rose2014social}.
+A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that the likelihood of dropout was lower for those who i.
+actively participated in the first week of the course; ii.
+served as an authority figure in the community on the forum, and iii.
+did not engage in a particular subcommunity on the forum \cite{yang2013turn}.
+In addition, a number of machine learning approaches yielded promising results for early dropout prediction (e.g.
+\cite{taylor2014likely,halawa2014dropout}), but fell short of offering insights into why a given learner would leave the course.
 
-A number of machine learning approaches to predicting attrition have been tested, incluging hidden Markov models \cite{balakrishnan2013predicting},
+\begin{description}
+\item[RQ3a] What reasons do learners report for dropping out of a MOOC?
+\item[RQ3b] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)?
+\item[RQ4] How does behavior in MOOCs vary by reasons for dropping out?
+\end{description}
 
@@
 
-TODO: develop
-
-\begin{description}
-  \item[RQ1] Why do people who drop out report they drop out?
-  \item[RQ2] Who is most likely to drop out in terms of demographics, intentions/motivations, geo location, prior experience with online courses?
-  \item[RQ3] What reasons do people from different subgroups (demographic, intention, etc.) report for dropping out?
-  \item[RQ4] What proportion of people who dropped out were satisfied with what they got out of the course?
-  \item[RQ5] What behaviors do people with different dropout reasons exhibit?
-  \item[H1] prior experience -- technical difficulties
-  \item[H2] demographics (education) -- course difficulty
-\end{description}
+% \begin{description}
+%   \item[H1] prior experience -- technical difficulties
+%   \item[H2] demographics (education) -- course difficulty
+% \end{description}
 
@@
 
-H2: people who report external committments (family, school), are more likely to specify the reason than people who have `no time' because they have low volition
+H2: people who report external committments (family, school), are more likely to specify the reason than people who have `no time' because they have low volition.
 

### 2014-10-17 paper_v1.Rnw
spell check and minor edits

@@
 %% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-\toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+\toappear{%$^\ast$ All authors contributed equally. \\
+Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
 Copyrights for components of this work owned by others than ACM must be honored.
@@
   \email{Email}\\
-%   \alignauthor Sherif Halawa\\
+%   \alignauthor Ren\'{e} F. Kizilcec$^\ast$\\
+%     \affaddr{Department of Communication}\\
+%     \affaddr{Stanford University}\\
+%     \email{kizilcec@stanford.edu}\\
+%   \alignauthor Sherif Halawa$^\ast$\\
 %     \affaddr{Department of Electrical Engineering}\\
@@
 %     \email{halawa@stanford.edu}\\
-%   \alignauthor Ren\'{e} Kizilcec\\
-%     \affaddr{Department of Communication}\\
-%     \affaddr{Stanford University}\\
-%     \email{kizilcec@stanford.edu}\\
 }
@@
 Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey.
+(abstract needs work...)
 \end{abstract}
@@
 Novel institutional structures have been developed to overcome particular constraints.
-Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accomodating people with less flexibile schedules \cite{goldrick2010challenges}.
+Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accommodating people with less flexible schedules \cite{goldrick2010challenges}.
 Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes.
@@
 By design, MOOCs provide course materials to millions of people worldwide.
-This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of poeple invovled (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}).
+This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}).
 Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon.
@@
 Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}.
-Early MOOC research cautioned against dichotomizing learners into sucesses and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}.
+Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}.
 Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs} (other?), motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested.
@@
 In Study 2, we sought to develop a deeper understanding of attrition in MOOCs by conducting a case study.
-Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey.
+Targeting around 7,000 online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey.
 TODO: complete paragraph based on what we actually do in Study 2.
@@
 We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work.
-The focus of this breif review is on how ways of thinking about attrition have developed over the last decades.
+The focus of this brief review is on how ways of thinking about attrition have developed over the last decades.
 
@@
 
-The majority of early work on attrition centered around theoretical models of students' decision to persist or dopout of a traditional higher education setting.
-An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, adn norms \cite{fishbein1975belief}.
+The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting.
+An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
 The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}.
@@
 Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college.
-While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
+While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
 Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
@@
 
-Dinstance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
+Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
 However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner.
 Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
-Persistence is thought to be a function of background characteristics (e.g., demographics), adademic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
+Persistence is thought to be a function of background characteristics (e.g., demographics), academic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
 The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
@@
 
-TODO: develop
-
-H1: People who report `not enough time' can be split into those who have external constraints and those who have low volition.
-
-H2: people who report external committments (family, school), are more likely to specify the reason than people who have `no time' because they have low volition.
+\begin{description}
+  \item[H4a] Learners who report a `lack of time' can be divided into those with too many higher priority obligations (work, school, family) and those with low volition.
+  \item[H4b] Learners who do not report specific obligations (e.g., work, school, family) tend to fall in the low volition than in the high-priority obligations group, and vice versa.
+\end{description}
 
@@
 
-Following the open resposne question, learners reported the difficulty of the course material and the extent to which their progress was hindred by a number of obstacles (see Table [XX]).
+Following the open response question, learners reported the difficulty of the course material and the extent to which their progress was hindered by a number of obstacles (see Table [XX]).
 The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.67, SD=0.72, \alpha=0.87$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=2.45, SD=0.90, \alpha=0.78$), and goal striving (5 items on motivation, perceived importance, committment, confidence, and distractions; $M=3.08, SD=0.83, \alpha=0.76$).

### 2014-10-20 paper_v1.Rnw
added some diagnostic results

@@
 \usepackage{url}      % llt: nicely formatted URLs
+\usepackage{booktabs}
 
@@
 
-TODO: Details of the prediction algorithm.
+SHERIF TODO: Details of the prediction algorithm.
 
@@
 Could you help us understand why?'' A very low response rate was expected, given that this subpopulation was defined by low engagement.
-535 out of 6,050 learners started the survey (8.8\% response rate), and 499 completed it (8.2\% completion rate).
+756 out of 6,050 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
+Missing values were multiply imputated using chained equations based on responses to all survey questions\footnote{reference to MICE}.
+All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
+85\% of respondents were male, 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.
 
@@
 
-Two research assistants independently developed a codebook for the resulting 435 non-empty open responses.
-Their codebooks were consolidated and applied on Mechanical Turk, where 4 `classification experts' independently coded all 200 randomly selected responses.
-Specific updates to the initial codebook were informed by category frequency, category correlations, and inter-coder agreement.
-The updated codebook was then applied by 4 Mechanical Turk `classification experts' to the 235 remaining open responses.
+Two research assistants independently developed a codebook for the resulting 448 non-empty open responses.
+Their codebooks were consolidated and applied on Mechanical Turk, where 250 randomly selected responses were each coded by four `classification experts'.
+Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement.
+The updated codebook was then applied by Mechanical Turk `classification experts' to the 198 remaining open responses (four coders per response).
 
-Following the open response question, learners reported the difficulty of the course material and the extent to which their progress was hindered by a number of obstacles (see Table [XX]).
-The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.67, SD=0.72, \alpha=0.87$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=2.45, SD=0.90, \alpha=0.78$), and goal striving (5 items on motivation, perceived importance, committment, confidence, and distractions; $M=3.08, SD=0.83, \alpha=0.76$).
-
-TODO: update numbers once finalized, explain use of imputation
+Besides these open responses, learners reported the difficulty of the course material and the extent to which their progress was hindered by a number of obstacles (see Table [XX]).
+The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and one item on volition (about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
 
@@
 
-TODO: Add results of prediction.
+\subsubsection{Predicting Likely Dropouts}
+
+SHERIF TODO: Add results of prediction.
 How well did the prediction work?
@@
 
-TODO: Write up results of diagnostic survey, keeping it simple for now.
+\subsubsection{Reasons for Dropout}
+
+Reasons for learner attrition were measured using open response as well as multiple-choice questions.
+As we were interested in learners who left the course unsatisfied, we focused on the subpopulation that reported being only somewhat or less satisfied with their progress in the course (73\% of respondents).
+Almost half (47\%) the respondents found the course somewhat, very, or extremely difficult.
+When asked about the extent their progress was hindered by certain challenges, 18\% reported problems with computer or Internet access, 8\% reported technical difficulties, 17\% had difficulties with their level of English, and 24\% found the course requirements unclear.
+While two in three (66\%) respondents were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
+volition).
+The two measures, which only differ in the specified level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.
+
+\subsubsection{Psychological Factors}
+
+Hypotheses H1, H2, and H3 about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
+Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
+The two measures were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
+Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
+Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
+Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress.
+The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction.
+The results fully support $H1$, and partially support $H2$ and $H3$.
+
+\begin{table*}[ht]
+\caption{Descriptive and inferrential statistics for psychological measures by learner success}
+\label{tab:psych}
+\small
+\center
+\begin{tabular}{lcccc}
+\toprule
+ & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
+\midrule
+\emph{Relative Progress} &  &  &  \\
+\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
+\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
+ &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
+ \emph{Satisfaction} &  &  &  \\
+\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
+\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
+&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
+\bottomrule
+% \multicolumn{4}{l}{Two-sided $t$-test: $^{***} < .001$, $^{**} < .01$, $^{*} < .05$} \\
+\end{tabular}
+\end{table*}
 

### 2014-10-25 paper_v1.Rnw
Added open response analysis to study 2.

@@
 
-To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller but more focused follow-up study.
+To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study.
 In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback.
-Identifying learners who were likely to drop out was achieved with machine learning.
+We developed a predictive model to identify learners who were likely to drop out and ask them to complete a survey.
 
 \begin{description}
-  \item[H4a] Learners who report a `lack of time' can be divided into those with too many higher priority obligations (work, school, family) and those with low volition.
+  \item[H4a] Learners who report having `not enough time' can be divided into those with too many higher priority obligations (work, school, family) and those with low volition.
   \item[H4b] Learners who do not report specific obligations (e.g., work, school, family) tend to fall in the low volition than in the high-priority obligations group, and vice versa.
@@
 Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently.
-Could you help us understand why?'' A very low response rate was expected, given that this subpopulation was defined by low engagement.
-756 out of 6,050 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
-Missing values were multiply imputated using chained equations based on responses to all survey questions\footnote{reference to MICE}.
+Could you help us understand why?'' A low response rate was expected for a subpopulation that was defined by low engagement.
+And yet, 756 out of 6,050 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
+Missing values were multiply imputated by predictive mean matching using responses to all survey questions\footnote{The $R$ mice package was used to perform multiple imputation and subsequent analyses.}.
 All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
-85\% of respondents were male, 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.
+The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.
 
@@
 
-Two research assistants independently developed a codebook for the resulting 448 non-empty open responses.
-Their codebooks were consolidated and applied on Mechanical Turk, where 250 randomly selected responses were each coded by four `classification experts'.
+Two research assistants independently developed codebooks for the resulting 448 non-empty open responses.
+Their codebooks were consolidated and applied on Mechanical Turk in three iterations.
+The codebook was updated in each iteration to reliably fit the open response data.
 Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement.
-The updated codebook was then applied by Mechanical Turk `classification experts' to the 198 remaining open responses (four coders per response).
+In the first iteration, 250 randomly selected responses were each coded by four `classification experts'.
+In the second iteration, the remaining responses were coded in the same way.
+In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see [URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
 
-Besides these open responses, learners reported the difficulty of the course material and the extent to which their progress was hindered by a number of obstacles (see Table [XX]).
+Besides these open responses, learners rated their relative progress and satisfaction with their progress in the course, which served as measures of personal success.
+They also rated the extent to which their progress was hindered by a number of obstacles.
 The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and one item on volition (about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
@@
 
+The majority of survey respondents (73\%; 550 of 756) reported being only somewhat or less satisfied with their progress in the course.
+While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
+volition).
+These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.
+
 Reasons for learner attrition were measured using open response as well as multiple-choice questions.
-As we were interested in learners who left the course unsatisfied, we focused on the subpopulation that reported being only somewhat or less satisfied with their progress in the course (73\% of respondents).
-Almost half (47\%) the respondents found the course somewhat, very, or extremely difficult.
-When asked about the extent their progress was hindered by certain challenges, 18\% reported problems with computer or Internet access, 8\% reported technical difficulties, 17\% had difficulties with their level of English, and 24\% found the course requirements unclear.
-While two in three (66\%) respondents were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
-volition).
-The two measures, which only differ in the specified level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.
+Based on learners' coded open responses ($n=399$), 84\% had `not enough time for the course', 18\% indicated that `the course format or teaching style was not a good fit', 10\% expressed that `the level of the course was too advanced', 7\% found that `the course was not interesting or valuable enough', 4\% expressed a need for `additional course materials to learn the topic', 2\% experienced `technical difficulties', and 1\% reported facing a `language barrier'.
+A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).
+
+To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time, which revealed that about half of them (53\%) did not specify a particular committment.
+This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders.
+Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
+We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volition, i.e.
+their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
+However, the two groups did not significantly differ in how much they were held back by committments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
+These findings lend strong support to hypotheses 4a and 4b: while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volition and reporting no specific reasons for having not enough time was indicative thereof.
 
@@
 
-Hypotheses H1, H2, and H3 about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
+Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
 Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
@@
 \bottomrule
-% \multicolumn{4}{l}{Two-sided $t$-test: $^{***} < .001$, $^{**} < .01$, $^{*} < .05$} \\
 \end{tabular}

### 2014-10-26 paper_v1.Rnw
added intro to study 2

@@
 In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback.
-We developed a predictive model to identify learners who were likely to drop out and ask them to complete a survey.
+To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey.
+One goal of this study was to further investiage research question $RQ3a$ based on learners' own descriptions of the challenges they faced.
+Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time.
+A variety of factors can influence the amount of time learners have at their disposal.
+Higher priority obligations, such as professional, academic, and personal committments, are certainly large contributors.
+Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}.
+For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV).
+We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:
 
 \begin{description}
-  \item[H4a] Learners who report having `not enough time' can be divided into those with too many higher priority obligations (work, school, family) and those with low volition.
-  \item[H4b] Learners who do not report specific obligations (e.g., work, school, family) tend to fall in the low volition than in the high-priority obligations group, and vice versa.
+  \item[H4a] Learners who report having `not enough time' can be grouped into those with higher priority obligations (professional, academic, personal) and those with low volitional control.
+  \item[H4b] Learners who do not report specific higher-priority obligations tend to exhibit lower volitional control than those who specify particular obligations.
 \end{description}
+
+Another goal of Study 2 was to collect data to test hypotheses $H1, H2, H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
+To this end, we included a self-report measures for each construct in the survey that was sent out to at-risk learners.
 

### 2014-10-26 paper_v1.Rnw
updating prediction section

@@
 It was offered in 2014 through Coursera.
-There were [N1] enrolled learners; [N2] watched more than one video, and [N3] attempted more than one assignment.
+There were  20,048 enrolled learners; 10,510 watched at least one video, and [N] attempted more than one assignment.
 
@@
 
-SHERIF TODO: Details of the prediction algorithm.
+The objective was to build an `early warning' prediction model that could flag learners who are likely to drop out in the future.
+We defined the following three criteria by which to evaluate the performance of the model.
+
+\textit{1.
+Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible.
+The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.
+
+\textit{2.
+Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short.
+A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.
+
+\textit{3.
+Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course.
+We thus had to train the prediction model using data from previous courses.
+The performance of the prediction model should be adequate in courses other than those on which it was trained.
+
+The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs.
+The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses.
+Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the test set ($test_1$).
+Another second test set of similar size was composed from 20 other MOOCs ($test_2$).
+Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).
+The model was fit by logistic regression.
+Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the difference in ROC and AUC.
+
+\begin{table}[h!]
+\caption{Performance of the dropout prediction model}
+\label{tab:drop_pred_rocs}
+\small
+\center
+\begin{tabular}{lccc}
+\toprule
+ & $train$ & $test_1$ & $test_2$ \\
+\midrule
+ROC -- AUC &  0.931 & 0.929 & 0.922  \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Two of the predictors in the fitted model are highly influential.
+The first is the number of days since the learner has last been active ($x_{1}$).
+The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more.
+However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model.
+Equation 1 is an approximation of the model with only these two predictors.
+A learner was flagged if their dropout probability exceeded a half:
+
+\begin{equation}
+Pr(\text{dropout}) = \frac{1}{1+e^{-0.431x_1 + 6.388x_2 + 1.626}} > 0.5
+\end{equation}
+
+Out of the 20,048 enrolled learners in the course, we only considered those 10,510 who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out.
+The survey was sent out to 6,053 leaners (excluding 21 learners with invalid email addresses or duplicate accounts).
 
@@
 Could you help us understand why?'' A low response rate was expected for a subpopulation that was defined by low engagement.
-And yet, 756 out of 6,050 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
+And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
 Missing values were multiply imputated by predictive mean matching using responses to all survey questions\footnote{The $R$ mice package was used to perform multiple imputation and subsequent analyses.}.
@@
 
-\subsubsection{Predicting Likely Dropouts}
+% \subsubsection{Predicting Likely Dropouts}
 
-SHERIF TODO: Add results of prediction.
+% SHERIF TODO: Add results of prediction.
 How well did the prediction work?
@@
 
+Learners drop out at different times during the course.
+To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis.
+However, learners can develop risk factors at different times in the course.
+The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times.
+Such increased survey load can increase attrition or dissatisfaction in the course.
+
+An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged.
+This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk).
+Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
+
 \section{General Discussion}

### 2014-10-26 paper_v1.Rnw
corrected ROC AUC
added study 2 discussion

@@
 To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey.
-One goal of this study was to further investiage research question $RQ3a$ based on learners' own descriptions of the challenges they faced.
+One goal of this study was to further investiage research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced.
 Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time.
@@
 \begin{description}
-  \item[H4a] Learners who report having `not enough time' can be grouped into those with higher priority obligations (professional, academic, personal) and those with low volitional control.
+  \item[H4a] Learners who report having `not enough time' can be grouped into those with higher priority obligations (professional, academic, and personal) and those with low volitional control.
   \item[H4b] Learners who do not report specific higher-priority obligations tend to exhibit lower volitional control than those who specify particular obligations.
@@
 
-Another goal of Study 2 was to collect data to test hypotheses $H1, H2, H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
-To this end, we included a self-report measures for each construct in the survey that was sent out to at-risk learners.
+Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
+To this end, we included a self-report measures for each construct in the survey that was sent to at-risk learners.
 
@@
 
-\textit{1.
-Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible.
+\textit{Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible.
 The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.
 
-\textit{2.
-Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short.
+\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short.
 A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.
 
-\textit{3.
-Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course.
+\textit{Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course.
 We thus had to train the prediction model using data from previous courses.
@@
 The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses.
-Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the test set ($test_1$).
-Another second test set of similar size was composed from 20 other MOOCs ($test_2$).
+Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$).
+A second test set of similar size was composed from 20 other MOOCs ($test_2$).
 Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).
 The model was fit by logistic regression.
-Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the difference in ROC and AUC.
+Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
 
@@
 \midrule
-ROC -- AUC &  0.931 & 0.929 & 0.922  \\
+AUC &  0.931 & 0.929 & 0.922  \\
 \bottomrule
@@
 Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently.
-Could you help us understand why?'' A low response rate was expected for a subpopulation that was defined by low engagement.
+Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement.
 And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
-Missing values were multiply imputated by predictive mean matching using responses to all survey questions\footnote{The $R$ mice package was used to perform multiple imputation and subsequent analyses.}.
-All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
+Missing values were multiply imputated by predictive mean matching using responses to all survey questions.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.} All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
 The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.
@@
 Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked.
-In addition, they reported intentions for engaging with each week's course content.
-Then, they were asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of.
+These items served as measures of personal success.
+% In addition, they reported intentions for engaging with each week's course content. 
+They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of.
 This question was deliberately asked prior to any survey questions that could suggest particular reasons for disengagement.
@@
 
-Besides these open responses, learners rated their relative progress and satisfaction with their progress in the course, which served as measures of personal success.
-They also rated the extent to which their progress was hindered by a number of obstacles.
-The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and one item on volition (about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
+% Learners also rated the extent to which their progress was hindered by a number of obstacles. 
+Learners also rated their sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
 
@@
 
-The majority of survey respondents (73\%; 550 of 756) reported being only somewhat or less satisfied with their progress in the course.
+The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course.
 While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
-volition).
-These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.
+exhibit lower volitional control).
+These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).
 
-Reasons for learner attrition were measured using open response as well as multiple-choice questions.
-Based on learners' coded open responses ($n=399$), 84\% had `not enough time for the course', 18\% indicated that `the course format or teaching style was not a good fit', 10\% expressed that `the level of the course was too advanced', 7\% found that `the course was not interesting or valuable enough', 4\% expressed a need for `additional course materials to learn the topic', 2\% experienced `technical difficulties', and 1\% reported facing a `language barrier'.
+Reasons for learner attrition were measured using iteratively coded open responses ($n=399$).
+Table \ref{tab:s2reas} provides each coded challenge and percentage of open responses that indicated it.
 A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).
 
-To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time, which revealed that about half of them (53\%) did not specify a particular committment.
+\begin{table}[h!]
+\caption{Iteratively-Coded Reasons for Attrition}
+\label{tab:s2reas}
+\small
+\center
+\begin{tabular}{lr}
+\toprule
+What challenges have you experienced while taking this course?  & \% \\
+\midrule
+Not enough time for the course & 84 \\
+The course format or teaching style is not a good fit & 18 \\
+The level of the course is too advanced & 10 \\
+The course is not interesting or valuable enough & 7 \\
+Additional course materials needed to learn the topic & 4 \\
+Technical difficulties & 2 \\
+Language barrier & 1 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time.
+It revealed that no particular commitment was specified in about half of them (53\%).
 This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders.
 Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
-We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volition, i.e.
+We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e.
 their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
-However, the two groups did not significantly differ in how much they were held back by committments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
-These findings lend strong support to hypotheses 4a and 4b: while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volition and reporting no specific reasons for having not enough time was indicative thereof.
+However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
+These findings lend strong support to hypotheses $H4a$ and $H4b$: that is, while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
 
@@
 Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
-The two measures were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
+The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
 Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
@@
 The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction.
-The results fully support $H1$, and partially support $H2$ and $H3$.
+Hence, the results fully support hypothesis $H1$, and partially support $H2$ and $H3$.
 
@@
 
-Learners drop out at different times during the course.
-To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis.
-However, learners can develop risk factors at different times in the course.
-The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times.
-Such increased survey load can increase attrition or dissatisfaction in the course.
+Study 2 focused on reasons for attrition and psychological factors of learners who were likely to drop out of the course.
+Through iterative coding, we developed a set of challenges that were faced by a subset of learners in a particular MOOC (cf.
+Table \ref{tab:s2reas}).
+The large majority of learners who indicate that they had not have enough time was found to be sperable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$).
+MOOCs are an environment that requires learners to be highly self-regulated to be successful.
+The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}.
+Very few learners explicitly indicated issues with volitional control in their responses, potentially because it is not socially desirable to admit.
+The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$).
+This could be a useful proxy for volitional issues in future research.
 
-An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged.
-This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk).
-Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
+We investigated the extent to which personal success in MOOCs is associated with three established psychological factors: goal striving, social belonging, and growth mindset.
+Prior evidence found the absence of each factor to cause lower persistence and performance in more traditional achievement-oriented environments.
+The present findings highlight the critical role of these factors in MOOCs based on observational evidence.
+Learners with higher levels of goal striving, social belonging, and growth mindset also reported higher levels of personal success ($H1,H2,H3$).
+The subpopulation of learners in this study was a small and intentionally skewed subset.
+The observed differences are expected to be larger in the full population of learners, compared to this homogenous sample of learners who are unlikely to persist.
+Goal striving was most strongly associated with both measures of personal success.
+Success in terms of relative progress was associated with a stronger growth mindset, but not significantly with social belonging.
+In contrast, success in terms of satisfaction with progress was associated with stronger social belonging, but not significantly with growth mindset.
+Future work should examine if this pattern reflects a difference in the underlying process by which these psychological factors affect learner behavior and perception.
+
+The methodology used in this study can be extended to deliver timely and targeted interventions.
+The predictive model was successful in identifying a large number of leaners who were facing challenges with taking the course.
+The majority of them were unsuccessful in their own terms.
+All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners.
+Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).
+Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their identity or social belonging.
 
@@
 Omitted for blind review.
+
+% Learners drop out at different times during the course.
+To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis.
+However, learners can develop risk factors at different times in the course.
+The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times.
+Such increased survey load can increase attrition or dissatisfaction in the course.
+
+% An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged.
+This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk).
+Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
 

### 2014-10-27 paper_v1.Rnw
Testing pushing changes to .Rnw only

--- 2014-10-26 paper_v1.Rnw
@@
 \begin{abstract}
-Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from [N] online learners in [M] courses.
+Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses.
 Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions.
@@
 
-TODO: Outline different courses and go through variabels in the course info table.
+% Which courses did we use? What data
+In this study, we used 17 MOOCs from different areas including humanities sciences (4 courses), engineering (2 courses), computer science (3 courses), mathematics and statistics (6 courses), and business (2 instances of 1 course).
+For each course, we acquired the data from the pre-course and post-course surveys, as well as the learner activity data from the platform hosting the course.
+From each course, we selected all learners who took the post-course survey, and we matched learners to their activity data and pre-course survey response (if available).
+Table \ref{tab:study1_courses_summary} shows the number of enrolled learners, number of learners who responses to the pre-course/post-course survey, distributions of age, gender, and education level, and the proportion of learner who self-reported that they stopped participating in the course.
+We also computed the proprtions of learners who viewed at least $p$\% of the course videos for $p = 30\%$, $50\%$, and $80\%$.
 
-Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+%Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+\newcommand{\specialcellcentered}[2][c]{%
+  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+\newcommand{\specialcellleft}[2][c]{%
+  \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
+
+\begin{table*}
+\label{tab:study1_courses_summary}
+\caption{Summary on courses and surveys used in Study 1}
+\small
+\center
+\begin{tabular}{llrrrrrrrrrr}
+\toprule
+ID & Topic Area & Enrollment & Pre-survey & Post-survey & Female & Age & Education & \specialcellcentered{\ensuremath{>} 30\% \\ videos} & \specialcellcentered{\ensuremath{>} 50\% \\ videos} & \specialcellcentered{\ensuremath{>} 80\% \\ videos} & \specialcellcentered{Self-Rep.
+\\ Dropout} \\
+\midrule
+C1 & Business & 61096 & 9943 & 2367 & 42\% & 39 (30,49) & n/a & 39\% & 20\% & 7\% & 10\% \\
+C2 & Business & 41081 & 6544 & 883 & 40\% & 39 (30,49) & 97\% & 32\% & 13\% & 2\% & 9\% \\
+C3 & Comp. Sc. & 27104 & 6104 & 2854 & 35\% & 30 (22,41) & 74\% & 67\% & 40\% & 7\% & 2\% \\
+C4 & Comp. Sc. & 56154 & 3406 & 1123 & 11\% & 33 (27,42) & 91\% & 34\% & 16\% & 11\% & 21\% \\
+C5 & Comp. Sc. & 57636 & 5738 & 402 & 17\% & 33 (26,43) & 91\% & 66\% & 55\% & 28\% & 14\% \\
+C6 & Engineering & 16220 & 2540 & 269 & 13\% & 33 (27,45) & 92\% & 62\% & 49\% & 17\% & 15\% \\
+C7 & Enginerring & 7178 & 1312 & 202 & 12\% & 27 (22,39) & 85\% & 51\% & 18\% & 2\% & 7\% \\
+C8 & Humanities & 7634 & 3631 & 152 & 92\% & 33 (26,45) & 93\% & 45\% & 11\% & 0\% & 7\% \\
+C9 & Humanities & 6475 & 3315 & 326 & 90\% & 32 (27,47) & 91\% & 69\% & 50\% & 17\% & 16\% \\
+C10 & Humanities & 9903 & 4227 & 1570 & 52\% & 39 (29,52) & 89\% & 84\% & 74\% & 21\% & 3\% \\
+C11 & Humanities & 630 & 167 & 80 & 50\% & 30 (27,40) & 97\% & 74\% & 55\% & 16\% & 16\% \\
+C12 & Math \& Stats & 49317 & 3090 & 862 & 33\% & 41 (29,55) & 88\% & 27\% & 10\% & 1\% & 31\% \\
+C13 & Math \& Stats & 32006 & 26142 & 5336 & 58\% & 28 (16,46) & n/a & 69\% & 38\% & 8\% & 4\% \\
+C14 & Math \& Stats & 16857 & 2892 & 363 & 9\% & 31 (26,38) & 99\% & 61\% & 45\% & 25\% & 14\% \\
+C15 & Math \& Stats & 37017 & 11629 & 1854 & 20\% & 35 (29,45) & 98\% & 78\% & 67\% & 34\% & 8\% \\
+C16 & Math \& Stats & 27236 & 2711 & 637 & 82\% & 43 (34,51) & n/a & 64\% & 36\% & 5\% & 1\% \\
+C17 & Math \& Stats & 13132 & 3629 & 843 & n/a & n/a & n/a & 70\% & 54\% & 14\% & 5\% \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+% What LMS data variables did we compute?
+We are interested in exploring correlations between reasons for dropout and learner behaviors.
+If such correlations exist, they can be an additional towards understanding why learners engage with or disengage from various learning activities.
+In this study, we computed the fraction of videos watched by each learner, fraction of problems attempted, average grade scored on attempted problems, and number of forum posts and replies posted.
 
@@
 
-Table: proportions for each dropout reason for each course
+%Table: proportions for each dropout reason for each course
+\begin{table*}
+\label{tab:proprtion_reason_for_dropout}
+\caption{Distribution of reported reasons for dropping out across courses.
+Percentage of learners who selected each reason for dropping out in each course, and median and interquartile range across courses}
+\small
+\center
+\begin{tabular}{llrrrrrrrrrrrrrrrrrr}
+\toprule
+Reason for dropout & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & Median & IQR \\
+\midrule
+\specialcellleft{Material not explained \\ properly} & 18 & 20 & 10 & 22 & 19 & 21 & 8 & 11 & 13 & 13 & 9 & 7 & 14 & 18 & 11 & 14 & 19 & 14 & 8 \\
+\specialcellleft{Course requires too \\ much time} & 57 & 39 & 13 & 48 & 42 & 51 & 33 & 60 & 65 & 42 & 18 & 52 & 21 & 68 & 39 & 28 & 68 & 42 & 24 \\
+\specialcellleft{Not what I was looking \\ for} & 21 & 23 & 25 & 14 & 13 & 19 & 8 & 0 & 0 & 23 & 9 & 12 & 20 & 20 & 13 & 28 & 12 & 14 & 9 \\
+I joined too late & 32 & 18 & 25 & 28 & 32 & 28 & 25 & 11 & 20 & 27 & 27 & 19 & 19 & 22 & 18 & 0 & 11 & 22 & 9 \\
+\specialcellleft{Didn't enjoy online \\ course format} & 15 & 14 & 8 & 16 & 13 & 12 & 0 & 0 & 9 & 10 & 9 & 9 & 19 & 12 & 7 & 14 & 14 & 12 & 5 \\
+Technical difficulties & 21 & 3 & 18 & 12 & 17 & 24 & 25 & 30 & 30 & 10 & 36 & 13 & 20 & 25 & 12 & 0 & 21 & 20 & 13 \\
+\specialcellleft{Couldn't keep up with \\ deadlines} & 69 & 58 & 38 & 76 & 80 & 66 & 66 & 55 & 69 & 67 & 27 & 71 & 28 & 64 & 54 & 0 & 71 & 66 & 15 \\
+Not challenging enough & 12 & 7 & 27 & 8 & 4 & 18 & 41 & 11 & 6 & 10 & 27 & 10 & 24 & 14 & 11 & 28 & 7 & 11 & 16 \\
+I learned all I wanted & 18 & 7 & 26 & 17 & 19 & 16 & 41 & 10 & 9 & 25 & 18 & 21 & 27 & 25 & 20 & 28 & 12 & 19 & 9 \\
+Videos were confusing & 13 & 12 & 10 & 19 & 19 & 21 & 0 & 11 & 7 & 13 & 9 & 9 & 11 & 16 & 9 & 14 & 19 & 12 & 7 \\
+I was only exploring & 16 & 5 & 22 & 17 & 14 & 21 & 16 & 11 & 7 & 18 & 20 & 16 & 23 & 20 & 16 & 16 & 12 & 16 & 6 \\
+\specialcellleft{Too advanced for my \\ background} & 14 & 10 & 10 & 23 & 24 & 16 & 8 & 11 & 2 & 13 & 18 & 19 & 10 & 25 & 17 & 0 & 29 & 14 & 9 \\
+\specialcellleft{Assn or exam was too \\ difficult} & 26 & 8 & 11 & 27 & 23 & 39 & 33 & 11 & 24 & 11 & 27 & 27 & 11 & 35 & 15 & 14 & 41 & 24 & 16 \\
+\specialcellleft{Credentials not valuable \\ to me} & 14 & 9 & 18 & 10 & 10 & 18 & 25 & 22 & 4 & 5 & 18 & 7 & 20 & 22 & 14 & 14 & 25 & 14 & 10 \\
+\bottomrule
+\end{tabular}
+\end{table*}
 

### 2014-10-27 paper_v1.Rnw
added figure, made changes to methods and tables

--- 2014-10-26 paper_v1.Rnw
@@
 \usepackage{booktabs}
+\usepackage{rotating}
 
@@
 
-TODO: Outline different courses and go through variabels in the course info table.
+% Which courses did we use? What data
+Seventeen MOOCs on topcis in a variety of disciplines were selected for investigation: four on topics in the humanities, two in engineering, three in computer science, two runs of the same course in business, and six in mathematics and statistics.
+For each course, we merged data from the course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey).
+Table \ref{tab:study1_courses_summary} provides the following summary statistics for each course: the number of enrolled learners, the number of learners who completed the course survey and the feedback surveys, age, gender, and education level, and the proportion of learners who self-reported that they stopped participating in the course.
+We also computed the proprtions of learners who viewed at least 30, 50, and 80\% of the course videos.
 
-Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+%Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+
+\begin{table*}
+\label{tab:study1_courses_summary}
+\caption{Summary statistics of 14 MOOCs}
+\small
+\center
+\begin{tabular}{llcccccccccc}
+\toprule
+%  & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & \specialcellcentered{\ensuremath{>} 30\% \\ videos} & \specialcellcentered{\ensuremath{>} 50\% \\ videos} & \specialcellcentered{\ensuremath{>} 80\% \\ videos} & \specialcellcentered{Self-Rep.
+\\ Dropout} \\
+& & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
+& Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
+\midrule
+C1 & Business & 61,096 & 9,943 & 2,367 & 42\% & 39 (30,49) & n/a & 39\% & 20\% & 7\% & 10\% \\
+C2 & Business & 41,081 & 6,544 & 883 & 40\% & 39 (30,49) & 97\% & 32\% & 13\% & 2\% & 9\% \\
+C3 & CS & 27,104 & 6,104 & 2,854 & 35\% & 30 (22,41) & 74\% & 67\% & 40\% & 7\% & 2\% \\
+C4 & CS & 56,154 & 3,406 & 1,123 & 11\% & 33 (27,42) & 91\% & 34\% & 16\% & 11\% & 21\% \\
+C5 & CS & 57,636 & 5738 & 402 & 17\% & 33 (26,43) & 91\% & 66\% & 55\% & 28\% & 14\% \\
+C6 & Engineering & 16,220 & 2,540 & 269 & 13\% & 33 (27,45) & 92\% & 62\% & 49\% & 17\% & 15\% \\
+C7 & Enginerring & 7,178 & 1,312 & 202 & 12\% & 27 (22,39) & 85\% & 51\% & 18\% & 2\% & 7\% \\
+C8 & Humanities & 7,634 & 3,631 & 152 & 92\% & 33 (26,45) & 93\% & 45\% & 11\% & 0\% & 7\% \\
+C9 & Humanities & 6,475 & 3,315 & 326 & 90\% & 32 (27,47) & 91\% & 69\% & 50\% & 17\% & 16\% \\
+C10 & Humanities & 9,903 & 4,227 & 1,570 & 52\% & 39 (29,52) & 89\% & 84\% & 74\% & 21\% & 3\% \\
+C11 & Humanities & 630 & 167 & 80 & 50\% & 30 (27,40) & 97\% & 74\% & 55\% & 16\% & 16\% \\
+C12 & Math \& Stats & 49,317 & 3,090 & 862 & 33\% & 41 (29,55) & 88\% & 27\% & 10\% & 1\% & 31\% \\
+C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 58\% & 28 (16,46) & n/a & 69\% & 38\% & 8\% & 4\% \\
+C14 & Math \& Stats & 16,857 & 2,892 & 363 & 9\% & 31 (26,38) & 99\% & 61\% & 45\% & 25\% & 14\% \\
+C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 20\% & 35 (29,45) & 98\% & 78\% & 67\% & 34\% & 8\% \\
+C16 & Math \& Stats & 27,236 & 2,711 & 637 & 82\% & 43 (34,51) & n/a & 64\% & 36\% & 5\% & 1\% \\
+C17 & Math \& Stats & 13,132 & 3,629 & 843 & n/a & n/a & n/a & 70\% & 54\% & 14\% & 5\% \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+% What LMS data variables did we compute?
+We are interested in exploring correlations between reasons for dropout and learner behaviors.
+If such correlations exist, they can be an additional towards understanding why learners engage with or disengage from various learning activities.
+In this study, we computed the fraction of videos watched by each learner, fraction of problems attempted, average grade scored on attempted problems, and number of forum posts and replies posted.
 
@@
 
-Table: proportions for each dropout reason for each course
+%Table: proportions for each dropout reason for each course
+<<prop-reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
+values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
+57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
+21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
+32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
+15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
+21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
+69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
+12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
+18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
+13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
+16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
+14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
+26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
+14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)
+
+reasons = c("Material not explained properly","Course requires too much time","Not what I was looking for","I joined too late","Didn't enjoy online course format","Technical difficulties","Couldn't keep up with deadlines","Not challenging enough","I learned all I wanted","Videos were confusing","I was only exploring","Too advanced for my background","Assignment or exam was too difficult","Credentials not valuable to me")
+
+dat = data.frame(reasons = rep(reasons, each=17), values)
+
+library(ggplot2)
+ggplot(dat, aes(reorder(reasons, values, median), values)) + geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion",x="")
+@
+
+% \begin{sidewaystable}
+% \label{tab:proprtion_reason_for_dropout}
+% \caption{Distribution of reported reasons for dropping out across courses.
+Percentage of learners who selected each reason for dropping out in each course, and median and interquartile range across courses}
+% \small
+% \center
+% \begin{tabular}{llrrrrrrrrrrrrrrrrrr}
+% \toprule
+% Reason for dropout & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & Median & IQR \\
+% \midrule
+% \specialcellleft{Material not explained \\ properly} & 18 & 20 & 10 & 22 & 19 & 21 & 8 & 11 & 13 & 13 & 9 & 7 & 14 & 18 & 11 & 14 & 19 & 14 & 8 \\
+% \specialcellleft{Course requires too \\ much time} & 57 & 39 & 13 & 48 & 42 & 51 & 33 & 60 & 65 & 42 & 18 & 52 & 21 & 68 & 39 & 28 & 68 & 42 & 24 \\
+% \specialcellleft{Not what I was looking \\ for} & 21 & 23 & 25 & 14 & 13 & 19 & 8 & 0 & 0 & 23 & 9 & 12 & 20 & 20 & 13 & 28 & 12 & 14 & 9 \\
+% I joined too late & 32 & 18 & 25 & 28 & 32 & 28 & 25 & 11 & 20 & 27 & 27 & 19 & 19 & 22 & 18 & 0 & 11 & 22 & 9 \\
+% \specialcellleft{Didn't enjoy online \\ course format} & 15 & 14 & 8 & 16 & 13 & 12 & 0 & 0 & 9 & 10 & 9 & 9 & 19 & 12 & 7 & 14 & 14 & 12 & 5 \\
+% Technical difficulties & 21 & 3 & 18 & 12 & 17 & 24 & 25 & 30 & 30 & 10 & 36 & 13 & 20 & 25 & 12 & 0 & 21 & 20 & 13 \\
+% \specialcellleft{Couldn't keep up with \\ deadlines} & 69 & 58 & 38 & 76 & 80 & 66 & 66 & 55 & 69 & 67 & 27 & 71 & 28 & 64 & 54 & 0 & 71 & 66 & 15 \\
+% Not challenging enough & 12 & 7 & 27 & 8 & 4 & 18 & 41 & 11 & 6 & 10 & 27 & 10 & 24 & 14 & 11 & 28 & 7 & 11 & 16 \\
+% I learned all I wanted & 18 & 7 & 26 & 17 & 19 & 16 & 41 & 10 & 9 & 25 & 18 & 21 & 27 & 25 & 20 & 28 & 12 & 19 & 9 \\
+% Videos were confusing & 13 & 12 & 10 & 19 & 19 & 21 & 0 & 11 & 7 & 13 & 9 & 9 & 11 & 16 & 9 & 14 & 19 & 12 & 7 \\
+% I was only exploring & 16 & 5 & 22 & 17 & 14 & 21 & 16 & 11 & 7 & 18 & 20 & 16 & 23 & 20 & 16 & 16 & 12 & 16 & 6 \\
+% \specialcellleft{Too advanced for my \\ background} & 14 & 10 & 10 & 23 & 24 & 16 & 8 & 11 & 2 & 13 & 18 & 19 & 10 & 25 & 17 & 0 & 29 & 14 & 9 \\
+% \specialcellleft{Assn or exam was too \\ difficult} & 26 & 8 & 11 & 27 & 23 & 39 & 33 & 11 & 24 & 11 & 27 & 27 & 11 & 35 & 15 & 14 & 41 & 24 & 16 \\
+% \specialcellleft{Credentials not valuable \\ to me} & 14 & 9 & 18 & 10 & 10 & 18 & 25 & 22 & 4 & 5 & 18 & 7 & 20 & 22 & 14 & 14 & 25 & 14 & 10 \\
+% \bottomrule
+% \end{tabular}
+% \end{sidewaystable}
 

### 2014-10-27 paper_v1.tex
added figure, made changes to methods and tables

--- 2014-10-26 paper_v1.tex
@@
 
-Table: proportions for each dropout reason for each course
+%Table: proportions for each dropout reason for each course
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
+
+\includegraphics[width=\maxwidth]{figure/prop-reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:prop-reason}}
+\end{figure*}
+
+\end{knitrout}
+
+% \begin{sidewaystable}
+% \label{tab:proprtion_reason_for_dropout}
+% \caption{Distribution of reported reasons for dropping out across courses.
+Percentage of learners who selected each reason for dropping out in each course, and median and interquartile range across courses}
+% \small
+% \center
+% \begin{tabular}{llrrrrrrrrrrrrrrrrrr}
+% \toprule
+% Reason for dropout & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & Median & IQR \\
+% \midrule
+% \specialcellleft{Material not explained \\ properly} & 18 & 20 & 10 & 22 & 19 & 21 & 8 & 11 & 13 & 13 & 9 & 7 & 14 & 18 & 11 & 14 & 19 & 14 & 8 \\
+% \specialcellleft{Course requires too \\ much time} & 57 & 39 & 13 & 48 & 42 & 51 & 33 & 60 & 65 & 42 & 18 & 52 & 21 & 68 & 39 & 28 & 68 & 42 & 24 \\
+% \specialcellleft{Not what I was looking \\ for} & 21 & 23 & 25 & 14 & 13 & 19 & 8 & 0 & 0 & 23 & 9 & 12 & 20 & 20 & 13 & 28 & 12 & 14 & 9 \\
+% I joined too late & 32 & 18 & 25 & 28 & 32 & 28 & 25 & 11 & 20 & 27 & 27 & 19 & 19 & 22 & 18 & 0 & 11 & 22 & 9 \\
+% \specialcellleft{Didn't enjoy online \\ course format} & 15 & 14 & 8 & 16 & 13 & 12 & 0 & 0 & 9 & 10 & 9 & 9 & 19 & 12 & 7 & 14 & 14 & 12 & 5 \\
+% Technical difficulties & 21 & 3 & 18 & 12 & 17 & 24 & 25 & 30 & 30 & 10 & 36 & 13 & 20 & 25 & 12 & 0 & 21 & 20 & 13 \\
+% \specialcellleft{Couldn't keep up with \\ deadlines} & 69 & 58 & 38 & 76 & 80 & 66 & 66 & 55 & 69 & 67 & 27 & 71 & 28 & 64 & 54 & 0 & 71 & 66 & 15 \\
+% Not challenging enough & 12 & 7 & 27 & 8 & 4 & 18 & 41 & 11 & 6 & 10 & 27 & 10 & 24 & 14 & 11 & 28 & 7 & 11 & 16 \\
+% I learned all I wanted & 18 & 7 & 26 & 17 & 19 & 16 & 41 & 10 & 9 & 25 & 18 & 21 & 27 & 25 & 20 & 28 & 12 & 19 & 9 \\
+% Videos were confusing & 13 & 12 & 10 & 19 & 19 & 21 & 0 & 11 & 7 & 13 & 9 & 9 & 11 & 16 & 9 & 14 & 19 & 12 & 7 \\
+% I was only exploring & 16 & 5 & 22 & 17 & 14 & 21 & 16 & 11 & 7 & 18 & 20 & 16 & 23 & 20 & 16 & 16 & 12 & 16 & 6 \\
+% \specialcellleft{Too advanced for my \\ background} & 14 & 10 & 10 & 23 & 24 & 16 & 8 & 11 & 2 & 13 & 18 & 19 & 10 & 25 & 17 & 0 & 29 & 14 & 9 \\
+% \specialcellleft{Assn or exam was too \\ difficult} & 26 & 8 & 11 & 27 & 23 & 39 & 33 & 11 & 24 & 11 & 27 & 27 & 11 & 35 & 15 & 14 & 41 & 24 & 16 \\
+% \specialcellleft{Credentials not valuable \\ to me} & 14 & 9 & 18 & 10 & 10 & 18 & 25 & 22 & 4 & 5 & 18 & 7 & 20 & 22 & 14 & 14 & 25 & 14 & 10 \\
+% \bottomrule
+% \end{tabular}
+% \end{sidewaystable}
 

### 2014-10-27 paper_v1.Rnw
Expanding study 1 Please enter the commit message for your changes. Lines starting

--- 2014-10-27 paper_v1.Rnw
@@
 In this study, we used 17 MOOCs from different areas including humanities sciences (4 courses), engineering (2 courses), computer science (3 courses), mathematics and statistics (6 courses), and business (2 instances of 1 course).
-For each course, we acquired the data from the pre-course and post-course surveys, as well as the learner activity data from the platform hosting the course.
+For each course, we acquired the data from the pre-and post-course surveys, as well as the learner activity data from the online platform.
 From each course, we selected all learners who took the post-course survey, and we matched learners to their activity data and pre-course survey response (if available).
-Table \ref{tab:study1_courses_summary} shows the number of enrolled learners, number of learners who responses to the pre-course/post-course survey, distributions of age, gender, and education level, and the proportion of learner who self-reported that they stopped participating in the course.
-We also computed the proprtions of learners who viewed at least $p$\% of the course videos for $p = 30\%$, $50\%$, and $80\%$.
+
+The post-course survey for each course asked the learners a range of questions about quality of experience with the course (Extremely dissatisfied - Extremely satisfied, 7 point scale), the extent to which they found it difficult/easy (Extremely easy - Extremely difficult, 7 point scale), whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
+For each reason, the student was asked to select one of two options: "Did influence my decision to stop" and "Did NOT influence my decision to stop".
+Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reportinga reason and those who express that the reason did not apply to them.
+
+Table \ref{tab:study1_courses_summary} shows the number of enrolled learners, percentage of learners who took the post-course survey, percentage who took both the pre- and post-course surveys, distributions of age, gender, and education level, and the proportion of learner who reported in the post-course survey that they stopped participating in the course before the end.
+The table also shows the proprtions of learners who viewed at least $30\%$/$50\%$/$80\%$ of the course videos.
+Note that the target of this study are the learners who took the post-course survey, whereas the demographic questions are asked in the pre-course survey.
+The age, gender, and education statistics are thus computed from the learners who took both surveys, which might be different from the statistics of the post-course survey takers, pre-course takers, or whole course learner population.
 
@@
 
-% What LMS data variables did we compute?
-We are interested in exploring correlations between reasons for dropout and learner behaviors.
-If such correlations exist, they can be an additional towards understanding why learners engage with or disengage from various learning activities.
-In this study, we computed the fraction of videos watched by each learner, fraction of problems attempted, average grade scored on attempted problems, and number of forum posts and replies posted.
+% Proportions for each dropout reason for each course
+We are interested in the proportions of respondents that reported each reason for dropout across the courses.
+These proportions are shown in [Ref to Box plot].
+The proprtion for each reason was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it aplies or that it does not apply).
+We also computed the median and interquartile range of the proprtions across all courses for each reason.
+
+% Correlations among rfd's and between rfd's and ei's.
+For each  course, we also computed the correlations between the reported reasons for dropout (Table \ref{tab:study1_rfd_corr}), and the correlations between reported reasons for dropout and enrollment intentions (Table \ref{tab:study1_rfd_ei_corr}).
+Each table shows a linear combination of the correlation matrices of the 17 courses.
+
+% Proportion videos watched for each disengagement reason
+We are also interested in the correlations between reported reasons for dropout and certain learner behaviors.
+For each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
+We then computed the matrix of correlations between reasons for dropout and these variables for each course.
+These matrices were averaged and the result is shown in Table \ref{tab:study1_rfd_video_behavior_corr}.
 
 \subsection{Results}
-
-%Table: proportions for each dropout reason for each course
-\begin{table*}
-\label{tab:proprtion_reason_for_dropout}
-\caption{Distribution of reported reasons for dropping out across courses.
-Percentage of learners who selected each reason for dropping out in each course, and median and interquartile range across courses}
-\small
-\center
-\begin{tabular}{llrrrrrrrrrrrrrrrrrr}
-\toprule
-Reason for dropout & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & Median & IQR \\
-\midrule
-\specialcellleft{Material not explained \\ properly} & 18 & 20 & 10 & 22 & 19 & 21 & 8 & 11 & 13 & 13 & 9 & 7 & 14 & 18 & 11 & 14 & 19 & 14 & 8 \\
-\specialcellleft{Course requires too \\ much time} & 57 & 39 & 13 & 48 & 42 & 51 & 33 & 60 & 65 & 42 & 18 & 52 & 21 & 68 & 39 & 28 & 68 & 42 & 24 \\
-\specialcellleft{Not what I was looking \\ for} & 21 & 23 & 25 & 14 & 13 & 19 & 8 & 0 & 0 & 23 & 9 & 12 & 20 & 20 & 13 & 28 & 12 & 14 & 9 \\
-I joined too late & 32 & 18 & 25 & 28 & 32 & 28 & 25 & 11 & 20 & 27 & 27 & 19 & 19 & 22 & 18 & 0 & 11 & 22 & 9 \\
-\specialcellleft{Didn't enjoy online \\ course format} & 15 & 14 & 8 & 16 & 13 & 12 & 0 & 0 & 9 & 10 & 9 & 9 & 19 & 12 & 7 & 14 & 14 & 12 & 5 \\
-Technical difficulties & 21 & 3 & 18 & 12 & 17 & 24 & 25 & 30 & 30 & 10 & 36 & 13 & 20 & 25 & 12 & 0 & 21 & 20 & 13 \\
-\specialcellleft{Couldn't keep up with \\ deadlines} & 69 & 58 & 38 & 76 & 80 & 66 & 66 & 55 & 69 & 67 & 27 & 71 & 28 & 64 & 54 & 0 & 71 & 66 & 15 \\
-Not challenging enough & 12 & 7 & 27 & 8 & 4 & 18 & 41 & 11 & 6 & 10 & 27 & 10 & 24 & 14 & 11 & 28 & 7 & 11 & 16 \\
-I learned all I wanted & 18 & 7 & 26 & 17 & 19 & 16 & 41 & 10 & 9 & 25 & 18 & 21 & 27 & 25 & 20 & 28 & 12 & 19 & 9 \\
-Videos were confusing & 13 & 12 & 10 & 19 & 19 & 21 & 0 & 11 & 7 & 13 & 9 & 9 & 11 & 16 & 9 & 14 & 19 & 12 & 7 \\
-I was only exploring & 16 & 5 & 22 & 17 & 14 & 21 & 16 & 11 & 7 & 18 & 20 & 16 & 23 & 20 & 16 & 16 & 12 & 16 & 6 \\
-\specialcellleft{Too advanced for my \\ background} & 14 & 10 & 10 & 23 & 24 & 16 & 8 & 11 & 2 & 13 & 18 & 19 & 10 & 25 & 17 & 0 & 29 & 14 & 9 \\
-\specialcellleft{Assn or exam was too \\ difficult} & 26 & 8 & 11 & 27 & 23 & 39 & 33 & 11 & 24 & 11 & 27 & 27 & 11 & 35 & 15 & 14 & 41 & 24 & 16 \\
-\specialcellleft{Credentials not valuable \\ to me} & 14 & 9 & 18 & 10 & 10 & 18 & 25 & 22 & 4 & 5 & 18 & 7 & 20 & 22 & 14 & 14 & 25 & 14 & 10 \\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-Table: correlations between reasons and one with cors between reasons and intentions
-
-Plot: ecdf curves for proportion videos watched for each disengagement reason
 

### 2014-10-27 paper_v1.Rnw

--- 2014-10-27 paper_v1.Rnw
@@
 Seventeen MOOCs on topcis in a variety of disciplines were selected for investigation: four on topics in the humanities, two in engineering, three in computer science, two runs of the same course in business, and six in mathematics and statistics.
-For each course, we merged data from the course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
 Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey).
 Table \ref{tab:study1_courses_summary} provides the following summary statistics for each course: the number of enrolled learners, the number of learners who completed the course survey and the feedback surveys, age, gender, and education level, and the proportion of learners who self-reported that they stopped participating in the course.
+Note that age, gender, and education level statistics were computed using only the subset of learners who took both the initial and feedback surveys.
 We also computed the proprtions of learners who viewed at least 30, 50, and 80\% of the course videos.
@@
 
-% What LMS data variables did we compute?
-We are interested in exploring correlations between reasons for dropout and learner behaviors.
-If such correlations exist, they can be an additional towards understanding why learners engage with or disengage from various learning activities.
-In this study, we computed the fraction of videos watched by each learner, fraction of problems attempted, average grade scored on attempted problems, and number of forum posts and replies posted.
+% Proportions for each dropout reason for each course
+We are interested in the proportions of respondents that reported each reason for dropout across the courses.
+These proportions are shown in Figure \ref{fig:prop-reason}.
+The proprtion for each reason was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it aplies or that it does not apply).
+We also computed the median and interquartile range of the proprtions across all courses for each reason.
 
-\subsection{Results}
-
-%Table: proportions for each dropout reason for each course
+%Figure: proportions for each dropout reason for each course
 <<prop-reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
@@
 
-% \begin{sidewaystable}
-% \label{tab:proprtion_reason_for_dropout}
-% \caption{Distribution of reported reasons for dropping out across courses.
-Percentage of learners who selected each reason for dropping out in each course, and median and interquartile range across courses}
-% \small
-% \center
-% \begin{tabular}{llrrrrrrrrrrrrrrrrrr}
-% \toprule
-% Reason for dropout & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & Median & IQR \\
-% \midrule
-% \specialcellleft{Material not explained \\ properly} & 18 & 20 & 10 & 22 & 19 & 21 & 8 & 11 & 13 & 13 & 9 & 7 & 14 & 18 & 11 & 14 & 19 & 14 & 8 \\
-% \specialcellleft{Course requires too \\ much time} & 57 & 39 & 13 & 48 & 42 & 51 & 33 & 60 & 65 & 42 & 18 & 52 & 21 & 68 & 39 & 28 & 68 & 42 & 24 \\
-% \specialcellleft{Not what I was looking \\ for} & 21 & 23 & 25 & 14 & 13 & 19 & 8 & 0 & 0 & 23 & 9 & 12 & 20 & 20 & 13 & 28 & 12 & 14 & 9 \\
-% I joined too late & 32 & 18 & 25 & 28 & 32 & 28 & 25 & 11 & 20 & 27 & 27 & 19 & 19 & 22 & 18 & 0 & 11 & 22 & 9 \\
-% \specialcellleft{Didn't enjoy online \\ course format} & 15 & 14 & 8 & 16 & 13 & 12 & 0 & 0 & 9 & 10 & 9 & 9 & 19 & 12 & 7 & 14 & 14 & 12 & 5 \\
-% Technical difficulties & 21 & 3 & 18 & 12 & 17 & 24 & 25 & 30 & 30 & 10 & 36 & 13 & 20 & 25 & 12 & 0 & 21 & 20 & 13 \\
-% \specialcellleft{Couldn't keep up with \\ deadlines} & 69 & 58 & 38 & 76 & 80 & 66 & 66 & 55 & 69 & 67 & 27 & 71 & 28 & 64 & 54 & 0 & 71 & 66 & 15 \\
-% Not challenging enough & 12 & 7 & 27 & 8 & 4 & 18 & 41 & 11 & 6 & 10 & 27 & 10 & 24 & 14 & 11 & 28 & 7 & 11 & 16 \\
-% I learned all I wanted & 18 & 7 & 26 & 17 & 19 & 16 & 41 & 10 & 9 & 25 & 18 & 21 & 27 & 25 & 20 & 28 & 12 & 19 & 9 \\
-% Videos were confusing & 13 & 12 & 10 & 19 & 19 & 21 & 0 & 11 & 7 & 13 & 9 & 9 & 11 & 16 & 9 & 14 & 19 & 12 & 7 \\
-% I was only exploring & 16 & 5 & 22 & 17 & 14 & 21 & 16 & 11 & 7 & 18 & 20 & 16 & 23 & 20 & 16 & 16 & 12 & 16 & 6 \\
-% \specialcellleft{Too advanced for my \\ background} & 14 & 10 & 10 & 23 & 24 & 16 & 8 & 11 & 2 & 13 & 18 & 19 & 10 & 25 & 17 & 0 & 29 & 14 & 9 \\
-% \specialcellleft{Assn or exam was too \\ difficult} & 26 & 8 & 11 & 27 & 23 & 39 & 33 & 11 & 24 & 11 & 27 & 27 & 11 & 35 & 15 & 14 & 41 & 24 & 16 \\
-% \specialcellleft{Credentials not valuable \\ to me} & 14 & 9 & 18 & 10 & 10 & 18 & 25 & 22 & 4 & 5 & 18 & 7 & 20 & 22 & 14 & 14 & 25 & 14 & 10 \\
-% \bottomrule
-% \end{tabular}
-% \end{sidewaystable}
+The feedback survey for each course asked the learners, among other questions, whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
+For each reason, the student was asked to select one of two options: "Did influence my decision to stop" and "Did NOT influence my decision to stop".
+Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reportinga reason and those who express that the reason did not apply to them.
+We are interested in examining the strength of correlations between these reasons.
+We computed the correlation matrix for the reasons for each course separately, and then linearly combined these matrices to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.
 
-Table: correlations between reasons and one with cors between reasons and intentions
+\begin{table*}
+\label{tab:study1_rfd_corr}
+\caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
+\small
+\center
+\begin{tabular}{lrrrrrrrrrrrrrr}
+\toprule
+ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
+\midrule
+R1: Assn or exam was too difficult & 1.0 & 0.17 & 0.28 & 0.3 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.2 & 0.09 & 0.33 & 0.4 & 0.28 \\
+R2: Couldn't keep up with deadlines &  & 1.0 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
+R3: Credentials not valuable to me &  &  & 1.0 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
+R4: Didn't enjoy online course format &  &  &  & 1.0 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
+R5: I was only exploring &  &  &  &  & 1.0 & 0.36 & 0.2 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
+R6: I learned all I wanted &  &  &  &  &  & 1.0 & 0.12 & 0.31 & 0.28 & 0.0 & 0.12 & 0.16 & 0.16 & 0.09 \\
+R7: Material not explained properly &  &  &  &  &  &  & 1.0 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
+R8: Not challenging enough &  &  &  &  &  &  &  & 1.0 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
+R9: Not what I was looking for &  &  &  &  &  &  &  &  & 1.0 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
+R10: Course requires too much time &  &  &  &  &  &  &  &  &  & 1.0 & 0.06 & 0.08 & 0.19 & 0.12 \\
+R11: I joined too late &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.22 & 0.11 & 0.14 \\
+R12: Technical difficulties &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.27 & 0.3 \\
+R13: Too advanced for my background &  &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.33 \\
+R14: Videos were confusing &  &  &  &  &  &  &  &  &  &  &  &  &  & 1 \\
+\bottomrule
+\end{tabular}
+\end{table*}
 
-Plot: ecdf curves for proportion videos watched for each disengagement reason
+In the initial surveys, learners were asked about their reasons for enrollment, and they were asked to mark whether or not each of a set of enrollment reasons \cite{} applies to them.
+For each course, we computed the correlation matrix of the reasons for dropout with reasons for enrollment.
+These matrices were then linearly combined to give the matrix shown in Table \ref{tab:study1_rfd_ei_corr}.
+
+\begin{table*}
+\label{tab:study1_rfd_ei_corr}
+\caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
+\small
+\center
+\begin{tabular}{lrrrrrrrrrrrrrr}
+\toprule
+ & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
+\midrule
+Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
+Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
+Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
+General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
+Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
+Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
+Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
+Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
+Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
+Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
+Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
+Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
+Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
+Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+% Proportion videos watched for each disengagement reason
+We are also interested in the correlations between reported reasons for dropout and certain learner behaviors.
+For each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
+We then computed the matrix of correlations between reasons for dropout and these variables for each course.
+These matrices were averaged and the result is shown in Table \ref{tab:study1_rfd_video_behavior_corr}.
+
+\subsection{Results}
 

### 2014-10-27 paper_v1.Rnw
Study 1 results

@@
 We then computed the matrix of correlations between reasons for dropout and these variables for each course.
-These matrices were averaged and the result is shown in Table \ref{tab:study1_rfd_video_behavior_corr}.
+These matrices were averaged and the result is shown in Table [TableToBeAdded].
 
 \subsection{Results}
+Study 1 shows that the two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
+The correlation between the two reasons is $r=0.32$.
+A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.
+
+Difficulty of an assignment or exam was the third most reported reason for dropout.
+It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).
+
+The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses.
+This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons.
+[What might we like to study more about respondents who selected this reason?]
+
+In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
+
+Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting ....
+[TBC]
+
+The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
+it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$)
+
+Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
+Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}
+
+\begin{table*}
+\label{tab:rfd_ei_strongest_correlations}
+\caption{Strongest observed correlations ($r >= 0.15) between reasons for disengagement and enrollment intentions}
+\center
+\small
+\begin{tabular}{llr}
+\toprule
+ Enrollment intention & Dropout reason & $r$ \\
+\midrule
+Take with others & Course not challenging enough & 0.24 \\
+Take with others & Too advanced for my background & 0.20 \\
+Experience an online course & I was only exploring the course & 0.17 \\
+Improve my English & Technical difficulties & 0.17 \\
+Personal growth & Could not keep up with deadlines & 0.16 \\
+Take with others & Credentials not valuable enough for me & 0.16 \\
+Take with others & Technical difficulties & 0.15 \\
+Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
+\bottomrule
+\end{tabular}
+\end{table*}
 
 \subsection{Discussion}
+Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward.
+Expectency-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain.
+In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks.
+Because learners can join or leave MOOCs freely.
+Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts.
+Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification.
+So what does lack of time for the course mean?
+In study 2, we attempt to investigate various possible dropout reasons that learners might report as "lack of time".
+
+[TBC]
 

### 2014-10-27 paper_v1.Rnw
some updates to s1

@@
 \begin{abstract}
-Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses.
+% Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses.
 Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions.
@@
 (abstract needs work...)
+Abstract goes here.
 \end{abstract}
@@
 
-Educational environments has become increasingly diverse.
+Educational environments have become increasingly diverse.
 Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit.
@@
 
+\begin{table*}[th]
+\label{tab:s1sum}
+\caption{Summary statistics of learner demographics and behavior in 14 MOOCs}
+\small
+\center
+\begin{tabular}{llccccccccccc}
+% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
+% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
+\toprule
+ &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
+ \cmidrule(r){4-6} \cmidrule(r){11-13}
+ & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
+\midrule
+C1 & Business & 61,096   & 9,943  & 2,367    & N    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
+C2 & Business & 41,081   & 6,544  & 883      & N    & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
+C3 & CS       & 27,104   & 6,104 & 2,854     & N    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
+C4 & CS       & 56,154   & 3,406 & 1,123     & N    & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
+C5 & CS       & 57,636   & 5738  & 402       & N    & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
+C6 & Engineering & 16,220 & 2,540 & 269      & N    & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
+C7 & Enginerring & 7,178 & 1,312 & 202       & N    & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
+C8 & Humanities & 7,634 & 3,631  & 152       & N    & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
+C9 & Humanities & 6,475 & 3,315  & 326       & N    & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
+C10 & Humanities & 9,903 & 4,227 & 1,570     & N    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
+C11 & Humanities & 630 & 167     & 80        & N    & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
+C12 & Math \& Stats & 49,317 & 3,090 & 862   & N    & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
+C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & N   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
+C14 & Math \& Stats & 16,857 & 2,892 & 363   & N    & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
+C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & N   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
+C16 & Math \& Stats & 27,236 & 2,711 & 637   & N    & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
+C17 & Math \& Stats & 13,132 & 3,629 & 843   & N    & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
+\bottomrule
+\multicolumn{13}{l}{$^1$ mean (SD)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
+\end{tabular}
+\end{table*}
+
 \subsection{MOOCs}
@@
 
-% \begin{description}
-%   \item[H1] prior experience -- technical difficulties
-%   \item[H2] demographics (education) -- course difficulty
-% \end{description}
+The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs.
+We address the research questions developed in the previous section with a combination of self-report and behavioral data from seventeen diverse MOOCs.
+Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments.
+This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf.
+Study 2).
 
@@
 
-% Which courses did we use? What data
-Seventeen MOOCs on topcis in a variety of disciplines were selected for investigation: four on topics in the humanities, two in engineering, three in computer science, two runs of the same course in business, and six in mathematics and statistics.
+Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}).
 For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
-Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey).
-Table \ref{tab:study1_courses_summary} provides the following summary statistics for each course: the number of enrolled learners, the number of learners who completed the course survey and the feedback surveys, age, gender, and education level, and the proportion of learners who self-reported that they stopped participating in the course.
-Note that age, gender, and education level statistics were computed using only the subset of learners who took both the initial and feedback surveys.
-We also computed the proprtions of learners who viewed at least 30, 50, and 80\% of the course videos.
+% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and behavioral indicators for each course.
+Self-reported dropout was measured in the feedback survey by asking learners if they stopped participating in the course.
+The three behavioral indicators are the proportions of learners who viewed at least 30, 50, and 80\% of the lecture videos.
+Information about age, gender, and education level were only available for learners who completed the initial course survey.
 
@@
 
-\begin{table*}
-\label{tab:study1_courses_summary}
-\caption{Summary statistics of 14 MOOCs}
-\small
-\center
-\begin{tabular}{llcccccccccc}
-\toprule
-%  & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & \specialcellcentered{\ensuremath{>} 30\% \\ videos} & \specialcellcentered{\ensuremath{>} 50\% \\ videos} & \specialcellcentered{\ensuremath{>} 80\% \\ videos} & \specialcellcentered{Self-Rep.
-\\ Dropout} \\
-& & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
-& Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
-\midrule
-C1 & Business & 61,096 & 9,943 & 2,367 & 42\% & 39 (30,49) & n/a & 39\% & 20\% & 7\% & 10\% \\
-C2 & Business & 41,081 & 6,544 & 883 & 40\% & 39 (30,49) & 97\% & 32\% & 13\% & 2\% & 9\% \\
-C3 & CS & 27,104 & 6,104 & 2,854 & 35\% & 30 (22,41) & 74\% & 67\% & 40\% & 7\% & 2\% \\
-C4 & CS & 56,154 & 3,406 & 1,123 & 11\% & 33 (27,42) & 91\% & 34\% & 16\% & 11\% & 21\% \\
-C5 & CS & 57,636 & 5738 & 402 & 17\% & 33 (26,43) & 91\% & 66\% & 55\% & 28\% & 14\% \\
-C6 & Engineering & 16,220 & 2,540 & 269 & 13\% & 33 (27,45) & 92\% & 62\% & 49\% & 17\% & 15\% \\
-C7 & Enginerring & 7,178 & 1,312 & 202 & 12\% & 27 (22,39) & 85\% & 51\% & 18\% & 2\% & 7\% \\
-C8 & Humanities & 7,634 & 3,631 & 152 & 92\% & 33 (26,45) & 93\% & 45\% & 11\% & 0\% & 7\% \\
-C9 & Humanities & 6,475 & 3,315 & 326 & 90\% & 32 (27,47) & 91\% & 69\% & 50\% & 17\% & 16\% \\
-C10 & Humanities & 9,903 & 4,227 & 1,570 & 52\% & 39 (29,52) & 89\% & 84\% & 74\% & 21\% & 3\% \\
-C11 & Humanities & 630 & 167 & 80 & 50\% & 30 (27,40) & 97\% & 74\% & 55\% & 16\% & 16\% \\
-C12 & Math \& Stats & 49,317 & 3,090 & 862 & 33\% & 41 (29,55) & 88\% & 27\% & 10\% & 1\% & 31\% \\
-C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 58\% & 28 (16,46) & n/a & 69\% & 38\% & 8\% & 4\% \\
-C14 & Math \& Stats & 16,857 & 2,892 & 363 & 9\% & 31 (26,38) & 99\% & 61\% & 45\% & 25\% & 14\% \\
-C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 20\% & 35 (29,45) & 98\% & 78\% & 67\% & 34\% & 8\% \\
-C16 & Math \& Stats & 27,236 & 2,711 & 637 & 82\% & 43 (34,51) & n/a & 64\% & 36\% & 5\% & 1\% \\
-C17 & Math \& Stats & 13,132 & 3,629 & 843 & n/a & n/a & n/a & 70\% & 54\% & 14\% & 5\% \\
-\bottomrule
-\end{tabular}
-\end{table*}
+\subsubsection{Measures}
 
-% Proportions for each dropout reason for each course
-We are interested in the proportions of respondents that reported each reason for dropout across the courses.
-These proportions are shown in Figure \ref{fig:prop-reason}.
-The proprtion for each reason was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it aplies or that it does not apply).
+Learners reported reasons for enrolling in the course (using the OLEI scale \cite{kizilcec2015motivation} on the course survey) and leaving the course (on the feedback survey) on a grid of binary-choice items.
+The feedback survey for asked learners whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
+For each reason, the student was asked to select one of two options: ``Did influence my decision to stop" and ``Did NOT influence my decision to stop".
+Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reporting reasons and those who express that the reason did not apply to them.
+We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior.
+
+We first report the proportion for each reason for dropping out.
+This was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it applies or that it does not apply).
 We also computed the median and interquartile range of the proprtions across all courses for each reason.
+The resulting proportions are shown in Figure \ref{fig:prop-reason}.
+Second, we performed three correlational analyses were performed to investigate associations between these measures.
 
-%Figure: proportions for each dropout reason for each course
+We examined the strength of correlations between reasons for dropout.
+The correlation matrix for the reasons were computed separately for each course, and then linearly combined to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.
+Following the same procedure, we computed the correlation matrix between reasons to enroll and reasons for attrition.
+Furthermore, we computed correlations between reported reasons for dropout and certain learner behaviors.
+Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
+We then computed the matrix of correlations between reasons for dropout and these variables for each course.
+These matrices were averaged and the result is shown in Table [TableToBeAdded].
+
+\subsection{Results}
+
 <<prop-reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
@@
 @
-
-The feedback survey for each course asked the learners, among other questions, whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
-For each reason, the student was asked to select one of two options: "Did influence my decision to stop" and "Did NOT influence my decision to stop".
-Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reportinga reason and those who express that the reason did not apply to them.
-We are interested in examining the strength of correlations between these reasons.
-We computed the correlation matrix for the reasons for each course separately, and then linearly combined these matrices to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.
 
@@
 
-In the initial surveys, learners were asked about their reasons for enrollment, and they were asked to mark whether or not each of a set of enrollment reasons \cite{} applies to them.
-For each course, we computed the correlation matrix of the reasons for dropout with reasons for enrollment.
-These matrices were then linearly combined to give the matrix shown in Table \ref{tab:study1_rfd_ei_corr}.
-
 \begin{table*}
@@
 
-% Proportion videos watched for each disengagement reason
-We are also interested in the correlations between reported reasons for dropout and certain learner behaviors.
-For each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
-We then computed the matrix of correlations between reasons for dropout and these variables for each course.
-These matrices were averaged and the result is shown in Table [TableToBeAdded].
+(Note: this section is work in progress, we will go through each RQ and present results that address them here.)
 
-\subsection{Results}
-Study 1 shows that the two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
+Table \ref{tab:study1_rfd_ei_corr}.
+
+How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
+
+What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
+
+What reasons do learners report for dropping out of a MOOC? (RQ3a)
+
+ How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b)
+
+How does behavior in MOOCs vary by reasons for dropping out? (RQ4)
+
+The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
 The correlation between the two reasons is $r=0.32$.
@@
 
-\begin{table*}
-\label{tab:rfd_ei_strongest_correlations}
-\caption{Strongest observed correlations ($r >= 0.15) between reasons for disengagement and enrollment intentions}
-\center
-\small
-\begin{tabular}{llr}
-\toprule
- Enrollment intention & Dropout reason & $r$ \\
-\midrule
-Take with others & Course not challenging enough & 0.24 \\
-Take with others & Too advanced for my background & 0.20 \\
-Experience an online course & I was only exploring the course & 0.17 \\
-Improve my English & Technical difficulties & 0.17 \\
-Personal growth & Could not keep up with deadlines & 0.16 \\
-Take with others & Credentials not valuable enough for me & 0.16 \\
-Take with others & Technical difficulties & 0.15 \\
-Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
-\bottomrule
-\end{tabular}
-\end{table*}
+% \begin{table*}
+% \label{tab:rfd_ei_strongest_correlations}
+% \caption{Strongest observed correlations ($r >= 0.15$) between reasons for disengagement and enrollment intentions}
+% \center
+% \small
+% \begin{tabular}{llr}
+% \toprule
+%  Enrollment intention & Dropout reason & $r$ \\
+% \midrule
+% Take with others & Course not challenging enough & 0.24 \\
+% Take with others & Too advanced for my background & 0.20 \\
+% Experience an online course & I was only exploring the course & 0.17 \\
+% Improve my English & Technical difficulties & 0.17 \\
+% Personal growth & Could not keep up with deadlines & 0.16 \\
+% Take with others & Credentials not valuable enough for me & 0.16 \\
+% Take with others & Technical difficulties & 0.15 \\
+% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 
 \subsection{Discussion}
+
+(Note: this section is also work in progress.)
+
 Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward.
@@
 So what does lack of time for the course mean?
-In study 2, we attempt to investigate various possible dropout reasons that learners might report as "lack of time".
-
-[TBC]
+In study 2, we attempt to investigate various possible dropout reasons that learners might report as ``lack of time".
 

### 2014-10-27 paper_v1.tex
some updates to s1

--- 2014-10-26 paper_v1.tex
@@
 \begin{abstract}
-Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from [N] online learners in [M] courses.
+% Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses.
 Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions.
@@
 
-TODO: Outline different courses and go through variabels in the course info table.
+Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}).
+For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and behavioral indicators for each course.
+Self-reported dropout was measured in the feedback survey by asking learners if they stopped participating in the course.
+The three behavioral indicators are the proportions of learners who viewed at least 30, 50, and 80\% of the lecture videos.
+Information about age, gender, and education level were only available for learners who completed the initial course survey.
 
-Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+%Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+
+\subsubsection{Measures}
+
+Learners reported reasons for enrolling in the course (using the OLEI scale \cite{kizilcec2015motivation} on the course survey) and leaving the course (on the feedback survey) on a grid of binary-choice items.
+The feedback survey for asked learners whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
+For each reason, the student was asked to select one of two options: ``Did influence my decision to stop" and ``Did NOT influence my decision to stop".
+Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reporting reasons and those who express that the reason did not apply to them.
+We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior.
+
+We first report the proportion for each reason for dropping out.
+This was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it applies or that it does not apply).
+We also computed the median and interquartile range of the proprtions across all courses for each reason.
+The resulting proportions are shown in Figure \ref{fig:prop-reason}.
+Second, we performed three correlational analyses were performed to investigate associations between these measures.
+
+We examined the strength of correlations between reasons for dropout.
+The correlation matrix for the reasons were computed separately for each course, and then linearly combined to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.
+Following the same procedure, we computed the correlation matrix between reasons to enroll and reasons for attrition.
+Furthermore, we computed correlations between reported reasons for dropout and certain learner behaviors.
+Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
+We then computed the matrix of correlations between reasons for dropout and these variables for each course.
+These matrices were averaged and the result is shown in Table [TableToBeAdded].
 
@@
 
-Table: proportions for each dropout reason for each course
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
 
-Table: correlations between reasons and one with cors between reasons and intentions
+\includegraphics[width=\maxwidth]{figure/prop-reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:prop-reason}}
+\end{figure*}
 
-Plot: ecdf curves for proportion videos watched for each disengagement reason
+\end{knitrout}
+
+\begin{table*}
+\label{tab:study1_rfd_corr}
+\caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
+\small
+\center
+\begin{tabular}{lrrrrrrrrrrrrrr}
+\toprule
+ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
+\midrule
+R1: Assn or exam was too difficult & 1.0 & 0.17 & 0.28 & 0.3 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.2 & 0.09 & 0.33 & 0.4 & 0.28 \\
+R2: Couldn't keep up with deadlines &  & 1.0 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
+R3: Credentials not valuable to me &  &  & 1.0 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
+R4: Didn't enjoy online course format &  &  &  & 1.0 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
+R5: I was only exploring &  &  &  &  & 1.0 & 0.36 & 0.2 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
+R6: I learned all I wanted &  &  &  &  &  & 1.0 & 0.12 & 0.31 & 0.28 & 0.0 & 0.12 & 0.16 & 0.16 & 0.09 \\
+R7: Material not explained properly &  &  &  &  &  &  & 1.0 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
+R8: Not challenging enough &  &  &  &  &  &  &  & 1.0 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
+R9: Not what I was looking for &  &  &  &  &  &  &  &  & 1.0 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
+R10: Course requires too much time &  &  &  &  &  &  &  &  &  & 1.0 & 0.06 & 0.08 & 0.19 & 0.12 \\
+R11: I joined too late &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.22 & 0.11 & 0.14 \\
+R12: Technical difficulties &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.27 & 0.3 \\
+R13: Too advanced for my background &  &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.33 \\
+R14: Videos were confusing &  &  &  &  &  &  &  &  &  &  &  &  &  & 1 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}
+\label{tab:study1_rfd_ei_corr}
+\caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
+\small
+\center
+\begin{tabular}{lrrrrrrrrrrrrrr}
+\toprule
+ & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
+\midrule
+Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
+Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
+Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
+General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
+Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
+Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
+Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
+Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
+Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
+Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
+Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
+Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
+Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
+Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+(Note: this section is work in progress, we will go through each RQ and present results that address them here.)
+
+Table \ref{tab:study1_rfd_ei_corr}.
+
+How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
+
+What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
+
+What reasons do learners report for dropping out of a MOOC? (RQ3a)
+
+ How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b)
+
+How does behavior in MOOCs vary by reasons for dropping out? (RQ4)
+
+The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
+The correlation between the two reasons is $r=0.32$.
+A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.
+
+Difficulty of an assignment or exam was the third most reported reason for dropout.
+It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).
+
+The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses.
+This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons.
+[What might we like to study more about respondents who selected this reason?]
+
+In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
+
+Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting ....
+[TBC]
+
+The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
+it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$)
+
+Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
+Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}
+
+% \begin{table*}
+% \label{tab:rfd_ei_strongest_correlations}
+% \caption{Strongest observed correlations ($r >= 0.15$) between reasons for disengagement and enrollment intentions}
+% \center
+% \small
+% \begin{tabular}{llr}
+% \toprule
+%  Enrollment intention & Dropout reason & $r$ \\
+% \midrule
+% Take with others & Course not challenging enough & 0.24 \\
+% Take with others & Too advanced for my background & 0.20 \\
+% Experience an online course & I was only exploring the course & 0.17 \\
+% Improve my English & Technical difficulties & 0.17 \\
+% Personal growth & Could not keep up with deadlines & 0.16 \\
+% Take with others & Credentials not valuable enough for me & 0.16 \\
+% Take with others & Technical difficulties & 0.15 \\
+% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 
 \subsection{Discussion}
+
+(Note: this section is also work in progress.)
+
+Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward.
+Expectency-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain.
+In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks.
+Because learners can join or leave MOOCs freely.
+Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts.
+Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification.
+So what does lack of time for the course mean?
+In study 2, we attempt to investigate various possible dropout reasons that learners might report as ``lack of time".
 

### 2014-10-27 paper_v1.Rnw
started editing results

@@
 % Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and behavioral indicators for each course.
-Self-reported dropout was measured in the feedback survey by asking learners if they stopped participating in the course.
-The three behavioral indicators are the proportions of learners who viewed at least 30, 50, and 80\% of the lecture videos.
-Information about age, gender, and education level were only available for learners who completed the initial course survey.
+On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
+On the feedback survey, learners reported their satisfaction with the course (N-item index, $\alpha=.XX, M=XX, SD=XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.}
 
-%Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
+The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, and 80\% of lecture videos in the course.
+The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
+The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
 
-\subsubsection{Measures}
-
-Learners reported reasons for enrolling in the course (using the OLEI scale \cite{kizilcec2015motivation} on the course survey) and leaving the course (on the feedback survey) on a grid of binary-choice items.
-The feedback survey for asked learners whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
-For each reason, the student was asked to select one of two options: ``Did influence my decision to stop" and ``Did NOT influence my decision to stop".
-Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reporting reasons and those who express that the reason did not apply to them.
-We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior.
-
-We first report the proportion for each reason for dropping out.
-This was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it applies or that it does not apply).
-We also computed the median and interquartile range of the proprtions across all courses for each reason.
-The resulting proportions are shown in Figure \ref{fig:prop-reason}.
-Second, we performed three correlational analyses were performed to investigate associations between these measures.
-
-We examined the strength of correlations between reasons for dropout.
-The correlation matrix for the reasons were computed separately for each course, and then linearly combined to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.
-Following the same procedure, we computed the correlation matrix between reasons to enroll and reasons for attrition.
-Furthermore, we computed correlations between reported reasons for dropout and certain learner behaviors.
+TODO, update based on what we do here: We computed correlations between reported reasons for dropout and certain learner behaviors.
 Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
@@
 
+% Information about age, gender, and education level were only available for learners who completed the initial course survey.
+% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior
+
 \subsection{Results}
 
-<<prop-reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
+Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.
+The distribution of proportions across fourteen MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
+Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
+Other reasons for attrition were less prevalent, though reported by around 10\% to 25\% of respondents.
+
+<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
 values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
@@
 \begin{table*}
-\label{tab:study1_rfd_corr}
+\label{tab:s1reascor}
 \caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
@@
 What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
-
-What reasons do learners report for dropping out of a MOOC? (RQ3a)
 

### 2014-10-27 paper_v1.tex
started editing results

@@
 % Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and behavioral indicators for each course.
-Self-reported dropout was measured in the feedback survey by asking learners if they stopped participating in the course.
-The three behavioral indicators are the proportions of learners who viewed at least 30, 50, and 80\% of the lecture videos.
-Information about age, gender, and education level were only available for learners who completed the initial course survey.
+On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
+On the feedback survey, learners reported their satisfaction with the course (N-item index, $\alpha=.XX, M=XX, SD=XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to a Yes/No or True/False scale.}
 
-%Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
+The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, and 80\% of lecture videos in the course.
+The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
+The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
 
-\subsubsection{Measures}
-
-Learners reported reasons for enrolling in the course (using the OLEI scale \cite{kizilcec2015motivation} on the course survey) and leaving the course (on the feedback survey) on a grid of binary-choice items.
-The feedback survey for asked learners whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop.
-For each reason, the student was asked to select one of two options: ``Did influence my decision to stop" and ``Did NOT influence my decision to stop".
-Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reporting reasons and those who express that the reason did not apply to them.
-We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior.
-
-We first report the proportion for each reason for dropping out.
-This was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it applies or that it does not apply).
-We also computed the median and interquartile range of the proprtions across all courses for each reason.
-The resulting proportions are shown in Figure \ref{fig:prop-reason}.
-Second, we performed three correlational analyses were performed to investigate associations between these measures.
-
-We examined the strength of correlations between reasons for dropout.
-The correlation matrix for the reasons were computed separately for each course, and then linearly combined to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.
-Following the same procedure, we computed the correlation matrix between reasons to enroll and reasons for attrition.
-Furthermore, we computed correlations between reported reasons for dropout and certain learner behaviors.
+TODO, update based on what we do here: We computed correlations between reported reasons for dropout and certain learner behaviors.
 Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
@@
 
+% Information about age, gender, and education level were only available for learners who completed the initial course survey.
+% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior
+
 \subsection{Results}
+
+Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.
+The distribution of proportions across fourteen MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
+Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
+Other reasons for attrition were less prevalent, though reported by around 10\% to 25\% of respondents.
 
@@
 
-\includegraphics[width=\maxwidth]{figure/prop-reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:prop-reason}}
+\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
 \end{figure*}

### 2014-10-28 paper_v1.Rnw
edits based on andreas' feedback

@@
 
-This paper presents a systematic investigation into attrition in MOOCs, based on self-report and behavioral data collected from [N] online learners in [M] courses.
+This paper presents a systematic investigation into attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses.
+A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition and thereby enables targeted interventions to help learners who want to persist.
 We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes.
 Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions.
-In Study 2, we sought to develop a deeper understanding of attrition in MOOCs by conducting a case study.
-Targeting around 7,000 online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey.
-TODO: complete paragraph based on what we actually do in Study 2.
+In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback.
+Their open response answers about what challenges they faced while taking the MOOC were iteratively coded.
+We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.
 
@@
 Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
-Other reasons for attrition were less prevalent, though reported by around 10\% to 25\% of respondents.
+Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents.
+Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.
+
+[.] What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
 
@@
 \center
-\begin{tabular}{lrrrrrrrrrrrrrr}
+\begin{tabular}{lccccccccccccc}
 \toprule
- & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
+ & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
 \midrule
-R1: Assn or exam was too difficult & 1.0 & 0.17 & 0.28 & 0.3 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.2 & 0.09 & 0.33 & 0.4 & 0.28 \\
-R2: Couldn't keep up with deadlines &  & 1.0 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
-R3: Credentials not valuable to me &  &  & 1.0 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
-R4: Didn't enjoy online course format &  &  &  & 1.0 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
-R5: I was only exploring &  &  &  &  & 1.0 & 0.36 & 0.2 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
-R6: I learned all I wanted &  &  &  &  &  & 1.0 & 0.12 & 0.31 & 0.28 & 0.0 & 0.12 & 0.16 & 0.16 & 0.09 \\
-R7: Material not explained properly &  &  &  &  &  &  & 1.0 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
-R8: Not challenging enough &  &  &  &  &  &  &  & 1.0 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
-R9: Not what I was looking for &  &  &  &  &  &  &  &  & 1.0 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
-R10: Course requires too much time &  &  &  &  &  &  &  &  &  & 1.0 & 0.06 & 0.08 & 0.19 & 0.12 \\
-R11: I joined too late &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.22 & 0.11 & 0.14 \\
-R12: Technical difficulties &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.27 & 0.3 \\
-R13: Too advanced for my background &  &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.33 \\
-R14: Videos were confusing &  &  &  &  &  &  &  &  &  &  &  &  &  & 1 \\
+1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
+2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
+3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
+4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
+5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
+6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
+7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
+8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
+9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
+10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
+11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
+12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
+13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
+14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
 \bottomrule
@@
 
-\begin{table*}
-\label{tab:study1_rfd_ei_corr}
-\caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
-\small
-\center
-\begin{tabular}{lrrrrrrrrrrrrrr}
-\toprule
- & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
-\midrule
-Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
-Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
-Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
-General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
-Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
-Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
-Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
-Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
-Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
-Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
-Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
-Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
-Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
-Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
-\bottomrule
-\end{tabular}
-\end{table*}
+[.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
+
+We expected reasons for disengaging from the course to be related to the motivations that drove the learner to initially enroll in the course.
+A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations.
+Taking the course with friends or colleagues was correlated with disengaging because the course is not challenging enough, the material is too advanced of the learner's background, and the credential is not valuable enough to the learner ($r=.24, .20, .16$, respectively).
+Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$).
+Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
+
+% \begin{table}
+% \label{tab:rfd_ei_strongest_correlations}
+% \caption{Strongest correlations between enrollment intentions and reasons for disengaging ($|r| \geq 0.15$)}
+% \center
+% \small
+% \begin{tabular}{llr}
+% \toprule
+%  Enrollment Intention & Reason for Disengaging & $r$ \\
+% \midrule
+% Take with others & Course not challenging enough & 0.24 \\
+% Take with others & Too advanced for my background & 0.20 \\
+% Take with others & Credentials not valuable enough for me & 0.16 \\
+% Take with others & Technical difficulties & 0.15 \\
+% Experience an online course & I was only exploring the course & 0.17 \\
+% Improve my English & Technical difficulties & 0.17 \\
+% Personal growth & Could not keep up with deadlines & 0.16 \\
+% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
+% \bottomrule
+% \end{tabular}
+% \end{table}
+
+% \begin{table*}
+% \label{tab:study1_rfd_ei_corr}
+% \caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
+% \small
+% \center
+% \begin{tabular}{lrrrrrrrrrrrrrr}
+% \toprule
+%  & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
+% \midrule
+% Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
+% Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
+% Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
+% General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
+% Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
+% Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
+% Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
+% Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
+% Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
+% Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
+% Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
+% Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
+% Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
+% Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 
@@
 
-Table \ref{tab:study1_rfd_ei_corr}.
+[.] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b) <- we may not be able to answer this
 
-How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
-
-What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
-
- How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b)
-
-How does behavior in MOOCs vary by reasons for dropping out? (RQ4)
+[.] How does behavior in MOOCs vary by reasons for dropping out? (RQ4) <- we need to see if we have enough data to answer this/if this is interesting
 
@@
 
-% \begin{table*}
-% \label{tab:rfd_ei_strongest_correlations}
-% \caption{Strongest observed correlations ($r >= 0.15$) between reasons for disengagement and enrollment intentions}
-% \center
-% \small
-% \begin{tabular}{llr}
-% \toprule
-%  Enrollment intention & Dropout reason & $r$ \\
-% \midrule
-% Take with others & Course not challenging enough & 0.24 \\
-% Take with others & Too advanced for my background & 0.20 \\
-% Experience an online course & I was only exploring the course & 0.17 \\
-% Improve my English & Technical difficulties & 0.17 \\
-% Personal growth & Could not keep up with deadlines & 0.16 \\
-% Take with others & Credentials not valuable enough for me & 0.16 \\
-% Take with others & Technical difficulties & 0.15 \\
-% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
-% \bottomrule
-% \end{tabular}
-% \end{table*}
-
 \subsection{Discussion}
 
-(Note: this section is also work in progress.)
-
 Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward.
-Expectency-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain.
+Expectancy-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain.
 In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks.
@@
 To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey.
-One goal of this study was to further investiage research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced.
+One goal of this study was to further investigate research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced.
 Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time.
 A variety of factors can influence the amount of time learners have at their disposal.
-Higher priority obligations, such as professional, academic, and personal committments, are certainly large contributors.
+Higher priority obligations, such as professional, academic, and personal committments, are presumably large contributors.
 Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}.
@@
 There were  20,048 enrolled learners; 10,510 watched at least one video, and [N] attempted more than one assignment.
+The following subsections describe how disengaging learners were identified, which measures were included in the feedback survey, and how open responses were iteratively coded.
 
@@
 These items served as measures of personal success.
-% In addition, they reported intentions for engaging with each week's course content. 
 They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of.
@@
 Two research assistants independently developed codebooks for the resulting 448 non-empty open responses.
-Their codebooks were consolidated and applied on Mechanical Turk in three iterations.
+Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations.
 The codebook was updated in each iteration to reliably fit the open response data.
 Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement.
-In the first iteration, 250 randomly selected responses were each coded by four `classification experts'.
+In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification.
 In the second iteration, the remaining responses were coded in the same way.
@@
 A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).
+Hence, if a learner had not enough time, this tended to be the main challenge.
 
@@
 Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
-We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e.
+We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e.
 their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
@@
 Table \ref{tab:s2reas}).
-The large majority of learners who indicate that they had not have enough time was found to be sperable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$).
+The large majority of learners who indicate that they had not enough time was found to be separable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$).
 MOOCs are an environment that requires learners to be highly self-regulated to be successful.

### 2014-10-28 paper_v1.tex
edits based on andreas' feedback

@@
 If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to a Yes/No or True/False scale.}
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.}
 

### 2014-10-28 paper_v1.Rnw
made edits based on emily's comments

@@
 Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes.
-Course materials, including assessments, were delivered through mail (``correspondence education''), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.
+Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.
 
@@
 And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}.
-Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} (other?) resemble those on online media platforms, such as YouTube or tumblr.
+Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} resemble those on online media platforms.
 This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.
@@
 Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}.
-Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs} (other?), motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested.
+Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested.
 Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized.
-Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one world is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other world adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}.
+Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}.
 Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition.
+The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.
 
-This paper presents a systematic investigation into attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses.
-A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition and thereby enables targeted interventions to help learners who want to persist.
+This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses.
+A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
 We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes.
@@
 The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting.
+The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course.
 An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
@@
 \begin{description}
-  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC?
+  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
   \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
@@
 \begin{description}
-  \item[RQ2] What proportion of learners who dropped out were satisfied with their achievements in the course?
+  \item[RQ2] What proportion of learners who disengaged were satisfied with their achievements in the course?
   \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
@@
 \begin{description}
-\item[RQ3a] What reasons do learners report for dropping out of a MOOC?
-\item[RQ3b] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)?
-\item[RQ4] How does behavior in MOOCs vary by reasons for dropping out?
+\item[RQ3a] What reasons do learners report for disengaging from a MOOC?
+\item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
+\item[RQ4] How does behavior in MOOCs vary by reasons for disengaging?
 \end{description}
@@
 
-(Note: this section is work in progress, we will go through each RQ and present results that address them here.)
-
 [.] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b) <- we may not be able to answer this
@@
 
-The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
+% The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
 The correlation between the two reasons is $r=0.32$.
 A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.
-
 Difficulty of an assignment or exam was the third most reported reason for dropout.
 It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).
-
 The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses.
 This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons.
-[What might we like to study more about respondents who selected this reason?]
-
-In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
-
+[What might we like to study more about respondents who selected this reason?] In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
 Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting ....
-[TBC]
-
-The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
-it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$)
-
-Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
+[TBC] The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
+it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$) Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
 Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}
@@
 % We thank Elise Ogle and Ruth Bram for helping with the development of the codebook. 
+% Thank Emily for feedback and Andreas for feedback and data support.
 Omitted for blind review.

### 2014-10-28 paper_v1.Rnw
abstract and restructuring

@@
 \begin{abstract}
-% Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses.
-Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions.
-Study 2 is a case study to develop a deeper understanding of attrition in MOOCs by conducting a case study.
-Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey.
-(abstract needs work...)
-Abstract goes here.
+Research on attrition has a long tradition in the field of education.
+Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for diverse motivations.
+A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses suggested time to be a critical obstacle and [other important take-aways from Study 1] (Study 1; $N = XX,XXX$).
+A follow-up study with online learners who disengaged from the course (Study 2; $ N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset.
+Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%).
+A low capacity for volitional control was identified as a large contributor to this issue, besides demands from higher priority obligations.
+A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
 \end{abstract}
@@
 Novel institutional structures have been developed to overcome particular constraints.
-Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accommodating people with less flexible schedules \cite{goldrick2010challenges}.
+Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}.
 Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes.
@@
 We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.
-
-\section{Related Work}
-
-Research and theorizing on attrition in education has a rich history.
-This review is intended to serve as a foundation to build on with the current research.
-We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work.
-The focus of this brief review is on how ways of thinking about attrition have developed over the last decades.
-
-\subsection{In-person Education}
-
-The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting.
-The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course.
-An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
-The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}.
-Hence, motivation alone is necessary but not sufficient for persistence.
-Students may fail to sustain efforts in the absence of strong self-regulatory skills.
-
-\begin{description}
-  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
-  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
-\end{description}
-
-The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution.
-Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college.
-While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
-Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
-This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2007question}.
-
-\begin{description}
-  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
-\end{description}
-
-Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration.
-Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.
-Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (i.e., instilling a growth mindset) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}.
-The results highlight the powerful influence of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.
-
-\begin{description}
-  \item[RQ2] What proportion of learners who disengaged were satisfied with their achievements in the course?
-  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
-\end{description}
-
-\subsection{Distance Education and e-Learning}
-
-Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
-However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner.
-Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
-Persistence is thought to be a function of background characteristics (e.g., demographics), academic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
-The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
-
-Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs.
-Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning.
-And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
-An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
-An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.
 
@@
 
+\section{Related Work}
+
+Research and theorizing on attrition in education has a rich history.
+This review is intended to serve as a foundation to build on with the current research.
+We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work.
+The focus of this brief review is on how ways of thinking about attrition have developed over the last decades.
+
+\subsection{In-person Education}
+
+The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting.
+The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course.
+An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
+The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}.
+Hence, motivation alone is necessary but not sufficient for persistence.
+Students may fail to sustain efforts in the absence of strong self-regulatory skills.
+
+\begin{description}
+  \item[RQ1] How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
+  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
+\end{description}
+
+The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution.
+Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college.
+While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
+Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
+This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2007question}.
+
+\begin{description}
+  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
+\end{description}
+
+Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration.
+Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.
+Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (i.e., instilling a growth mindset) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}.
+The results highlight the powerful influence of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.
+
+\begin{description}
+  \item[RQ2a] What proportion of learners who disengaged were satisfied with their achievements in the course?
+  \item[RQ2b] How does learners' behavior in the MOOC vary by their level of satisfaction?
+  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
+\end{description}
+
+\subsection{Distance Education and e-Learning}
+
+Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
+However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner.
+Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
+Persistence is thought to be a function of background characteristics (e.g., demographics), academic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
+The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
+
+Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs.
+Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning.
+And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
+An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
+An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.
+
 \subsection{MOOCs}
@@
 \item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
-\item[RQ4] How does behavior in MOOCs vary by reasons for disengaging?
 \end{description}
-
-\section{Study 1: Reasons for Disengaging}
-
-The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs.
-We address the research questions developed in the previous section with a combination of self-report and behavioral data from seventeen diverse MOOCs.
-Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments.
-This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf.
-Study 2).
-
-\subsection{Methods}
-
-Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}).
-For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
-% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
-On the feedback survey, learners reported their satisfaction with the course (N-item index, $\alpha=.XX, M=XX, SD=XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.}
-
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
-The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, and 80\% of lecture videos in the course.
-The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
-The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
-
-TODO, update based on what we do here: We computed correlations between reported reasons for dropout and certain learner behaviors.
-Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
-We then computed the matrix of correlations between reasons for dropout and these variables for each course.
-These matrices were averaged and the result is shown in Table [TableToBeAdded].
-
-% Information about age, gender, and education level were only available for learners who completed the initial course survey.
-% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior
-
-\subsection{Results}
-
-Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.
-The distribution of proportions across fourteen MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
-Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
-Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents.
-Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.
-
-[.] What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
 
@@
 
-\begin{table*}
-\label{tab:s1reascor}
-\caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
-\small
-\center
-\begin{tabular}{lccccccccccccc}
-\toprule
- & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
-\midrule
-1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
-2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
-3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
-4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
-5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
-6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
-7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
-8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
-9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
-10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
-11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
-12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
-13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
-14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
-\bottomrule
-\end{tabular}
-\end{table*}
+\section{Study 1: Reasons for Disengaging}
 
-[.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
+The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs.
+We address the research questions developed in the previous section with a combination of self-report and behavioral data from seventeen diverse MOOCs.
+Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments.
+This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf.
+Study 2).
+
+\subsection{Methods}
+
+Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}).
+For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
+On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
+On the feedback survey, learners reported their satisfaction with the course (N-item index, M=XX, SD=XX, $\alpha=.XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.}
+
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
+The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, and 80\% of lecture videos in the course.
+The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
+The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
+
+TODO, update based on what we do here: We computed correlations between reported reasons for dropout and certain learner behaviors.
+Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
+We then computed the matrix of correlations between reasons for dropout and these variables for each course.
+These matrices were averaged and the result is shown in Table [TableToBeAdded].
+
+% Information about age, gender, and education level were only available for learners who completed the initial course survey.
+% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior
+
+\subsection{Results}
+
+Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.
+The distribution of proportions across fourteen MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
+Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
+Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents.
+Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.
+
+Report results of factor analysis here.
 
@@
 Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
+
+Attrition can be a sign of success if a learner has achieved their personal goal.
+We therefore investigated what proportion of learners who dropped out were satisfied with their achievements in the course (RQ2).
+TODO add satisfaction analysis here
+
+% \begin{table*}
+% \label{tab:s1reascor}
+% \caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
+% \small
+% \center
+% \begin{tabular}{lccccccccccccc}
+% \toprule
+%  & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
+% \midrule
+% 1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
+% 2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
+% 3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
+% 4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
+% 5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
+% 6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
+% 7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
+% 8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
+% 9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
+% 10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
+% 11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
+% 12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
+% 13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
+% 14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
+
+% [.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
 
@@
 % \end{table*}
-
-[.] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b) <- we may not be able to answer this
-
-[.] How does behavior in MOOCs vary by reasons for dropping out? (RQ4) <- we need to see if we have enough data to answer this/if this is interesting
 
@@
 
-To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time.
-It revealed that no particular commitment was specified in about half of them (53\%).
-This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders.
-Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
-We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e.
-their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
-However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
-These findings lend strong support to hypotheses $H4a$ and $H4b$: that is, while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
-
-\subsubsection{Psychological Factors}
-
-Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
-Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
-The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
-Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
-Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
-Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress.
-The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction.
-Hence, the results fully support hypothesis $H1$, and partially support $H2$ and $H3$.
-
 \begin{table*}[ht]
@@
 \end{table*}
+
+To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time.
+It revealed that no particular commitment was specified in about half of them (53\%).
+This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders.
+Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
+We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e.
+their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
+However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
+These findings lend strong support to hypotheses $H4a$ and $H4b$: that is, while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
+
+\subsubsection{Psychological Factors}
+
+Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
+Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
+The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
+Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
+Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
+Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress.
+The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction.
+Hence, the results fully support hypothesis $H1$, and partially support $H2$ and $H3$.
 

### 2014-10-28 paper_v1.tex
abstract and restructuring

@@
 \item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
-\item[RQ4] How does behavior in MOOCs vary by reasons for disengaging?
 \end{description}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
+
+\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
+\end{figure*}
+
+\end{knitrout}
 
@@
 On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
-On the feedback survey, learners reported their satisfaction with the course (N-item index, $\alpha=.XX, M=XX, SD=XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+On the feedback survey, learners reported their satisfaction with the course (N-item index, M=XX, SD=XX, $\alpha=.XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
 If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
@@
 
-[.] What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
-
-\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
-\end{figure*}
-
-\end{knitrout}
-
-\begin{table*}
-\label{tab:s1reascor}
-\caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
-\small
-\center
-\begin{tabular}{lccccccccccccc}
-\toprule
- & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
-\midrule
-1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
-2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
-3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
-4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
-5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
-6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
-7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
-8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
-9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
-10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
-11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
-12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
-13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
-14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-[.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
+Report results of factor analysis here.
 

### 2014-10-28 paper_v1.Rnw
more restructuring

@@
 
+\subsubsection{Predictors}
+
+RQ1: How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
+
+TODO Rene adds these analyses here
+
+\subsubsection{Satisfaction}
+
+RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
+
+RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?
+
+Attrition can be a sign of success if a learner has achieved their personal goal.
+We therefore investigated what proportion of learners who dropped out were satisfied with their achievements in the course (RQ2).
+
+TODO Sherif adds satisfaction analyses here
+
+\subsubsection{Reasons}
+
+RQ3a: What reasons do learners report for disengaging from a MOOC?
+
+RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
+
 Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.
@@
 Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
-
-Attrition can be a sign of success if a learner has achieved their personal goal.
-We therefore investigated what proportion of learners who dropped out were satisfied with their achievements in the course (RQ2).
-TODO add satisfaction analysis here
 

### 2014-10-29 paper_v1.Rnw
added results of factor analysis

@@
 
-Report results of factor analysis here.
+To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis.
+The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
+The following five factors emerged: first, general time issues (course requires too much time; can't keep up with deadlines); second, late start (starting course late); third, difficulty (too advanced, exam too difficult); fourth, format and content (didn't enjoy online format, confusing videos, materials not explained well), and fifth, expectations (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
+Only the technical difficulties item failed to load onto any factor.
 

### 2014-10-29 paper_v1.tex
structured S1 results

@@
 
-\section{Study 1: Reasons for Disengaging}
+\section{Study 1: Understanding Attrition}
 
@@
 
-\subsubsection{Predictors}
-
-RQ1: How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
+\subsubsection{Predictors of Attrition}
+% RQ1: How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
 
@@
 
-\subsubsection{Satisfaction}
-
-RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
-
-RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?
+\subsubsection{Satisfaction: Personal Success}
+% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
+% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?
 
@@
 
-\subsubsection{Reasons}
-
-RQ3a: What reasons do learners report for disengaging from a MOOC?
-
-RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
+\subsubsection{Reasons for Disengaging}
+% RQ3a: What reasons do learners report for disengaging from a MOOC?
+% RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
 
@@
 
-We expected reasons for disengaging from the course to be related to the motivations that drove the learner to initially enroll in the course.
+We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course.
 A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations.
@@
 
-\section{Conclusion}
-
 \section{Acknowledgments}

### 2014-10-29 paper_v1.Rnw
edited abstract

@@
 Research on attrition has a long tradition in the field of education.
-Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for diverse motivations.
+Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in motivation.
 A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses suggested time to be a critical obstacle and [other important take-aways from Study 1] (Study 1; $N = XX,XXX$).
@@
 Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%).
-A low capacity for volitional control was identified as a large contributor to this issue, besides demands from higher priority obligations.
-A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
+A low capacity for volitional control was identified as a large contributor to this issue.
+A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
 \end{abstract}

### 2014-10-29 paper_v1.Rnw
First satisfaction results

--- 2014-10-29 paper_v1.Rnw
@@
 \documentclass{sigchi}
+
+<<loadcsv, echo=FALSE>>=
+#library(RCurl)
+#data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+#a = read.csv(text=data)
+a = read.csv("C:/Research/Analytics/pycode/user_code/halawa/publications/LAS2015/data_masters/compiled_data_all_courses.csv")
+
+# Shift scores for satisfaction index components up to 1-N
+a$qoe = a$qoe + 4
+a$satis_amount_learnt = a$satis_amount_learnt + 4
+a$satis_vs_time_spent = a$satis_vs_time_spent + 4
+a$satis_time_mgmt = a$satis_time_mgmt + 4
+a$how_met_expectations = a$how_met_expectations + 3
+a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
+
+satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
+a$satisfaction = rowMeans(satM, na.rm=TRUE)
+a$satisfaction[!is.finite(a$satisfaction)] = NA
+
+library(plyr)
+a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
+@
 
@@
 \category{K.3.1.}{Computers and Education}{Computer Uses in Education}
+
+<<test1, echo=TRUE>>=
+@
 
@@
 
-TODO Sherif adds satisfaction analyses here
+First, we compared the distributions of satisfaction scores for learners who reported they stopped participating and those who reported not to have stopped.
+As figure \ref{fig:study1_satis_si_score_dist_comparison} shows, the mean satisfaction score ($mean=4.70$) for learners who reported to have stopped is 0.64 points lower than the other group ($mean=5.34$).
+
+<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
+y0 = a$satisfaction[a$did_dropout == "No"]
+y1 = a$satisfaction[a$did_dropout == "Yes"]
+b = a[!is.na(a$did_dropout),]
+ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
+@
 

### 2014-10-29 paper_v1.Rnw
adding pred analysis

--- 2014-10-29 paper_v1.Rnw
@@
 Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
-The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, and 80\% of lecture videos in the course.
-The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
+The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, 80\% of lecture videos and attempted over 30\%, 50\%, 80\% of assessments in the course.
+In addition, to capture performance, we  The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
 The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
@@
 % RQ1: How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
+
+<<>>=
+res = read.csv("pred_coefs.csv")
+ggplot(subset(res, coef2!="Intercept"), aes(coef2, est, ymin=est-1.96*se, ymax=est+1.96*se, color=factor(sign))) + 
+  geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
+  theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
+  scale_color_manual(values=c('red','grey30','darkgreen')) +
+  labs(y="Estimated change in probability (95% CIs)", x="")
+@
 

### 2014-10-30 paper_v1.Rnw
More satisfaction results

--- 2014-10-29 paper_v1.Rnw
@@
 
-<<loadcsv, echo=FALSE>>=
-#library(RCurl)
-#data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-#a = read.csv(text=data)
-a = read.csv("C:/Research/Analytics/pycode/user_code/halawa/publications/LAS2015/data_masters/compiled_data_all_courses.csv")
+<<loadcsv, echo=FALSE, warning=FALSE>>=
+library(bitops)
+library(RCurl)
+data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+a = read.csv(text=data)
 
@@
 First, we compared the distributions of satisfaction scores for learners who reported they stopped participating and those who reported not to have stopped.
-As figure \ref{fig:study1_satis_si_score_dist_comparison} shows, the mean satisfaction score ($mean=4.70$) for learners who reported to have stopped is 0.64 points lower than the other group ($mean=5.34$).
+A significant difference was observed ($t_{1469}=20.7, p \ll .0001, d=0.7$).
+The mean satisfaction score in the group that reported stopped (no stopped) participation was 4.70 (5.34).
 
-<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
-y0 = a$satisfaction[a$did_dropout == "No"]
-y1 = a$satisfaction[a$did_dropout == "Yes"]
+<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=5, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
+#b = a[!is.na(a$did_dropout),]
+#ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
+@
+
+We also examined differences in satisfaction between learners grouped by certain behaviors.
+First, we classified learners into 4 groups based on the fraction of course videos they viewed.
+The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}.
+The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group.
+Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.
+
+<<study1_satsfaction_by_vid_groups, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
 b = a[!is.na(a$did_dropout),]
-ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
+course_ids = unique(a$course_id)
+course_num_videos = list()
+for (cid in course_ids) {
+  course_num_videos[[cid]] = quantile(a$num_videos_watched[a$course_id == cid], 0.95, na.rm=TRUE)
+}
+for(i in 1:nrow(b)) {
+  x = b[i,]
+  f = x[['num_videos_watched']]/course_num_videos[[x[['course_id']]]]
+  if (is.na(f)) {level=NA}
+  else if (f <= 0.25) {level = "0-25%"}
+  else if  (f <= 0.50) {level = "25-50%"}
+  else if  (f <= 0.75) {level = "50-75%"}
+  else {level = "75-100%"}
+  b[i,"fraction_videos_watched"] = level
+}
+b$fraction_videos_watched = factor(b$fraction_videos_watched)
+c = b[!is.na(b$fraction_videos_watched),]
+ggplot(c) + geom_boxplot(aes(x=c$fraction_videos_watched,y=c$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Fraction videos watched")
 @

### 2014-10-30 paper_v1.Rnw
added pred results

--- 2014-10-29 paper_v1.Rnw
@@
 #library(RCurl)
-#data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-#a = read.csv(text=data)
-a = read.csv("C:/Research/Analytics/pycode/user_code/halawa/publications/LAS2015/data_masters/compiled_data_all_courses.csv")
+# data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+# a = read.csv(text=data)
+a = read.csv("compiled_data_all_courses.csv")
 
@@
 
-<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
-values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
-57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
-21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
-32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
-15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
-21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
-69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
-12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
-18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
-13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
-16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
-14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
-26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
-14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)
-
-reasons = c("Material not explained properly","Course requires too much time","Not what I was looking for","I joined too late","Didn't enjoy online course format","Technical difficulties","Couldn't keep up with deadlines","Not challenging enough","I learned all I wanted","Videos were confusing","I was only exploring","Too advanced for my background","Assignment or exam was too difficult","Credentials not valuable to me")
-
-dat = data.frame(reasons = rep(reasons, each=17), values)
-
-library(ggplot2)
-ggplot(dat, aes(reorder(reasons, values, median), values)) + geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion",x="")
-@
-
 \section{Study 1: Understanding Attrition}
@@
 % Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
-On the feedback survey, learners reported their satisfaction with the course (N-item index, M=XX, SD=XX, $\alpha=.XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.}
+On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course.
+On the feedback survey, learners reported their satisfaction with the course (6-item index, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see reasons in Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.} Geographic location was determined by IP address.
 
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
-The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, 80\% of lecture videos and attempted over 30\%, 50\%, 80\% of assessments in the course.
-In addition, to capture performance, we  The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
-The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
-
-TODO, update based on what we do here: We computed correlations between reported reasons for dropout and certain learner behaviors.
-Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
-We then computed the matrix of correlations between reasons for dropout and these variables for each course.
-These matrices were averaged and the result is shown in Table [TableToBeAdded].
-
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
+Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
+A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 14 MOOCs in our dataset.
+Out of the 65,320 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
+Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.
+% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
+% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
 % Information about age, gender, and education level were only available for learners who completed the initial course survey.
@@
 
+We address the research questions developed in the previous section in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across subsets of learners ($RQ2ab$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3ab$).
+
 \subsubsection{Predictors of Attrition}
-% RQ1: How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
 
-<<>>=
+We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features.
+Correlations between predictors were low enough to allow for simple interpretation of coefficients.
+The 14 MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.
+Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model.
+The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.
+
+A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% points less likely than men to persist with lectures and assessments.
+Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men.
+Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America (N.A.).
+Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in N.A..
+While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from N.A..
+
+Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
+More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to perform highly on them.
+Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
+Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.
+
+<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop.
+videos watched, prop.
+assessments completed, and grade percentile).
+The baseline (intercept) are males in N.
+America of average age who have not earned a bachelors\', no prior experience in the topic, not intending to complete the course, prepared to spend under 2 hours/week, this being their first MOOC.
+An estimate of -20 indicates a 20\\% points lower probability of achieving a milestone.'>>=
+library(ggplot2)
 res = read.csv("pred_coefs.csv")
-ggplot(subset(res, coef2!="Intercept"), aes(coef2, est, ymin=est-1.96*se, ymax=est+1.96*se, color=factor(sign))) + 
+res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
+res$y = factor(res$y, labels=c("> 30% videos", "> 50% videos", "> 80% videos", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))
+res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
+ggplot(subset(res, coef2!="Intercept"), 
+       aes(coef2, 100*est, ymin=100*(est-1.96*se), ymax=100*(est+1.96*se), color=factor(sign))) + 
   geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
   theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
-  scale_color_manual(values=c('red','grey30','darkgreen')) +
-  labs(y="Estimated change in probability (95% CIs)", x="")
+  scale_color_manual(values=c('red','grey30','green3')) +
+  labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
 @
-
-TODO Rene adds these analyses here
 
@@
 Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
+
+<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
+values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
+57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
+21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
+32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
+15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
+21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
+69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
+12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
+18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
+13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
+16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
+14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
+26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
+14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)
+
+reasons = c("Material not explained properly","Course requires too much time","Not what I was looking for","I joined too late","Didn't enjoy online course format","Technical difficulties","Couldn't keep up with deadlines","Not challenging enough","I learned all I wanted","Videos were confusing","I was only exploring","Too advanced for my background","Assignment or exam was too difficult","Credentials not valuable to me")
+
+dat = data.frame(reasons = rep(reasons, each=17), values)
+
+library(ggplot2)
+ggplot(dat, aes(reorder(reasons, values, median), values)) + geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion",x="")
+@
 

### 2014-10-30 paper_v1.tex
added pred results

@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
-
-\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
-\end{figure*}
-
-\end{knitrout}
-
 \section{Study 1: Understanding Attrition}
@@
 % Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-On the initial course surveys, learners reported their gender, age, prior experience, and enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}).
-On the feedback survey, learners reported their satisfaction with the course (N-item index, M=XX, SD=XX, $\alpha=.XX$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-If they indicated that they stopped participating, learners reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.}
+On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course.
+On the feedback survey, learners reported their satisfaction with the course (6-item index, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see reasons in Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.} Geographic location was determined by IP address.
 
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and three behavioral indicators for each course.
-The behavioral indicators represent course milestones: the proportions of learners who viewed over 30\%, 50\%, and 80\% of lecture videos in the course.
-The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
-The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
-
-TODO, update based on what we do here: We computed correlations between reported reasons for dropout and certain learner behaviors.
-Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes).
-We then computed the matrix of correlations between reasons for dropout and these variables for each course.
-These matrices were averaged and the result is shown in Table [TableToBeAdded].
-
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
+Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
+A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 14 MOOCs in our dataset.
+Out of the 65,320 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
+Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.
+% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
+% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
 % Information about age, gender, and education level were only available for learners who completed the initial course survey.
@@
 
+We address the research questions developed in the previous section in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across subsets of learners ($RQ2ab$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3ab$).
+
 \subsubsection{Predictors of Attrition}
-% RQ1: How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
 
-TODO Rene adds these analyses here
+We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features.
+Correlations between predictors were low, which allows for simple interpretations of coefficients.
+The 14 MOOCs were modeled as a random effect, given our goal of generalizing to the population of similar MOOCs.
+Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model.
+The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.
+
+A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% points less likely than men to persist with lectures and assessments.
+Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men.
+Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America (N.A.).
+Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in N.A..
+While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from N.A..
+Besides geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
+More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to perform highly on them.
+Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
+Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
+
+\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop.
+videos watched, prop.
+assessments completed, and grade percentile).
+The baseline (intercept) are males in N.
+America of average age who have not earned a bachelors', no prior experience in the topic, not intending to complete the course, prepared to spend under 2 hours/week, this being their first MOOC.
+An estimate of -20 indicates a 20\% points lower probability of achieving a milestone.\label{fig:s1coefs}}
+\end{figure*}
+
+\end{knitrout}
 
@@
 
-TODO Sherif adds satisfaction analyses here
+First, we compared the distributions of satisfaction scores for learners who reported they stopped participating and those who reported not to have stopped.
+As figure \ref{fig:study1_satis_si_score_dist_comparison} shows, the mean satisfaction score ($mean=4.70$) for learners who reported to have stopped is 0.64 points lower than the other group ($mean=5.34$).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\color{warningcolor}{\#\# Warning: Removed 4 rows containing non-finite values (stat\_boxplot).}}\end{kframe}\begin{figure*}[ht]
+
+\includegraphics[width=\maxwidth]{figure/study1_satis_si_score_dist_comparison} \caption[Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course]{Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course\label{fig:study1_satis_si_score_dist_comparison}}
+\end{figure*}
+
+\end{knitrout}
 
@@
 Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
+
+\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
+\end{figure*}
+
+\end{knitrout}
 

### 2014-10-30 paper_v1.Rnw

@@
 
-<<loadcsv, echo=FALSE>>=
-#library(RCurl)
-# data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-# a = read.csv(text=data)
-a = read.csv("compiled_data_all_courses.csv")
+<<loadcsv, echo=FALSE, warning=FALSE>>=
+library(bitops)
+library(RCurl)
+data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+a = read.csv(text=data)
 

### 2014-10-30 paper_v1.Rnw
updated S1 results

--- 2014-10-30 paper_v1.Rnw
@@
 
-<<loadcsv, echo=FALSE>>=
-#library(RCurl)
-# data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-# a = read.csv(text=data)
-a = read.csv("compiled_data_all_courses.csv")
-
-# Shift scores for satisfaction index components up to 1-N
-a$qoe = a$qoe + 4
-a$satis_amount_learnt = a$satis_amount_learnt + 4
-a$satis_vs_time_spent = a$satis_vs_time_spent + 4
-a$satis_time_mgmt = a$satis_time_mgmt + 4
-a$how_met_expectations = a$how_met_expectations + 3
-a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
-
-satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
-a$satisfaction = rowMeans(satM, na.rm=TRUE)
-a$satisfaction[!is.finite(a$satisfaction)] = NA
-
-library(plyr)
-a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
-@
+% 
+% <<loadcsv, echo=FALSE, warning=FALSE>>=
+% library(bitops)
+% library(RCurl)
+% data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+% a = read.csv(text=data)
+% 
+% # Shift scores for satisfaction index components up to 1-N
+% a$qoe = a$qoe + 4
+% a$satis_amount_learnt = a$satis_amount_learnt + 4
+% a$satis_vs_time_spent = a$satis_vs_time_spent + 4
+% a$satis_time_mgmt = a$satis_time_mgmt + 4
+% a$how_met_expectations = a$how_met_expectations + 3
+% a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
+% 
+% satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
+% library(psych)
+% 
+% a$satisfaction = rowMeans(satM, na.rm=TRUE)
+% a$satisfaction[!is.finite(a$satisfaction)] = NA
+% a$sat
+% d=subset(a, select=c('course_id', 'num_videos_watched', 'did_dropout', 'satisfaction'))
+% write.csv(d, "satisfaction.csv", row.names=F)
+% library(plyr)
+% a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
+% @
 
@@
 \bottomrule
-\multicolumn{13}{l}{$^1$ mean (SD)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
+\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
 \end{tabular}
@@
 Attrition can be a sign of success if a learner has achieved their personal goal.
-We therefore investigated what proportion of learners who dropped out were satisfied with their achievements in the course (RQ2).
+We therefore investigated to what extent learners were satisfied with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$).
+To account for differences across courses, we compared satisfaction scores using mixed-effects models.
 
-First, we compared the distributions of satisfaction scores for learners who reported they stopped participating and those who reported not to have stopped.
-As figure \ref{fig:study1_satis_si_score_dist_comparison} shows, the mean satisfaction score ($mean=4.70$) for learners who reported to have stopped is 0.64 points lower than the other group ($mean=5.34$).
+Learners who reported that they stopped participating were significantly 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$).
+Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied.
+While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=12, p<.001$), it was a relatively weak effect.
+Those who watched all videos were only around 3\% more satisfied than those who watched none, based on a linear fit.
 
-<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
-y0 = a$satisfaction[a$did_dropout == "No"]
-y1 = a$satisfaction[a$did_dropout == "Yes"]
-b = a[!is.na(a$did_dropout),]
-ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
+<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=5, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
+#b = a[!is.na(a$did_dropout),]
+#ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
+@
+
+% We also examined differences in satisfaction between learners grouped by certain behaviors.
+First, we classified learners into 4 groups based on the fraction of course videos they viewed.
+The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}.
+The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group.
+Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.
+
+<<echo=FALSE,eval=FALSE>>=
+library(dplyr)
+sat = read.csv("satisfaction.csv")
+sat$num_videos_watched[is.na(sat$num_videos_watched)] = 0
+sat = sat %>% group_by(course_id) %>% 
+  mutate(prop_vid = num_videos_watched / quantile(num_videos_watched[num_videos_watched!=0], .90))
+
+sat$prop_vid4 = cut(sat$prop_vid, breaks=c(-Inf, .25, .5, .75, Inf))
+
+require(lmerTest)
+summary(lmer(satisfaction~ did_dropout + (1|course_id), sat))
+5.170/(5.170-0.4456)
+
+summary(lmer(satisfaction~0+prop_vid4 + (1|course_id), sat))
+summary(lmer(satisfaction~prop_vid + (1|course_id), sat))
+(5.005+0.1679)/5.005
 @

### 2014-10-30 paper_v1.tex
updated S1 results

@@
 \usepackage{alltt}
+
+% 
+% <<loadcsv, echo=FALSE, warning=FALSE>>=
+% library(bitops)
+% library(RCurl)
+% data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+% a = read.csv(text=data)
+% 
+% # Shift scores for satisfaction index components up to 1-N
+% a$qoe = a$qoe + 4
+% a$satis_amount_learnt = a$satis_amount_learnt + 4
+% a$satis_vs_time_spent = a$satis_vs_time_spent + 4
+% a$satis_time_mgmt = a$satis_time_mgmt + 4
+% a$how_met_expectations = a$how_met_expectations + 3
+% a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
+% 
+% satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
+% library(psych)
+% 
+% a$satisfaction = rowMeans(satM, na.rm=TRUE)
+% a$satisfaction[!is.finite(a$satisfaction)] = NA
+% a$sat
+% d=subset(a, select=c('course_id', 'num_videos_watched', 'did_dropout', 'satisfaction'))
+% write.csv(d, "satisfaction.csv", row.names=F)
+% library(plyr)
+% a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
+% @
 
@@
 We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features.
-Correlations between predictors were low, which allows for simple interpretations of coefficients.
-The 14 MOOCs were modeled as a random effect, given our goal of generalizing to the population of similar MOOCs.
+Correlations between predictors were low enough to allow for simple interpretation of coefficients.
+The 14 MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.
 Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model.
@@
 While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from N.A..
-Besides geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
+
+Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
 More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to perform highly on them.
@@
 Attrition can be a sign of success if a learner has achieved their personal goal.
-We therefore investigated what proportion of learners who dropped out were satisfied with their achievements in the course (RQ2).
+We therefore investigated to what extent learners were satisfied with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$).
+To account for differences across courses, we compared satisfaction scores using mixed-effects models.
 
-First, we compared the distributions of satisfaction scores for learners who reported they stopped participating and those who reported not to have stopped.
-As figure \ref{fig:study1_satis_si_score_dist_comparison} shows, the mean satisfaction score ($mean=4.70$) for learners who reported to have stopped is 0.64 points lower than the other group ($mean=5.34$).
+Learners who reported that they stopped participating were significantly 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$).
+Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied.
+While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=12, p<.001$), it was a relatively weak effect.
+Those who watched all videos were only around 3\% more satisfied than those who watched none, based on a linear fit.
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning: Removed 4 rows containing non-finite values (stat\_boxplot).}}\end{kframe}\begin{figure*}[ht]
-
-\includegraphics[width=\maxwidth]{figure/study1_satis_si_score_dist_comparison} \caption[Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course]{Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course\label{fig:study1_satis_si_score_dist_comparison}}
-\end{figure*}
-
-\end{knitrout}
+% We also examined differences in satisfaction between learners grouped by certain behaviors.
+First, we classified learners into 4 groups based on the fraction of course videos they viewed.
+The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}.
+The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group.
+Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.
 

### 2014-10-30 paper_v1.Rnw
wrote S1 discussion, edited abstract

@@
 \documentclass{sigchi}
-
-% 
-% <<loadcsv, echo=FALSE, warning=FALSE>>=
-% library(bitops)
-% library(RCurl)
-% data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-% a = read.csv(text=data)
-% 
-% # Shift scores for satisfaction index components up to 1-N
-% a$qoe = a$qoe + 4
-% a$satis_amount_learnt = a$satis_amount_learnt + 4
-% a$satis_vs_time_spent = a$satis_vs_time_spent + 4
-% a$satis_time_mgmt = a$satis_time_mgmt + 4
-% a$how_met_expectations = a$how_met_expectations + 3
-% a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
-% 
-% satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
-% library(psych)
-% 
-% a$satisfaction = rowMeans(satM, na.rm=TRUE)
-% a$satisfaction[!is.finite(a$satisfaction)] = NA
-% a$sat
-% d=subset(a, select=c('course_id', 'num_videos_watched', 'did_dropout', 'satisfaction'))
-% write.csv(d, "satisfaction.csv", row.names=F)
-% library(plyr)
-% a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
-% @
 
@@
 
-\title{Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses}
+\title{Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors}
 
@@
 Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in motivation.
-A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses suggested time to be a critical obstacle and [other important take-aways from Study 1] (Study 1; $N = XX,XXX$).
+A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1).
+Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence.
 A follow-up study with online learners who disengaged from the course (Study 2; $ N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset.
@@
 The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
-The following five factors emerged: first, general time issues (course requires too much time; can't keep up with deadlines); second, late start (starting course late); third, difficulty (too advanced, exam too difficult); fourth, format and content (didn't enjoy online format, confusing videos, materials not explained well), and fifth, expectations (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
+The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em  difficulty} (too advanced, exam too difficult); fourth, {\em format and content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
 Only the technical difficulties item failed to load onto any factor.
@@
 
-<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
+<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
 values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
@@
 
-% \begin{table*}
-% \label{tab:s1reascor}
-% \caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
-% \small
-% \center
-% \begin{tabular}{lccccccccccccc}
-% \toprule
-%  & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
-% \midrule
-% 1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
-% 2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
-% 3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
-% 4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
-% 5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
-% 6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
-% 7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
-% 8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
-% 9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
-% 10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
-% 11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
-% 12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
-% 13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
-% 14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
-% \bottomrule
-% \end{tabular}
-% \end{table*}
-
-% [.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
-
-% \begin{table}
-% \label{tab:rfd_ei_strongest_correlations}
-% \caption{Strongest correlations between enrollment intentions and reasons for disengaging ($|r| \geq 0.15$)}
-% \center
-% \small
-% \begin{tabular}{llr}
-% \toprule
-%  Enrollment Intention & Reason for Disengaging & $r$ \\
-% \midrule
-% Take with others & Course not challenging enough & 0.24 \\
-% Take with others & Too advanced for my background & 0.20 \\
-% Take with others & Credentials not valuable enough for me & 0.16 \\
-% Take with others & Technical difficulties & 0.15 \\
-% Experience an online course & I was only exploring the course & 0.17 \\
-% Improve my English & Technical difficulties & 0.17 \\
-% Personal growth & Could not keep up with deadlines & 0.16 \\
-% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
-% \bottomrule
-% \end{tabular}
-% \end{table}
-
-% \begin{table*}
-% \label{tab:study1_rfd_ei_corr}
-% \caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
-% \small
-% \center
-% \begin{tabular}{lrrrrrrrrrrrrrr}
-% \toprule
-%  & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
-% \midrule
-% Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
-% Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
-% Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
-% General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
-% Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
-% Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
-% Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
-% Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
-% Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
-% Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
-% Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
-% Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
-% Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
-% Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
-% \bottomrule
-% \end{tabular}
-% \end{table*}
-
-% The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
-The correlation between the two reasons is $r=0.32$.
-A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.
-Difficulty of an assignment or exam was the third most reported reason for dropout.
-It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).
-The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses.
-This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons.
-[What might we like to study more about respondents who selected this reason?] In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
-Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting ....
-[TBC] The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
-it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$) Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
-Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}
-
 \subsection{Discussion}
 
-Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward.
-Expectancy-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain.
-In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks.
-Because learners can join or leave MOOCs freely.
-Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts.
-Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification.
-So what does lack of time for the course mean?
-In study 2, we attempt to investigate various possible dropout reasons that learners might report as ``lack of time".
+A number of striking findings emerged from the preceeding analyses.
+We found strong evidence for gender inequalities in attrition and, to some extent, performance.
+Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship.
+The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women.
+Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched.
+We also identified a large achievement gap in terms of grades between learners from these continents.
+The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings.
+It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.
+
+Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end.
+Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations.
+General time issues were overall the most pronounced.
+
+Learners may disengage from the course when they become too occupied by higher-priority tasks.
+Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course.
+According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain.
+A low capacity for volitional control \cite{corno2001volitional} could lead learners misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
+This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.
 
@@
 \subsection{Results}
+
+We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control.
+In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
 
@@
 
-% Learners drop out at different times during the course.
-To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis.
-However, learners can develop risk factors at different times in the course.
-The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times.
-Such increased survey load can increase attrition or dissatisfaction in the course.
-
-% An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged.
-This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk).
-Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
-
 % Balancing columns in a ref list is a bit of a pain because you
@@
 \end{document}
+
+% Learners drop out at different times during the course.
+To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis.
+However, learners can develop risk factors at different times in the course.
+The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times.
+Such increased survey load can increase attrition or dissatisfaction in the course.
+
+% An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged.
+This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk).
+Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
+
+% \begin{table*}
+% \label{tab:s1reascor}
+% \caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
+% \small
+% \center
+% \begin{tabular}{lccccccccccccc}
+% \toprule
+%  & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
+% \midrule
+% 1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
+% 2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
+% 3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
+% 4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
+% 5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
+% 6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
+% 7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
+% 8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
+% 9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
+% 10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
+% 11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
+% 12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
+% 13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
+% 14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
+
+% [.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
+
+% \begin{table}
+% \label{tab:rfd_ei_strongest_correlations}
+% \caption{Strongest correlations between enrollment intentions and reasons for disengaging ($|r| \geq 0.15$)}
+% \center
+% \small
+% \begin{tabular}{llr}
+% \toprule
+%  Enrollment Intention & Reason for Disengaging & $r$ \\
+% \midrule
+% Take with others & Course not challenging enough & 0.24 \\
+% Take with others & Too advanced for my background & 0.20 \\
+% Take with others & Credentials not valuable enough for me & 0.16 \\
+% Take with others & Technical difficulties & 0.15 \\
+% Experience an online course & I was only exploring the course & 0.17 \\
+% Improve my English & Technical difficulties & 0.17 \\
+% Personal growth & Could not keep up with deadlines & 0.16 \\
+% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
+% \bottomrule
+% \end{tabular}
+% \end{table}
+
+% \begin{table*}
+% \label{tab:study1_rfd_ei_corr}
+% \caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
+% \small
+% \center
+% \begin{tabular}{lrrrrrrrrrrrrrr}
+% \toprule
+%  & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
+% \midrule
+% Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
+% Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
+% Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
+% General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
+% Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
+% Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
+% Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
+% Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
+% Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
+% Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
+% Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
+% Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
+% Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
+% Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
+
+% The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
+The correlation between the two reasons is $r=0.32$.
+A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.
+Difficulty of an assignment or exam was the third most reported reason for dropout.
+It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).
+The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses.
+This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons.
+[What might we like to study more about respondents who selected this reason?] In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
+Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting ....
+[TBC] The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
+it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$) Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
+Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}
+
+% 
+% <<loadcsv, echo=FALSE, warning=FALSE>>=
+% library(bitops)
+% library(RCurl)
+% data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
+% a = read.csv(text=data)
+% 
+% # Shift scores for satisfaction index components up to 1-N
+% a$qoe = a$qoe + 4
+% a$satis_amount_learnt = a$satis_amount_learnt + 4
+% a$satis_vs_time_spent = a$satis_vs_time_spent + 4
+% a$satis_time_mgmt = a$satis_time_mgmt + 4
+% a$how_met_expectations = a$how_met_expectations + 3
+% a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
+% 
+% satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
+% library(psych)
+% 
+% a$satisfaction = rowMeans(satM, na.rm=TRUE)
+% a$satisfaction[!is.finite(a$satisfaction)] = NA
+% a$sat
+% d=subset(a, select=c('course_id', 'num_videos_watched', 'did_dropout', 'satisfaction'))
+% write.csv(d, "satisfaction.csv", row.names=F)
+% library(plyr)
+% a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
+% @

### 2014-10-30 paper_v1.tex
wrote S1 discussion, edited abstract

@@
 \usepackage{alltt}
-
-% 
-% <<loadcsv, echo=FALSE, warning=FALSE>>=
-% library(bitops)
-% library(RCurl)
-% data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-% a = read.csv(text=data)
-% 
-% # Shift scores for satisfaction index components up to 1-N
-% a$qoe = a$qoe + 4
-% a$satis_amount_learnt = a$satis_amount_learnt + 4
-% a$satis_vs_time_spent = a$satis_vs_time_spent + 4
-% a$satis_time_mgmt = a$satis_time_mgmt + 4
-% a$how_met_expectations = a$how_met_expectations + 3
-% a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
-% 
-% satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
-% library(psych)
-% 
-% a$satisfaction = rowMeans(satM, na.rm=TRUE)
-% a$satisfaction[!is.finite(a$satisfaction)] = NA
-% a$sat
-% d=subset(a, select=c('course_id', 'num_videos_watched', 'did_dropout', 'satisfaction'))
-% write.csv(d, "satisfaction.csv", row.names=F)
-% library(plyr)
-% a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
-% @
 

### 2014-10-30 paper_v1.Rnw
Table 1 missing column values added

--- 2014-10-30 paper_v1.Rnw
@@
 \midrule
-C1 & Business & 61,096   & 9,943  & 2,367    & N    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
-C2 & Business & 41,081   & 6,544  & 883      & N    & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
-C3 & CS       & 27,104   & 6,104 & 2,854     & N    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
-C4 & CS       & 56,154   & 3,406 & 1,123     & N    & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
-C5 & CS       & 57,636   & 5738  & 402       & N    & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
-C6 & Engineering & 16,220 & 2,540 & 269      & N    & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
-C7 & Enginerring & 7,178 & 1,312 & 202       & N    & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
-C8 & Humanities & 7,634 & 3,631  & 152       & N    & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
-C9 & Humanities & 6,475 & 3,315  & 326       & N    & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
-C10 & Humanities & 9,903 & 4,227 & 1,570     & N    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
-C11 & Humanities & 630 & 167     & 80        & N    & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
-C12 & Math \& Stats & 49,317 & 3,090 & 862   & N    & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
-C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & N   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
-C14 & Math \& Stats & 16,857 & 2,892 & 363   & N    & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
-C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & N   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
-C16 & Math \& Stats & 27,236 & 2,711 & 637   & N    & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
-C17 & Math \& Stats & 13,132 & 3,629 & 843   & N    & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
+C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
+C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
+C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
+C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
+C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
+C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
+C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
+C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
+C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
+C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
+C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
+C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
+C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
+C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
+C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
+C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
+C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
 \bottomrule

### 2014-10-30 paper_v1.Rnw
started general discussion, need to eat
Fixes and changes

@@
 Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized.
-Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}.
+Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}.
 Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition.
@@
 
-This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses.
+This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses.
 A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
@@
 It was offered in 2014 through Coursera.
-There were  20,048 enrolled learners; 10,510 watched at least one video, and [N] attempted more than one assignment.
+There were  20,048 enrolled learners; 10,510 watched at least one video, and 5,912 attempted at least one assignment.
 The following subsections describe how disengaging learners were identified, which measures were included in the feedback survey, and how open responses were iteratively coded.
@@
 
-The objective was to build an `early warning' prediction model that could flag learners who are likely to drop out in the future.
-We defined the following three criteria by which to evaluate the performance of the model.
+The objective was to build an `early warning' prediction model that could flag learners who are likely to have dropped out or to drop out in the future.
+In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.
 
@@
 Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).
-The model was fit by logistic regression.
-Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
+
+To constrain prediction lag, we stopped feeding the model with a learner's activity data once they have been absent for 14 consecutive days.
+During training, if the learner had dropped out but the model had not detected her within 14 days of her last appearnce, the learner is counted as a miss (false negative).
+This construction causes the training algorithm to choose from among features that only operate within the prediction lag constraints we define.
+
+The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
 
@@
 \end{equation}
+
+The almost identical results obtained for $test_1$ and $test_2$ imply good transferability of the model.
 

### 2014-10-30 paper_v1.Rnw
final additions and edits before sumission

@@
 Research on attrition has a long tradition in the field of education.
-Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in motivation.
+Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
 A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1).
 Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence.
-A follow-up study with online learners who disengaged from the course (Study 2; $ N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset.
+A follow-up study with online learners who disengaged from the course (Study 2; $N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset.
 Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%).
@@
 
-This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses.
+This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall.
 A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
@@
 \label{tab:s1sum}
-\caption{Summary statistics of learner demographics and behavior in 14 MOOCs}
+\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
 \small
@@
 \bottomrule
-\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
+\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
 \end{tabular}
@@
 \begin{description}
-  \item[RQ1] How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
   \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
@@
 \begin{description}
+  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from a MOOC?
   \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
@@
 
-\section{Study 1: Understanding Attrition}
+\section{Study 1: Attrition \& Satisfaction}
 
@@
 Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
-A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 14 MOOCs in our dataset.
-Out of the 65,320 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
-Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.
+A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 17 MOOCs in our dataset.
+Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
+Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.}
 % The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
@@
 
-\subsubsection{Predictors of Attrition}
+\subsubsection{Individual Differences in Attrition}
 
@@
 Correlations between predictors were low enough to allow for simple interpretation of coefficients.
-The 14 MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.
-Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model.
+Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{The required data for 3 of the 17 MOOCs could not be obtained in time for submission due to technical issues.} Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model.
 The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.
@@
 Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men.
-Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America (N.A.).
-Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in N.A..
-While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from N.A..
+Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America.
+Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in North America.
+While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from North America.
 
@@
 
-<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop.
+<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop.
 videos watched, prop.
 assessments completed, and grade percentile).
-The baseline (intercept) are males in N.
-America of average age who have not earned a bachelors\', no prior experience in the topic, not intending to complete the course, prepared to spend under 2 hours/week, this being their first MOOC.
+The baseline group were male learners in North America of average age who had not earned a bachelor\'s degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it.
 An estimate of -20 indicates a 20\\% points lower probability of achieving a milestone.'>>=
@@
 
-\subsubsection{Satisfaction: Personal Success}
+\subsubsection{Learner Satisfaction}
 % RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
@@
 
-\subsubsection{Reasons for Disengaging}
+\subsubsection{Reasons for Attrition}
 % RQ3a: What reasons do learners report for disengaging from a MOOC?
@@
 
-Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.
-The distribution of proportions across fourteen MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
+Research question $RQ3a$ about learners' reasons for disengaging from a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.\footnote{This researcher-defined list may not have been collectively exhaustive, which is addressed in Study 2.} The distribution of proportions across 17 MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
 Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
@@
 According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain.
-A low capacity for volitional control \cite{corno2001volitional} could lead learners misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
+A low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
 This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.
 
-\section{Study 2: Understanding Disengagement}
+\section{Study 2: Understanding Attrition}
 
@@
 In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback.
-To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey.
+To this end, we developed a predictive model to identify disengaging learners and asked them to complete a survey.
 One goal of this study was to further investigate research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced.
@@
 Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
-To this end, we included a self-report measures for each construct in the survey that was sent to at-risk learners.
+To this end, we included a self-report measure for each construct in the survey that was sent to disengaging learners.
 
@@
 
-The objective was to build an `early warning' prediction model that could flag learners who are likely to have dropped out or to drop out in the future.
-In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.
+We constructed an `early warning' prediction model to flag learners who had disengaged from the course or were about to disengage.
+The following three criteria were chosen as benchmarks to evaluate the practical and theoretical value of our prediction model.%   In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.
 
-\textit{Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible.
-The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.
+\textit{Prediction accuracy}: Prediction recall (fraction of correctly predicted disengaged learners) should be high to deliver the survey to as many disengaged learners as possible.
+The false positive rate (fraction of active learners who are incorrectly predicted as disengaged) should be low to avoid sending surveys to learners who do not experience challenges in the course.
 
-\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short.
-A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.
+\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as disengaged should be short.
+A shorter prediction lag improves the likelihood of successfully intervening to help learners resume the course.
 
-\textit{Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course.
-We thus had to train the prediction model using data from previous courses.
-The performance of the prediction model should be adequate in courses other than those on which it was trained.
+\textit{Transferability}: Whether or not a learner disengages from the course cannot be determined before the end of the course.
+We thus trained the prediction model using data from previous courses.
+The performance of the prediction model should be good in courses other than those on which it was trained.
 
-The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs.
+The training set was composed of learners' interaction data (predictors) and disengagement states (outcome) from 20 MOOCs.
 The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses.
@@
 A second test set of similar size was composed from 20 other MOOCs ($test_2$).
-Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).
+Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see omitted for blind review%\cite{halawa2014dropout}
+, for additional information on the prediction model).
 
-To constrain prediction lag, we stopped feeding the model with a learner's activity data once they have been absent for 14 consecutive days.
-During training, if the learner had dropped out but the model had not detected her within 14 days of her last appearnce, the learner is counted as a miss (false negative).
-This construction causes the training algorithm to choose from among features that only operate within the prediction lag constraints we define.
-
-The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
+To control the length of the prediction lag, the model was only provided with learners' activity data until they were absent for 14 consecutive days.
+In the training phase, if the learner disengaged but the model did not detect it within 14 days of their last activity, the case was counted as a false negative.
+This construction results in a model with features that all fall within the defined prediction lag period.
+The model was fit by logistic regression with two-fold learner-level and course-level cross-validation.
+Table \ref{tab:predAuc} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
 
 \begin{table}[h!]
-\caption{Performance of the dropout prediction model}
-\label{tab:drop_pred_rocs}
+\caption{Performance of the attrition prediction model}
+\label{tab:predAuc}
 \small
@@
 Equation 1 is an approximation of the model with only these two predictors.
+The fact that almost identical results were obtained for $test_1$ and $test_2$ suggest good transferability of the model.
 A learner was flagged if their dropout probability exceeded a half:
@@
 
-The almost identical results obtained for $test_1$ and $test_2$ imply good transferability of the model.
-
-Out of the 20,048 enrolled learners in the course, we only considered those 10,510 who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out.
-The survey was sent out to 6,053 leaners (excluding 21 learners with invalid email addresses or duplicate accounts).
+Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out.
+The survey was sent out to 6,053 learners (excluding 21 learners with invalid email addresses or duplicate accounts).
 
@@
 And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
-Missing values were multiply imputated by predictive mean matching using responses to all survey questions.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.} All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
-The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.
+Missing values were multiply imputated by predictive mean matching using responses to all survey questions.
+All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
+The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelor's or more advanced degree.
 
@@
 The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}.
-Very few learners explicitly indicated issues with volitional control in their responses, potentially because it is not socially desirable to admit.
+Very few learners explicitly indicated issues with volitional control in their open responses, potentially because it is not socially desirable to admit.
 The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$).
@@
 The methodology used in this study can be extended to deliver timely and targeted interventions.
-The predictive model was successful in identifying a large number of leaners who were facing challenges with taking the course.
+The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course.
 The majority of them were unsuccessful in their own terms.
@@
 
+The present studies extend the empirical base of research on attrition in online learning environments.
+A large-scale study of over 60,000 online learners revealed strong individual differences in persistence and performance by demographic and geographic features.
+Learners who reported that they dropped out were less satisfied with their course achievements.
+Time was a critical obstacle for many learners, and a low capacity for volitional control was found to contribute to this issue, though learners seemingly avoid admitting this.
+Successful online learners, according to their own judgment, were found to exhibit higher levels of goal-striving, social belonging, and growth mindset than subjectively unsuccessful learners.
+We also began to develop a survey item for measuring reasons for disengaging from a course (cf.
+Table \ref{tab:s2reas}), which should be tested and extended, if needed, based on more empirical evidence from additional MOOCs.
+
+This work yields a number of theoretical and practical implications.
+In addressing research questions and testing hypotheses which emerged from a review of the literature, we sought to advance theory on attrition and provide empirical evidence.
+In line with early models of attrition that emphasized the importance of goal-directed behavior, we found goal striving to be higher in subjectively successful than unsuccessful learners.
+Moreover, volitional control was identified as a major contributor in reported time-related challenges with taking a MOOC.
+Tinto's \cite{tinto1975dropout} classic model emphasizes the role of academic and social integration, and also builds on various learner characteristics as contributors to learners' decision to persist in a course of study.
+Our findings provide strong support for the critical role of individual characteristics, such as gender, age, intentions, experiences, and geographic location; and, leveraging the more recent notion of social belonging \cite{walton2007question}, we found evidence in support of Tinto's social integration component.
+Parallel to positive attributions of attrition in community college settings, we found progress in the MOOC and learner satisfaction to only be weakly related: learners who watched very few lectures were almost as satisfied as those who watched most lectures.
+Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
+
+There are many practical implications of this research and the one we would like to emphasize is the need for and feasibility of targeted psychological interventions in online learning environments.
+While the current work does not provide explanations for the presence of substantial achievement gaps, we can speculate that similar psychological processes as in traditional educational environments are involved.
+Brief and targeted online interventions (e.g., mindset, social belonging, self-affirmation, goal striving) can potentially narrow existing achievement gaps in online learning.
+We hope to see more research on developing accurate prediction models that identify struggling learners and suitable interventions to best support them.
+
 \section{Acknowledgments}
-% We thank Elise Ogle and Ruth Bram for helping with the development of the codebook. 
-% Thank Emily for feedback and Andreas for feedback and data support.
+% We thank Elise Ogle and Ruth Bram for helping with the development of the initial codebook. 
+% Thank Emily Schneider for her helpful feedback and Andreas P. for his feedback and data support.
 Omitted for blind review.

### 2014-10-30 paper_v1.tex
started general discussion, need to eat
final additions and edits before sumission

@@
 
-This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses.
+This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall.
 A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
@@
 \midrule
-C1 & Business & 61,096   & 9,943  & 2,367    & N    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
-C2 & Business & 41,081   & 6,544  & 883      & N    & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
-C3 & CS       & 27,104   & 6,104 & 2,854     & N    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
-C4 & CS       & 56,154   & 3,406 & 1,123     & N    & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
-C5 & CS       & 57,636   & 5738  & 402       & N    & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
-C6 & Engineering & 16,220 & 2,540 & 269      & N    & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
-C7 & Enginerring & 7,178 & 1,312 & 202       & N    & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
-C8 & Humanities & 7,634 & 3,631  & 152       & N    & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
-C9 & Humanities & 6,475 & 3,315  & 326       & N    & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
-C10 & Humanities & 9,903 & 4,227 & 1,570     & N    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
-C11 & Humanities & 630 & 167     & 80        & N    & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
-C12 & Math \& Stats & 49,317 & 3,090 & 862   & N    & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
-C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & N   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
-C14 & Math \& Stats & 16,857 & 2,892 & 363   & N    & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
-C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & N   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
-C16 & Math \& Stats & 27,236 & 2,711 & 637   & N    & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
-C17 & Math \& Stats & 13,132 & 3,629 & 843   & N    & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
+C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
+C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
+C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
+C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
+C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
+C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
+C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
+C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
+C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
+C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
+C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
+C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
+C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
+C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
+C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
+C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
+C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
 \bottomrule
-\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
+\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
 \end{tabular}
@@
 
-\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop.
+\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop.
 videos watched, prop.
 assessments completed, and grade percentile).
-The baseline (intercept) are males in N.
-America of average age who have not earned a bachelors', no prior experience in the topic, not intending to complete the course, prepared to spend under 2 hours/week, this being their first MOOC.
+The baseline group were male learners in North America of average age who had not earned a bachelor's degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it.
 An estimate of -20 indicates a 20\% points lower probability of achieving a milestone.\label{fig:s1coefs}}
@@
 
-The objective was to build an `early warning' prediction model that could flag learners who are likely to drop out in the future.
-We defined the following three criteria by which to evaluate the performance of the model.
+We constructed an `early warning' prediction model to flag learners who had disengaged from the course or were about to disengage.
+The following three criteria were chosen as benchmarks to evaluate the practical and theoretical value of our prediction model.%   In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.
 
-\textit{Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible.
-The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.
+\textit{Prediction accuracy}: Prediction recall (fraction of correctly predicted disengaged learners) should be high to deliver the survey to as many disengaged learners as possible.
+The false positive rate (fraction of active learners who are incorrectly predicted as disengaged) should be low to avoid sending surveys to learners who do not experience challenges in the course.
 
-\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short.
-A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.
+\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as disengaged should be short.
+A shorter prediction lag improves the likelihood of successfully intervening to help learners resume the course.
 
-\textit{Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course.
-We thus had to train the prediction model using data from previous courses.
-The performance of the prediction model should be adequate in courses other than those on which it was trained.
+\textit{Transferability}: Whether or not a learner disengages from the course cannot be determined before the end of the course.
+We thus trained the prediction model using data from previous courses.
+The performance of the prediction model should be good in courses other than those on which it was trained.
 
-The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs.
+The training set was composed of learners' interaction data (predictors) and disengagement states (outcome) from 20 MOOCs.
 The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses.
@@
 A second test set of similar size was composed from 20 other MOOCs ($test_2$).
-Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).
-The model was fit by logistic regression.
-Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
+Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see omitted for blind review%\cite{halawa2014dropout}
+, for additional information on the prediction model).
+
+To control the length of the prediction lag, the model was only provided with learners' activity data until they were absent for 14 consecutive days.
+In the training phase, if the learner disengaged but the model did not detect it within 14 days of their last activity, the case was counted as a false negative.
+This construction results in a model with features that all fall within the defined prediction lag period.
+The model was fit by logistic regression with two-fold learner-level and course-level cross-validation.
+Table \ref{tab:predAuc} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
 
 \begin{table}[h!]
-\caption{Performance of the dropout prediction model}
-\label{tab:drop_pred_rocs}
+\caption{Performance of the attrition prediction model}
+\label{tab:predAuc}
 \small
@@
 
-Out of the 20,048 enrolled learners in the course, we only considered those 10,510 who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out.
-The survey was sent out to 6,053 leaners (excluding 21 learners with invalid email addresses or duplicate accounts).
+Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out.
+The survey was sent out to 6,053 learners (excluding 21 learners with invalid email addresses or duplicate accounts).
 

### 2014-10-30 paper_v1.Rnw
pred plot correction for final sub

@@
 res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
+
+# This is what we submitted
 ggplot(subset(res, coef2!="Intercept"), 
@@
   labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
+
+# This is what it should be
+# ggplot(subset(res, coef2!="Intercept"), 
+#   aes(coef2, 100*(exp(est)-1), ymin=100*(exp(est-1.96*se)-1), ymax=100*(exp(est+1.96*se)-1), color=factor(sign))) + 
+#   geom_hline(yintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
+#   theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
+#   scale_color_manual(values=c('red','grey30','green3')) +
+#   labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
+
 @

### 2014-11-01 paper_v2.Rnw
small change to submitted draft. added new draft with updated fig

--- 2014-10-30 paper_v1.Rnw
@@
 
-A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% points less likely than men to persist with lectures and assessments.
+A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% less likely than men to persist with lectures and assessments.
 Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men.
-Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America.
-Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in North America.
-While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from North America.
+Learners in Africa, Asia, and South America were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America.
+In contrast, learners in Europe exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America.
+While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America.
 
@@
 The baseline group were male learners in North America of average age who had not earned a bachelor\'s degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it.
-An estimate of -20 indicates a 20\\% points lower probability of achieving a milestone.'>>=
+An estimate of -20 indicates a 20\\% lower probability of achieving a milestone.'>>=
 library(ggplot2)
@@
 res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
+
+# This is what we submitted
+# ggplot(subset(res, coef2!="Intercept"), 
+#        aes(coef2, 100*est, ymin=100*(est-1.96*se), ymax=100*(est+1.96*se), color=factor(sign))) + 
+#   geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
+#   theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
+#   scale_color_manual(values=c('red','grey30','green3')) +
+#   labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
+
+# This is what it should be
 ggplot(subset(res, coef2!="Intercept"), 
-       aes(coef2, 100*est, ymin=100*(est-1.96*se), ymax=100*(est+1.96*se), color=factor(sign))) + 
-  geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
+  aes(coef2, 100*(exp(est)-1), ymin=100*(exp(est-1.96*se)-1), ymax=100*(exp(est+1.96*se)-1), color=factor(sign))) + 
+  geom_hline(yintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
   theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
   scale_color_manual(values=c('red','grey30','green3')) +
+  scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60)) +
   labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
+coefs = subset(res, coef2=="Intercept", select=c("est","y"))  # should report these in the text of reference
+coefs$exp_est = exp(coefs$est) # estimates seem high
 @

### 2014-11-01 paper_v2.tex
small change to submitted draft. added new draft with updated fig

--- 2014-10-30 paper_v1.tex
@@
 The baseline group were male learners in North America of average age who had not earned a bachelor's degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it.
-An estimate of -20 indicates a 20\% points lower probability of achieving a milestone.\label{fig:s1coefs}}
+An estimate of -20 indicates a 20\% lower probability of achieving a milestone.\label{fig:s1coefs}}
 \end{figure*}

### 2015-01-12 paper_v2.Rnw
Updating paper after first review

@@
 %% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-\toappear{%$^\ast$ All authors contributed equally. \\
+\toappear{
 Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
@@
 \hypersetup{
-pdftitle={SIGCHI Conference Proceedings Format},
-pdfauthor={LaTeX},
+pdftitle={Attrition in Massive Open Online Courses},
+pdfauthor={Rene F. Kizilcec, Sherif Halawa},
 pdfkeywords={SIGCHI, proceedings, archival format},
@@
 
-\title{Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors}
+\title{Attrition in Massive Open Online Courses}
 
@@
 \author{
-  \alignauthor Omitted for blind review\\
-  \affaddr{Institution}\\
-  \affaddr{Address}\\
-  \email{Email}\\
-%   \alignauthor Ren\'{e} F. Kizilcec$^\ast$\\
-%     \affaddr{Department of Communication}\\
-%     \affaddr{Stanford University}\\
-%     \email{kizilcec@stanford.edu}\\
-%   \alignauthor Sherif Halawa$^\ast$\\
-%     \affaddr{Department of Electrical Engineering}\\
-%     \affaddr{Stanford University}\\
-%     \email{halawa@stanford.edu}\\
+  \alignauthor Ren\'{e} F. Kizilcec\\
+    \affaddr{Department of Communication}\\
+    \affaddr{Stanford University}\\
+    \email{kizilcec@stanford.edu}\\
+  \alignauthor Sherif Halawa\\
+    \affaddr{Department of Electrical Engineering}\\
+    \affaddr{Stanford University}\\
+    \email{halawa@stanford.edu}\\
 }
@@
 \begin{abstract}
-Research on attrition has a long tradition in the field of education.
-Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
-A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1).
-Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence.
-A follow-up study with online learners who disengaged from the course (Study 2; $N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset.
-Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%).
-A low capacity for volitional control was identified as a large contributor to this issue.
+Attrition in online learning is generally higher than in traditional settings.
+High attrition in large-scale open learning environments prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
+A systematic analysis of individual differences in attrition in 17 diverse massive open online courses ($N>60,000$) revealed substantial achievement gaps: women were less likely to persist than men, and learners in Africa, Asia, and South America persisted less and scored lower than learners in Europe, North America, and Oceania.
+More persistent learners were, however, only marginally more satisfied with their achievement.
+Successful online learners who left the course early reported higher levels of goal striving, feelings of social belonging, and growth mindset than unsuccessful ones ($N=756$).
+The shortage of time was the primary obstacle for most learners, and many were hindered by insufficient volitional control.
 A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
@@
 
-\keywords{Online learning, persistence, attrition, dropout, disengagement, massive open online courses, MOOC, psychological factors}
+\keywords{Online learning, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}
 
@@
 \category{K.3.1.}{Computers and Education}{Computer Uses in Education}
-
-<<test1, echo=TRUE>>=
-@
 
@@
 Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}.
-Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes.
+Distance learning programs deliver education to remote parts of the world and to people who simply could not attend in-person classes.
 Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.
@@
 By design, MOOCs provide course materials to millions of people worldwide.
-This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}).
+This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer assessment).
 Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon.
@@
 
-Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}.
+Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc}.
 Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}.
@@
 Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized.
-Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}.
-Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition.
+For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}.
+This view of MOOC participation clearly adds complexity to interpretations of attrition.
 The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.
@@
 Their open response answers about what challenges they faced while taking the MOOC were iteratively coded.
-We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.
+We investigated the role of volitional control for learners who reported having not enough time, and examined goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.
 
 \begin{table*}[th]
-\label{tab:s1sum}
 \caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
@@
 \center
+\label{tab:s1sum}
 \begin{tabular}{llccccccccccc}
@@
 We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work.
-The focus of this brief review is on how ways of thinking about attrition have developed over the last decades.
+The focus of this brief review is on how ways of thinking about attrition have evolved over the last decades.
 
@@
 An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
-The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}.
-Hence, motivation alone is necessary but not sufficient for persistence.
-Students may fail to sustain efforts in the absence of strong self-regulatory skills.
+The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, the extent to which the student engages in goal-directed behavior in the face of distraction \cite{corno1993role}.
+Motivation alone is necessary but not sufficient for persistence, because students may fail to sustain efforts in the absence of strong self-regulatory skills.
 
 \begin{description}
-  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
+  \item[H1] Successful learners exhibit higher levels of goal striving than unsuccessful learners.
 \end{description}
@@
 Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively.
-This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2007question}.
+This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments (e.g.
+\cite{walton2007question}).
 
 \begin{description}
-  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from a MOOC?
-  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
+  \item[RQ1] How do prior characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from the course?
+  \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
 \end{description}
@@
 Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.
-Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (i.e., instilling a growth mindset) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}.
-The results highlight the powerful influence of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.
+Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (instilling a growth mindset \cite{dweck2006mindset}) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}.
+This highlights the critical role of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.
 
@@
   \item[RQ2a] What proportion of learners who disengaged were satisfied with their achievements in the course?
-  \item[RQ2b] How does learners' behavior in the MOOC vary by their level of satisfaction?
-  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
+  \item[RQ2b] How does learners' level of satisfaction depend on their behavior in the course?
+  \item[H3] Successful learners have a more growth mindset than unsuccessful learners.
 \end{description}
@@
 Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
-Persistence is thought to be a function of background characteristics (e.g., demographics), academic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction).
-The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
+Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.).
+The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
+% (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
 
@@
 Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning.
-And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
-An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
+And yet, these increasingly complex theoretical models were difficult to apply to real-world settings, and hence, the empirical evidence to support them remained sparse.
+An exploratory study of reasons for attrition of over 1,000 online students found eight relevant factors: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.
 An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.
@@
 
-A survey of over one hundred learners who dropped out of a MOOC showed that the majority indicated having too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}.
+A survey of over a hundred learners who dropped out of a MOOC showed that the majority indicated having too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}.
 A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}.
@@
 \begin{description}
-\item[RQ3a] What reasons do learners report for disengaging from a MOOC?
+\item[RQ3a] What reasons do learners report for disengaging from a course?
 \item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
@@
 The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs.
-We address the research questions developed in the previous section with a combination of self-report and behavioral data from seventeen diverse MOOCs.
+We address the research questions developed in the previous section with a combination of self-report and behavioral data from 17 diverse MOOCs.
 Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments.
@@
 
-Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}).
+Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (cf.
+Table \ref{tab:s1sum}).
 For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
@@
 On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course.
-On the feedback survey, learners reported their satisfaction with the course (6-item index, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see reasons in Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.} Geographic location was determined by IP address.
+On the feedback survey, learners reported their satisfaction with the course (6-item index, range from 1 to 7, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (Figure \ref{fig:s1reason} lists reasons).\footnote{Instead of a select-all-that-apply question, learners were encouraged to consider each reason in turn by using binary-choice items.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to mitigate acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
 
@@
 Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
-Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.}
+Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{Multiple imputation and subsequent pooled analyses were performed with the $R$ mice package.}
 % The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
@@
 
-We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features.
-Correlations between predictors were low enough to allow for simple interpretation of coefficients.
-Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{The required data for 3 of the 17 MOOCs could not be obtained in time for submission due to technical issues.} Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model.
-The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.
+We fit a logistic mixed-effects (hierarchical) model to predict multiple binary persistence and performance metrics (milestones) to estimate individual differences in learner characteristics, including demographic and geographic features.
+Correlations between predictors were low ($|r|<0.16$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
+The coefficient on education, for instance, reflects the association between having a bachelor's degree and achieving a milestone, partialing out all other predictors, including age.
+Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{Data for 3 of the 17 MOOCs was unavailable due to anonymization.}
 
-A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% less likely than men to persist with lectures and assessments.
-Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men.
-Learners in Africa, Asia, and South America were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America.
-In contrast, learners in Europe exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America.
+Figure \ref{fig:s1coefs} presents transformed regression coefficients with 95\% confidence intervals from each model.
+The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population.
+The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average aged male learners in North America without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
+In this sample of survey respondents, the baseline probability for each milestone was high: videos watched (27\% above 30\%, 16\% above 50\%, 8\% above 80\%) assessments attempted (78\% above 30\%, 55\% above 50\%, 33\% above 80\%), and grade (42\% above the 60$^\text{th}$, 31\% above the 80$^\text{th}$ percentile).
+
+A number of significant and substantial individual differences emerged.
+Women were 10 to 15\% less likely than men to persist with lectures and assessments.
+Women, who constituted 41\% of learners in the sample, were also less likely to score a grade above the 60$^\text{th}$ percentile than men.
+Learners in Africa ($n = 2,097$), Asia ($n = 8,199$), and South America ($n = 2,574$) were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America ($n = 18,353$).
+In contrast, learners in Europe ($n = 11,992$) exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America.
 While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America.
+Learners in Oceania ($n = 1,166$) performed similarly to those in North America and Europe.
 
 Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
-More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to perform highly on them.
+More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to score highly on them.
 Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
@@
 
-<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop.
-videos watched, prop.
-assessments completed, and grade percentile).
-The baseline group were male learners in North America of average age who had not earned a bachelor\'s degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it.
-An estimate of -20 indicates a 20\\% lower probability of achieving a milestone.'>>=
+<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Estimates from logistic mixed-effects model predicting persistence and performance milestones.
+An estimate of -10 indicates a 10\\% lower chance of achieving a milestone.'>>=
 library(ggplot2)
 res = read.csv("pred_coefs.csv")
+
 res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
-res$y = factor(res$y, labels=c("> 30% videos", "> 50% videos", "> 80% videos", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))
+res$y = factor(res$y, labels=c("watched >30% videos", "watched >50% videos", "watched >80% videos", ">60th percentile grade", "attempted >30% assessments", "attempted >50% assessments", "attempted >80% assessments", ">80th percentile grade"))
+
 res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
+res$coef2 = factor(res$coef2, labels=rev(c("Intercept", "Gender: Female", "Age (normalized)", "Education: Bachelor+", "Prior Experience", "Intent to complete", "Hours: 2-4", "Hours: 4+ ", "Prior MOOCs: 1-3", "Prior MOOCs: 4+ ", "Location: Europe", "Location: Oceania", "Location: Africa", "Location: Asia", "Location: South America")))
 
@@
   geom_hline(yintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
-  theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
-  scale_color_manual(values=c('red','grey30','green3')) +
-  scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60)) +
+  theme_minimal() + theme(legend.position="none",
+                          axis.title=element_text(vjust=0),
+                          panel.border=element_rect(fill="transparent",size=.5,color="grey50")) + 
+  scale_color_manual(values=c('black','grey50','black')) +
+  scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60,80)) +
   labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
 coefs = subset(res, coef2=="Intercept", select=c("est","y"))  # should report these in the text of reference
-coefs$exp_est = exp(coefs$est) # estimates seem high
+coefs$exp_est = round(100*exp(coefs$est)) # estimates seem high
+# >30% videos      27
+# >50% videos      16
+# >80% videos       8
+# >30% assessments      78
+# >50% assessments      55
+# >80% assessments      33
+# >60% pctile grade      42
+# >80% pctile grade      31
 @
@@
 Attrition can be a sign of success if a learner has achieved their personal goal.
-We therefore investigated to what extent learners were satisfied with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$).
-To account for differences across courses, we compared satisfaction scores using mixed-effects models.
-
-Learners who reported that they stopped participating were significantly 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$).
+We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$).
+Satisfaction ratings were compared with a mixed-effects regression to model differences across courses and generalize to a larger population of MOOCs.
+Learners who reported that they stopped participating were 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$).
 Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied.
-While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=12, p<.001$), it was a relatively weak effect.
-Those who watched all videos were only around 3\% more satisfied than those who watched none, based on a linear fit.
+While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=14, p<.001$), the effect was relatively weak.
+Those who watched over 75\% of videos were only 5\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=4.97, M_{>\text{75\%}}=5.22$).
 
@@
 
+sat$prop_vid[sat$prop_vid>1] = 1
 sat$prop_vid4 = cut(sat$prop_vid, breaks=c(-Inf, .25, .5, .75, Inf))
+table(sat$prop_vid4)
 
-require(lmerTest)
 summary(lmer(satisfaction~ did_dropout + (1|course_id), sat))
@@
 
-summary(lmer(satisfaction~0+prop_vid4 + (1|course_id), sat))
+summary(lmer(satisfaction~prop_vid4 + (1|course_id), sat))
 summary(lmer(satisfaction~prop_vid + (1|course_id), sat))
-(5.005+0.1679)/5.005
+(4.969+.2466)/4.969
 @
@@
 
-Research question $RQ3a$ about learners' reasons for disengaging from a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.\footnote{This researcher-defined list may not have been collectively exhaustive, which is addressed in Study 2.} The distribution of proportions across 17 MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots.
-Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time.
-Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents.
-Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.
+% Say how many actually reported N=?
+Learners' reasons for disengaging from a MOOC were examined using survey responses ($RQ3a$).
+Across 17 MOOCs, learners selected reasons for disengaging from a pre-defined list.
+This list may not be collectively exhaustive, which is addressed in Study 2.
+The distribution of chosen reasons is illustrated in Figure \ref{fig:s1reason}.
+Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating.
+Learners faced issues keeping up with deadlines and found that the course demanded too much time.
+Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents.
+Notably, just under 20\% of respondents in a typical course (median proportion) stopped participating because they learned all they intended to learn.
 
+% say how this was done (linear combination of correlation matrices; exluding any? if so, N=?)
 To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis.
 The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
-The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em  difficulty} (too advanced, exam too difficult); fourth, {\em format and content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
-Only the technical difficulties item failed to load onto any factor.
+The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em difficulty} (too advanced, exam too difficult); fourth, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em goals \& expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
+The technical difficulties item failed to load onto any factor.
 
+% NEED TO SAY THAT IT'S LITTLE DATA HERE (n=?)
 We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course.
 A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations.
-Taking the course with friends or colleagues was correlated with disengaging because the course is not challenging enough, the material is too advanced of the learner's background, and the credential is not valuable enough to the learner ($r=.24, .20, .16$, respectively).
+Taking the course with friends or colleagues was correlated with disengaging because the course was not challenging enough, the material was too advanced for the learner's background, and the credential was not valuable enough to the learner ($r=.24, .20, .16$, respectively).
 Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$).
 Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
+
+% other covariates: e.g. do we see geolocation difference for language and tech?
+% also gender differences?
 
@@
 \subsection{Discussion}
-
+% need to adjust once the rest is finalized above
 A number of striking findings emerged from the preceeding analyses.
@@
 General time issues were overall the most pronounced.
+Attrition in community colleges has been interpreted as a sign of success in prior work.
+This view resonates in the MOOC setting, where around 20\% of learners who disengaged reportedly had achieved their goal, and where progress in the course and learner satisfaction were only weakly related.
 
@@
 To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study.
-In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback.
-To this end, we developed a predictive model to identify disengaging learners and asked them to complete a survey.
-One goal of this study was to further investigate research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced.
-Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time.
-A variety of factors can influence the amount of time learners have at their disposal.
-Higher priority obligations, such as professional, academic, and personal committments, are presumably large contributors.
+This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced.
+We invited learners who were likely to drop out of a course to provide feedback.
+These learners were identified with a predictive model and invited to complete a survey.
+The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
+Moreover, most learners indicated in Study 1 that they disengaged due to a shortage of time.
+A variety of factors could influence the amount of time learners have at their disposal.
+Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors.
 Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}.
-For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV).
-We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:
+Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV).
+We constructed the following hypotheses about the role of volition in attrition due to a shortage of time:
 
@@
 
-Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
-To this end, we included a self-report measure for each construct in the survey that was sent to disengaging learners.
-
 \subsection{Methods}
 
-The particular MOOC under observation was an undergraduate level course on an advanced topic in computer science.
+The particular MOOC under observation was an undergraduate-level course on an advanced topic in computer science.
 It was offered in 2014 through Coursera.
@@
 A second test set of similar size was composed from 20 other MOOCs ($test_2$).
-Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see omitted for blind review%\cite{halawa2014dropout}
-, for additional information on the prediction model).
+Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (cf.
+\cite{halawa2014dropout}, for additional information on the prediction model).
 
@@
 The model was fit by logistic regression with two-fold learner-level and course-level cross-validation.
-Table \ref{tab:predAuc} provides information on the prediction performance, measured by the area under the ROC curve (AUC).
-
-\begin{table}[h!]
-\caption{Performance of the attrition prediction model}
-\label{tab:predAuc}
-\small
-\center
-\begin{tabular}{lccc}
-\toprule
- & $train$ & $test_1$ & $test_2$ \\
-\midrule
-AUC &  0.931 & 0.929 & 0.922  \\
-\bottomrule
-\end{tabular}
-\end{table}
-
+Prediction performance measured by the area under the ROC curve (AUC) very high (0.931 for $train$, 0.929 for $test_1$, 0.922 for $test_2$).
 Two of the predictors in the fitted model are highly influential.
-The first is the number of days since the learner has last been active ($x_{1}$).
+The first is the number of days since the learner has last been active.
 The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more.
-However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model.
-Equation 1 is an approximation of the model with only these two predictors.
+However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed prior to the absence; this is the second most influential feature in the model.
 The fact that almost identical results were obtained for $test_1$ and $test_2$ suggest good transferability of the model.
-A learner was flagged if their dropout probability exceeded a half:
-
-\begin{equation}
-Pr(\text{dropout}) = \frac{1}{1+e^{-0.431x_1 + 6.388x_2 + 1.626}} > 0.5
-\end{equation}
-
+A learner was flagged if their dropout probability exceeded 50\%.
 Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out.
@@
 They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of.
-This question was deliberately asked prior to any survey questions that could suggest particular reasons for disengagement.
+This question was asked prior to any survey questions that could suggest particular reasons for disengaging.
 
@@
 In the second iteration, the remaining responses were coded in the same way.
-In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see [URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
+In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see !!!!!!![URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
 
 % Learners also rated the extent to which their progress was hindered by a number of obstacles. 
-Learners also rated their sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
+Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adpated from \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
 
@@
 
-\subsubsection{Reasons for Dropout}
-
-The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course.
-While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
-exhibit lower volitional control).
-These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).
-
-Reasons for learner attrition were measured using iteratively coded open responses ($n=399$).
-Table \ref{tab:s2reas} provides each coded challenge and percentage of open responses that indicated it.
-A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).
-Hence, if a learner had not enough time, this tended to be the main challenge.
-
-\begin{table}[h!]
-\caption{Iteratively-Coded Reasons for Attrition}
-\label{tab:s2reas}
-\small
-\center
-\begin{tabular}{lr}
-\toprule
-What challenges have you experienced while taking this course?  & \% \\
-\midrule
-Not enough time for the course & 84 \\
-The course format or teaching style is not a good fit & 18 \\
-The level of the course is too advanced & 10 \\
-The course is not interesting or valuable enough & 7 \\
-Additional course materials needed to learn the topic & 4 \\
-Technical difficulties & 2 \\
-Language barrier & 1 \\
-\bottomrule
-\end{tabular}
-\end{table}
+\subsubsection{Psychological Factors}
 
@@
 
-To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time.
+Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
+Success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
+The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
+Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
+Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
+Social belonging was significantly higher for learners who were successful in terms of satisfaction, but those who were successful in terms of progress.
+The pattern was reversed for growth mindset, such that learners who were successful in terms of progress (but not satisfaction) exhibited a stronger growth mindset.
+This pattern could reflect a difference in the underlying process by which these psychological factors affect learner behavior and perception.
+The observed differences are expected to be larger in the full population of learners, compared to the small, homogenous sample in this study, consisting of learners unlikely to persist.
+
+\subsubsection{Reasons for Dropout}
+
+The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course.
+While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
+exhibit lower volitional control).
+These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).
+Reasons for learner attrition were measured using iteratively coded open responses.
+Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it.
+A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).
+Hence, if a learner had not enough time, this tended to be the primary challenge.
+
+To test hypotheses $H4ab$, we examined the 336 open responses that indicated having too little time.
 It revealed that no particular commitment was specified in about half of them (53\%).
@@
 However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
-These findings lend strong support to hypotheses $H4a$ and $H4b$: that is, while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
+These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
 
-\subsubsection{Psychological Factors}
-
-Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
-Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
-The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
-Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
-Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
-Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress.
-The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction.
-Hence, the results fully support hypothesis $H1$, and partially support $H2$ and $H3$.
+\begin{table}[h!]
+\caption{Iteratively-Coded Reasons for Attrition ($n=399$)}
+\label{tab:s2reas}
+\small
+\center
+\begin{tabular}{lr}
+\toprule
+What challenges have you experienced while taking this course?  & \% \\
+\midrule
+Not enough time for the course & 84 \\
+The course format or teaching style is not a good fit & 18 \\
+The level of the course is too advanced & 10 \\
+The course is not interesting or valuable enough & 7 \\
+Additional course materials needed to learn the topic & 4 \\
+Technical difficulties & 2 \\
+Language barrier & 1 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-Study 2 focused on reasons for attrition and psychological factors of learners who were likely to drop out of the course.
-Through iterative coding, we developed a set of challenges that were faced by a subset of learners in a particular MOOC (cf.
-Table \ref{tab:s2reas}).
-The large majority of learners who indicate that they had not enough time was found to be separable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$).
-MOOCs are an environment that requires learners to be highly self-regulated to be successful.
-The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}.
-Very few learners explicitly indicated issues with volitional control in their open responses, potentially because it is not socially desirable to admit.
-The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$).
-This could be a useful proxy for volitional issues in future research.
-
-We investigated the extent to which personal success in MOOCs is associated with three established psychological factors: goal striving, social belonging, and growth mindset.
-Prior evidence found the absence of each factor to cause lower persistence and performance in more traditional achievement-oriented environments.
-The present findings highlight the critical role of these factors in MOOCs based on observational evidence.
-Learners with higher levels of goal striving, social belonging, and growth mindset also reported higher levels of personal success ($H1,H2,H3$).
-The subpopulation of learners in this study was a small and intentionally skewed subset.
-The observed differences are expected to be larger in the full population of learners, compared to this homogenous sample of learners who are unlikely to persist.
-Goal striving was most strongly associated with both measures of personal success.
-Success in terms of relative progress was associated with a stronger growth mindset, but not significantly with social belonging.
-In contrast, success in terms of satisfaction with progress was associated with stronger social belonging, but not significantly with growth mindset.
-Future work should examine if this pattern reflects a difference in the underlying process by which these psychological factors affect learner behavior and perception.
+In this follow-up study, we examined the role of psychological factors in attrition and further investigated reasons for attrition.
+Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset.
+Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance.
+The current observational findings highlight the critical role of these factors in an online learning environment.
+Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$).
+An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging.
+These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues.
+Pschological factors could thereby, at least partially, account for individual differences in persistence and performance (Study 1).
 
@@
 Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).
-Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their identity or social belonging.
+Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
+Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their social belonging.
+
+Through iterative coding, we arrived at a list of challenges faced by a subset of learners in a particular MOOC (Table \ref{tab:s2reas}).
+The list maps onto the factors identified in Study 1 (general time issues, late start, difficulty, format \& content, expectations, and technical difficulties), with the additional challenge of a language barrier, which was not included as a response option in Study 1.
+The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$).
+A learner needs to be highly self-regulated to be successful in a MOOC.
+Very few learners explicitly indicated issues with volitional control in their open responses.
+Learners may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait.
+Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$).
+This could be a useful proxy for identifying issues with volitional control in future work.
+Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.
 
@@
 
-The present studies extend the empirical base of research on attrition in online learning environments.
-A large-scale study of over 60,000 online learners revealed strong individual differences in persistence and performance by demographic and geographic features.
-Learners who reported that they dropped out were less satisfied with their course achievements.
-Time was a critical obstacle for many learners, and a low capacity for volitional control was found to contribute to this issue, though learners seemingly avoid admitting this.
-Successful online learners, according to their own judgment, were found to exhibit higher levels of goal-striving, social belonging, and growth mindset than subjectively unsuccessful learners.
-We also began to develop a survey item for measuring reasons for disengaging from a course (cf.
-Table \ref{tab:s2reas}), which should be tested and extended, if needed, based on more empirical evidence from additional MOOCs.
+This study of over 60,000 online learners, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments.
+The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners with different demographic features and from different geographic locations; \emph{second}, the relation between personal success in online learning (by learners' own judgment) and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which was frequently related to learners' low capacity for volitional control.
+Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf.
+Table \ref{tab:s2reas}), which should be tested and potentially extended based on more empirical evidence from additional courses.
 
-This work yields a number of theoretical and practical implications.
 In addressing research questions and testing hypotheses which emerged from a review of the literature, we sought to advance theory on attrition and provide empirical evidence.
@@
 Moreover, volitional control was identified as a major contributor in reported time-related challenges with taking a MOOC.
-Tinto's \cite{tinto1975dropout} classic model emphasizes the role of academic and social integration, and also builds on various learner characteristics as contributors to learners' decision to persist in a course of study.
-Our findings provide strong support for the critical role of individual characteristics, such as gender, age, intentions, experiences, and geographic location; and, leveraging the more recent notion of social belonging \cite{walton2007question}, we found evidence in support of Tinto's social integration component.
-Parallel to positive attributions of attrition in community college settings, we found progress in the MOOC and learner satisfaction to only be weakly related: learners who watched very few lectures were almost as satisfied as those who watched most lectures.
-Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
+Tinto's \cite{tinto1975dropout} classic model of attrition emphasizes the role of academic and social integration, and incorporates individual learner characteristics as contributors to learners' decision to persist in a course of study.
+Our findings provide strong support for the critical role of individual characteristics, such as gender, age, intentions, experiences, and geographic location.
+Furthermore, we found evidence for the importance of social integration in terms of perceived social belonging \cite{walton2007question}.
+In contrast to Tinto's concept of social and academic integration, which characterizes a general state in an institution, social belonging is a subjective experience based on perceptions and interpretations of environmental cues.
+Yet, this subjective experience is typically shared by individuals with a common identity; for example,  female learners in male-dominated fields or black minority learners in college.
+Models of attrition should be updated to acknowledge the potential impact of psychological threats, which can cause achievement gaps that would otherwise be overlooked or misinterpreted.
 
@@
 While the current work does not provide explanations for the presence of substantial achievement gaps, we can speculate that similar psychological processes as in traditional educational environments are involved.
-Brief and targeted online interventions (e.g., mindset, social belonging, self-affirmation, goal striving) can potentially narrow existing achievement gaps in online learning.
-We hope to see more research on developing accurate prediction models that identify struggling learners and suitable interventions to best support them.
+Brief and targeted online interventions (mindset, social belonging, goal striving, volitional control, etc.) can potentially narrow existing achievement gaps in online learning.
+More research on prediction models that identify struggling learners and targeted interventions to support them can make a big difference in online learning.
 
 \section{Acknowledgments}
-% We thank Elise Ogle and Ruth Bram for helping with the development of the initial codebook. 
-% Thank Emily Schneider for her helpful feedback and Andreas P. for his feedback and data support.
-Omitted for blind review.
+We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for helping with the development of the codebook.
+The authors contributed equally to this work.
 

### 2015-01-12 paper_v2.tex
Updating paper after first review

@@
 
-\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop.
-videos watched, prop.
-assessments completed, and grade percentile).
-The baseline group were male learners in North America of average age who had not earned a bachelor's degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it.
-An estimate of -20 indicates a 20\% lower probability of achieving a milestone.\label{fig:s1coefs}}
+\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Estimates from logistic mixed-effects model predicting persistence and performance milestones]{Estimates from logistic mixed-effects model predicting persistence and performance milestones.
+An estimate of -10 indicates a 10\% lower chance of achieving a milestone.\label{fig:s1coefs}}
 \end{figure*}

### 2015-01-12 course_data.R

@@ -4,8 +4,8 @@ library(psych)
 library(foreach)
 library(ggplot2)
 
-### Analysis of correlations between OLEI and dropout reasons
-setwd("~/Dropbox/r-code/dropout_reasons/data_masters/")
+### Analysis of correlations between dropout reasons
+setwd("~/Dropbox/r-code/dropout_reasons/data_all_pre3/")
 files = dir()
 Mcor = list()
 i = 1
@@ -14,9 +14,10 @@ N = 0
 for(f in files){
   print(f)
   dat = read.csv(f)
-  n = sum(!is.na(dat[,7]))
+  n = sum(!is.na(dat[,12]))
+  print(n)
   if(n > 30) {
-    m = data.frame(cor(dat[,7:20], use="pair"))
+    m = data.frame(cor(dat[,12:25], use="pair"))
     m[is.na(m)] = 0
     Mcor[[i]] = m
     N = N + n
@@ -31,6 +32,59 @@ fa.diagram(fa(CombMcor, 5, n.obs=N))
 fa(CombMcor, 5, n.obs=N)
 plot(princomp(CombMcor, cor=T))
 summary(princomp(CombMcor, cor=T))
+
+
+### Analysis of correlations between demographics/geo and dropout reasons
+library(foreach)
+setwd("~/Dropbox/r-code/dropout_reasons/data_all_pre3/")
+files = dir()
+
+dat = foreach(f=files, .combine=rbind) %do% {
+  dat = read.csv(f)
+  temp = dat[,c(4,6,12:25,30,33)]
+  temp$course = f
+  subset(temp, did_dropout==1)
+}  
+
+# Geolocation
+dat$ip_address = as.character(dat$ip_address)
+# write.table(dat$ip_address, file="../../ip_lookup/ips.csv", col.names=F, row.names=F, sep=",", quote=F)
+ips = read.csv("../../ip_lookup/ips_geo.csv")
+ips$continent = as.character(ips$continent)
+ips[is.na(ips$continent),]$continent = "NAmerica"
+ips[ips$continent=="SA",]$continent = "SAmerica"
+ips[ips$continent%in%c("--","unk"),]$continent = NA
+plot(ips$longitude,ips$latitude, col=as.numeric(factor(ips$continent)))
+dat$country = ips$country_name
+dat$continent = ips$continent
+dat$continent = factor(dat$continent, levels=c("NAmerica","EU","OC","AF","AS","SAmerica"))
+rm(ips)
+
[…]

### 2015-01-24 paper_v2.Rnw
updated version

@@
 
-\title{Attrition in Massive Open Online Courses}
+\title{Attrition and Achievement Gaps in Online Learning}
 
@@
 \begin{abstract}
-Attrition in online learning is generally higher than in traditional settings.
-High attrition in large-scale open learning environments prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
-A systematic analysis of individual differences in attrition in 17 diverse massive open online courses ($N>60,000$) revealed substantial achievement gaps: women were less likely to persist than men, and learners in Africa, Asia, and South America persisted less and scored lower than learners in Europe, North America, and Oceania.
+Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments.
+This has prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
+A systematic analysis of individual differences in attrition in 20 diverse massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap.
+Women and learners in Africa, Asia, and Latin America scored lower grades and persisted less than men and learners in Europe, Oceania, and Northern America.
 More persistent learners were, however, only marginally more satisfied with their achievement.
@@
 
-\keywords{Online learning, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}
+\keywords{Massive open online course, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}
 
@@
 
-This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall.
-A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals.
-We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes.
-Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions.
-In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback.
-Their open response answers about what challenges they faced while taking the MOOC were iteratively coded.
-We investigated the role of volitional control for learners who reported having not enough time, and examined goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.
+This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total.
+Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggeling to achieve their learning goals.
+These learners could benefit from changes to the course or targeted interventions.
+We briefly review the large literature on attrition in educational environments, with a focus on key developments in understanding the causes of attrition.
+Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance.
+For Study 2, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback.
+We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time.
+Learners' open responses about challenges they faced while taking the course were iteratively coded to develop an empirically-grounded list of reasons for attrition.
 
 \begin{table*}[th]
-\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
+\caption{Summary statistics for 20 massive open online courses}
 \small
@@
 \midrule
-C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
-C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
-C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
-C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
-C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
-C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
-C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
-C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
-C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
-C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
-C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
-C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
-C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
-C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
-C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
-C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
-C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
+C1 & Business & 61,233 & 8,618 & 1,796 & 1,282 & 47\% & 36 (28,43) & 94\% & 11\% & 11\% & 9\% & 6\%\\
+C2 & Business & 45,202 & 5,951 & 789 & 631 & 43\% & 36 (27,42) & 90\% & 10\% & 11\% & 8\% & 6\%\\
+C3 & Computer Science & 28,086 & 5,918 & 2,815 & 2,295 & 36\% & 33 (23,40) & 63\% & 3\% & 15\% & 11\% & 8\%\\
+C4 & Computer Science & 56,314 & 2,686 & 749 & 352 & 14\% & 33 (25,39) & 79\% & 21\% & 13\% & 10\% & 8\%\\
+C5 & Computer Science & 79,273 & 7,319 & 1,740 & 1,236 & 16\% & 31 (24,36) & 81\% & 11\% & 15\% & 11\% & 8\%\\
+C6 & Education & 40,214 & 24,192 & 5,174 & 4,969 & 56\% & 28 (15,40) & n.a. & 5\% & 17\% & 13\% & 8\%\\
+C7 & Education & 5,325 & 2,429 & 624 & 615 & 80\% & 43 (35,50) & 99\% & 2\% & 24\% & 17\% & 12\%\\
+C8 & Engineering & 11,958 & 2,341 & 262 & 196 & 16\% & 35 (26,43) & 84\% & 16\% & 14\% & 9\% & 6\%\\
+C9 & Engineering & 5,701 & 5,652 & 396 & 347 & 16\% & 33 (24,38) & 73\% & 16\% & 18\% & 11\% & 7\%\\
+C10 & Engineering & 7,139 & 1,250 & 197 & 168 & 17\% & 31 (22,36) & 72\% & 7\% & 11\% & 8\% & 6\%\\
+C11 & Humanities & 7,656 & 2,512 & 142 & 131 & 89\% & 33 (24,38) & 82\% & 9\% & 14\% & 9\% & 6\%\\
+C12 & Humanities & 4,949 & 2,921 & 292 & 270 & 91\% & 33 (25,38) & 83\% & 17\% & 15\% & 10\% & 7\%\\
+C13 & Mathematics & 49,509 & 3,031 & 836 & 510 & 35\% & 37 (26,47) & 76\% & 32\% & 12\% & 9\% & 6\%\\
+C14 & Mathematics & 13,310 & 2,861 & 346 & 298 & 10\% & 31 (25,34) & 93\% & 15\% & 15\% & 10\% & 6\%\\
+C15 & Natural Sciences & 9,918 & 3,957 & 1,351 & 1,223 & 51\% & 37 (26,47) & 78\% & 4\% & 19\% & 15\% & 11\%\\
+C16 & Natural Sciences & 543 & 163 & 79 & 50 & 47\% & 34 (26,39) & 98\% & 18\% & 27\% & 20\% & 12\%\\
+C17 & Physics & 6,452 & 1,069 & 269 & 269 & 14\% & 36 (25,46) & 82\% & 11\% & 15\% & 10\% & 6\%\\
+C18 & Statistics & 11,733 & 11,465 & 1,837 & 1,569 & 22\% & 36 (28,41) & 96\% & 9\% & 34\% & 25\% & 15\%\\
+C19 & Statistics & 13,161 & 3,495 & 819 & 627 & n.a. & n.a. & n.a. & 6\% & 13\% & 9\% & 7\%\\
+C20 & Writing & 35,374 & 12,514 & 2,877 & 2,876 & n.a. & n.a. & n.a. & 8\% & 23\% & 14\% & 8\%\\
 \bottomrule
@@
 
-Research and theorizing on attrition in education has a rich history.
-This review is intended to serve as a foundation to build on with the current research.
-We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work.
-The focus of this brief review is on how ways of thinking about attrition have evolved over the last decades.
+Research and theory on attrition has a rich history in education.
+This review focuses on how conceptions of attrition have evolved over the last decades.
+As this review is intended to serve as a foundation to build on with the current research, we concurrently develop hypotheses and research questions to investigate.
 
@@
 \item[RQ3a] What reasons do learners report for disengaging from a course?
-\item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
+\item[RQ3b] How do reasons for disengaging vary by learner characteristics?
 \end{description}
@@
 
-The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs.
-We address the research questions developed in the previous section with a combination of self-report and behavioral data from 17 diverse MOOCs.
-Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments.
-This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf.
-Study 2).
+The present study extends the existing literature on attrition with a large-scale quantitative treatment of attrition in MOOCs.
+We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners.
+This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggeling learners.
 
@@
 
-Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (cf.
-Table \ref{tab:s1sum}).
-For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
-% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course.
-On the feedback survey, learners reported their satisfaction with the course (6-item index, range from 1 to 7, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (Figure \ref{fig:s1reason} lists reasons).\footnote{Instead of a select-all-that-apply question, learners were encouraged to consider each reason in turn by using binary-choice items.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to mitigate acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
+Twenty MOOCs on topics in a variety of disciplines were selected for investigation.
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
+For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course.
+On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf.
+Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
+Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
 
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
-Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
-A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 17 MOOCs in our dataset.
-Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
-Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{Multiple imputation and subsequent pooled analyses were performed with the $R$ mice package.}
-% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
-% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
-% Information about age, gender, and education level were only available for learners who completed the initial course survey.
-% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior
+To analyze individual differences, we fit a logistic mixed-effects (hierarchical) model predicting binary persistence and performance milestones using learner characteristics, including demographic and geographic features.
+The analysis included 67,333 learners who completed the initial course survey, but excluded 381 learners who left all questions unanswered and learners in courses C6, C7, C19, C20 due to missing survey items.
+Any remaining missing survey responses were multiply imputed using other survey responses.
+We report pooled estimates that account for added uncertainty.\footnote{Multiple imputation ($m=10$) using the $R$ mice package.} Correlations between predictors were low ($|r|<0.15$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
+Courses were modeled as a random effect (instead of including dummy variables) to generalize to a larger population of similar MOOCs.
 
-\subsection{Results}
+\subsection{Results and Discussion}
 
-We address the research questions developed in the previous section in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across subsets of learners ($RQ2ab$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3ab$).
+We address the posed research questions in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across learner subgroups ($RQ2$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3$).
 
-\subsubsection{Individual Differences in Attrition}
+\subsubsection{Individual Differences in Attrition and Performance}
 
-We fit a logistic mixed-effects (hierarchical) model to predict multiple binary persistence and performance metrics (milestones) to estimate individual differences in learner characteristics, including demographic and geographic features.
-Correlations between predictors were low ($|r|<0.16$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
-The coefficient on education, for instance, reflects the association between having a bachelor's degree and achieving a milestone, partialing out all other predictors, including age.
-Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{Data for 3 of the 17 MOOCs was unavailable due to anonymization.}
-
-Figure \ref{fig:s1coefs} presents transformed regression coefficients with 95\% confidence intervals from each model.
+Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone.
 The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population.
-The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average aged male learners in North America without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
-In this sample of survey respondents, the baseline probability for each milestone was high: videos watched (27\% above 30\%, 16\% above 50\%, 8\% above 80\%) assessments attempted (78\% above 30\%, 55\% above 50\%, 33\% above 80\%), and grade (42\% above the 60$^\text{th}$, 31\% above the 80$^\text{th}$ percentile).
+The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
+The baseline probability for each milestone was relatively high: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).
 
 A number of significant and substantial individual differences emerged.
-Women were 10 to 15\% less likely than men to persist with lectures and assessments.
-Women, who constituted 41\% of learners in the sample, were also less likely to score a grade above the 60$^\text{th}$ percentile than men.
-Learners in Africa ($n = 2,097$), Asia ($n = 8,199$), and South America ($n = 2,574$) were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America ($n = 18,353$).
-In contrast, learners in Europe ($n = 11,992$) exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America.
-While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America.
-Learners in Oceania ($n = 1,166$) performed similarly to those in North America and Europe.
+Women were 12 to 20\% less likely than men to persist with lectures and assessments.
+Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.
+Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$).
+In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America.
+While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America.
+Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe.
+Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners.
+More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile.
+Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
+Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.
 
-Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
-More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to score highly on them.
-Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
-Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.
+The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning.
+Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America.
+There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates.
+The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model.
+The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings.
+The present work cannot move beyond presenting evidence for the existence of these gaps and discussing potential causes.
+Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps; but rather, gender and location are associated with countless individual characteristics, some of which impact academic achievement.
+The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}.
+More work is needed to identify potential causal mechanisms underlying these gaps.
 
-<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Estimates from logistic mixed-effects model predicting persistence and performance milestones.
+<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Individual differnces in persistence and course grades showing substantial achievement gaps.
 An estimate of -10 indicates a 10\\% lower chance of achieving a milestone.'>>=
 library(ggplot2)
-res = read.csv("pred_coefs.csv")
+res = read.csv("pred_coefs2014-1-15.csv")
 
 res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
-res$y = factor(res$y, labels=c("watched >30% videos", "watched >50% videos", "watched >80% videos", ">60th percentile grade", "attempted >30% assessments", "attempted >50% assessments", "attempted >80% assessments", ">80th percentile grade"))
+res$y = factor(res$y, labels=c("> 30% lectures", "> 50% lectures", "> 80% lectures", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))
 
-res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
-res$coef2 = factor(res$coef2, labels=rev(c("Intercept", "Gender: Female", "Age (normalized)", "Education: Bachelor+", "Prior Experience", "Intent to complete", "Hours: 2-4", "Hours: 4+ ", "Prior MOOCs: 1-3", "Prior MOOCs: 4+ ", "Location: Europe", "Location: Oceania", "Location: Africa", "Location: Asia", "Location: South America")))
+res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:Latin America")))
+res$coef2 = factor(res$coef2, labels=rev(c("Intercept", "Gender: Female", "Age (normalized)", "Education: Bachelor+", "Prior Experience", "Intent to complete", "Hours: 2-4", "Hours: 4+ ", "Prior MOOCs: 1-3", "Prior MOOCs: 4+ ", "Location: Europe", "Location: Oceania", "Location: Africa", "Location: Asia", "Location: Latin America")))
 
@@
   scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60,80)) +
-  labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
+  labs(y="Predicted change in probability of achieving each milestone (95% CIs)", x="")
 coefs = subset(res, coef2=="Intercept", select=c("est","y"))  # should report these in the text of reference
 coefs$exp_est = round(100*exp(coefs$est)) # estimates seem high
-# >30% videos      27
-# >50% videos      16
-# >80% videos       8
-# >30% assessments      78
-# >50% assessments      55
-# >80% assessments      33
-# >60% pctile grade      42
-# >80% pctile grade      31
+# 1   -1.63          > 30% lectures      20
+# 16  -2.20          > 50% lectures      11
+# 31  -3.13          > 80% lectures       4
+# 46  -1.00       > 30% assessments      37
+# 61  -1.38       > 50% assessments      25
+# 76  -1.73       > 80% assessments      18
+# 91  -1.21 > 60th percentile grade      30
+# 106 -1.51 > 80th percentile grade      22
 @
 
-\subsubsection{Learner Satisfaction}
+\subsubsection{Satisfaction}
 % RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
@@
 
-Attrition can be a sign of success if a learner has achieved their personal goal.
-We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$).
-Satisfaction ratings were compared with a mixed-effects regression to model differences across courses and generalize to a larger population of MOOCs.
-Learners who reported that they stopped participating were 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$).
+Attrition can be a sign of success if a learner has achieved their personal learning objectives.
+We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who persisted to different extents ($RQ2b$).
+Satisfaction ratings were compared using mixed-effects regression ($N=23,390$).
+Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$).
 Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied.
-While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=14, p<.001$), the effect was relatively weak.
-Those who watched over 75\% of videos were only 5\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=4.97, M_{>\text{75\%}}=5.22$).
-
-<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=5, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
-#b = a[!is.na(a$did_dropout),]
-#ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
-@
-
-% We also examined differences in satisfaction between learners grouped by certain behaviors.
-First, we classified learners into 4 groups based on the fraction of course videos they viewed.
-The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}.
-The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group.
-Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.
-
-<<echo=FALSE,eval=FALSE>>=
-library(dplyr)
-sat = read.csv("satisfaction.csv")
-sat$num_videos_watched[is.na(sat$num_videos_watched)] = 0
-sat = sat %>% group_by(course_id) %>% 
-  mutate(prop_vid = num_videos_watched / quantile(num_videos_watched[num_videos_watched!=0], .90))
-
-sat$prop_vid[sat$prop_vid>1] = 1
-sat$prop_vid4 = cut(sat$prop_vid, breaks=c(-Inf, .25, .5, .75, Inf))
-table(sat$prop_vid4)
-
-summary(lmer(satisfaction~ did_dropout + (1|course_id), sat))
-5.170/(5.170-0.4456)
-
-summary(lmer(satisfaction~prop_vid4 + (1|course_id), sat))
-summary(lmer(satisfaction~prop_vid + (1|course_id), sat))
-(4.969+.2466)/4.969
-@
+The proportion of videos significantly predicted higher satisfaction ($t_{101}=20, p<.001$), such that those who watched over 75\% of videos were 8\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=3.80, M_{>\text{75\%}}=4.09$).
+As levels of self-reported satisfaction were generally high, it remains unclear if the estimated differences in satisfaction are consequential.
+Learners who disengaged early were only somewhat less satisfied than those who persisted in the course, which suggests that some learners arrived with modest course goals and achieved them.
+The following analysis of reasons for attrition can shed more light on this hypothesis.
 
@@
 
-% Say how many actually reported N=?
-Learners' reasons for disengaging from a MOOC were examined using survey responses ($RQ3a$).
-Across 17 MOOCs, learners selected reasons for disengaging from a pre-defined list.
-This list may not be collectively exhaustive, which is addressed in Study 2.
-The distribution of chosen reasons is illustrated in Figure \ref{fig:s1reason}.
+A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a pre-defined list ($RQ3a$).
+This list may not be collectively exhaustive, which is addressed in Study 2, but responses were collected from 1,698 learners in 20 different courses.
+Figure \ref{fig:s1reason} illustrates how frequently each reason was selected across courses.
 Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating.
-Learners faced issues keeping up with deadlines and found that the course demanded too much time.
+In a typical course (median proportion), 66\% faced issues keeping up with deadlines and for 46\% the course demanded too much time.
+According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gains.
+The shortage of time could be due to an abundance of higher-priority tasks, as enrolling in a MOOC neither guarantees that a learner can nor intends to allocate enough time to complete the course.
+Alternatively, a low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
+Learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.
+We investigate this hypothesis in Study 2.
+
 Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents.
-Notably, just under 20\% of respondents in a typical course (median proportion) stopped participating because they learned all they intended to learn.
+Notably, 17\% of respondents in a typical course stopped participating because they learned all they intended to learn.
+This finding resonates with some prior work on attrition in community colleges, where attrition has been interpreted as a sign of success.
+It also supports the observation that progress in the course and satisfaction were only weakly related.
+We investigated the underlying factor structure of the reported reasons for disengaging using an exploratory factor analysis.
+The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
+The following four-factor solution emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines);  second, {\em difficulty} (too advanced, exam too difficult); third, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well, not what I'm looking for, technical difficulties), and fourth, {\em goals \& expectations} (only exploring, learned all I wanted, not challenging, credential not valuable).
+The `late start' item failed to load onto any factor.
+General time issues and difficulty are very interpretable factors, while the other two factors comprise a complex set of reasons that should probably be further separated.
 
-% say how this was done (linear combination of correlation matrices; exluding any? if so, N=?)
-To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis.
-The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
-The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em difficulty} (too advanced, exam too difficult); fourth, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em goals \& expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
-The technical difficulties item failed to load onto any factor.
+Reasons for disengaging from a course are likely to vary by learner characteristics ($RQ3b$).
+To explore if the substantial individual differences in attrition identified above could be associated with different reasons for disengaging, we investigated variation by geographic region ($N = 1,698$) and gender ($N = 1,039$ due to missing gender information).
+Learners in Africa, Asia, and Latin America were more likely to disengage due to technical difficulties (61\% more likely), too demanding deadlines, and starting late ($p\text{s} < .03$), while learners in Europe, Oceania, and Northern America were more likely to disengage because the course was not a good fit for them ($p=.02$).
+Women were more likely than men to report disengaging due to technical difficulties and because the course required too much time ($p<.01$).
+Technical difficulties were a critical individual difference that could be a source of the observed achievement gaps.
+The pre-defined list of reasons for attrition unfortunately did not include an item on language difficulties, which could be another plausible source of the global achievement gap.
 
-% NEED TO SAY THAT IT'S LITTLE DATA HERE (n=?)
-We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course.
-A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations.
-Taking the course with friends or colleagues was correlated with disengaging because the course was not challenging enough, the material was too advanced for the learner's background, and the credential was not valuable enough to the learner ($r=.24, .20, .16$, respectively).
-Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$).
-Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
-
-% other covariates: e.g. do we see geolocation difference for language and tech?
-% also gender differences?
-
-<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
-values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
-57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
-21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
-32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
-15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
-21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
-69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
-12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
-18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
-13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
-16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
-14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
-26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
-14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)
-
-reasons = c("Material not explained properly","Course requires too much time","Not what I was looking for","I joined too late","Didn't enjoy online course format","Technical difficulties","Couldn't keep up with deadlines","Not challenging enough","I learned all I wanted","Videos were confusing","I was only exploring","Too advanced for my background","Assignment or exam was too difficult","Credentials not valuable to me")
-
-dat = data.frame(reasons = rep(reasons, each=17), values)
-
+<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Distribution of self-reported reasons for attrition across 20 MOOCs.'>>=
 library(ggplot2)
-ggplot(dat, aes(reorder(reasons, values, median), values)) + geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion",x="")
+x = read.csv("drop_reasons.csv")
+ggplot(x, aes(reorder(variable, value, median), round(100*value))) + 
+  geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion of learners in the course who selected the reason",x="") + scale_y_continuous(breaks=c(0,10,20,30,40,50,60,70))
 @
-
-\subsection{Discussion}
-% need to adjust once the rest is finalized above
-A number of striking findings emerged from the preceeding analyses.
-We found strong evidence for gender inequalities in attrition and, to some extent, performance.
-Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship.
-The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women.
-Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched.
-We also identified a large achievement gap in terms of grades between learners from these continents.
-The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings.
-It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.
-
-Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end.
-Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations.
-General time issues were overall the most pronounced.
-Attrition in community colleges has been interpreted as a sign of success in prior work.
-This view resonates in the MOOC setting, where around 20\% of learners who disengaged reportedly had achieved their goal, and where progress in the course and learner satisfaction were only weakly related.
-
-Learners may disengage from the course when they become too occupied by higher-priority tasks.
-Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course.
-According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain.
-A low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
-This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.
 
@@
 This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced.
-We invited learners who were likely to drop out of a course to provide feedback.
-These learners were identified with a predictive model and invited to complete a survey.
-The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners.
-Moreover, most learners indicated in Study 1 that they disengaged due to a shortage of time.
+We collected feedback from learners who were likely to have stopped following the course.
+They were identified with a predictive model and invited to complete a survey.
+The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of online learners.
+Moreover, in Study 1, most learners indicated that they disengaged due to a shortage of time.
 A variety of factors could influence the amount of time learners have at their disposal.
 Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors.
-Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}.
-Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV).
-We constructed the following hypotheses about the role of volition in attrition due to a shortage of time:
+Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}.
+Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.).
+We constructed the following hypotheses about the role of volition in time-related attrition:
 
@@
 
-\subsubsection{Feedback Survey}
-
-Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently.
-Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement.
-And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
-Missing values were multiply imputated by predictive mean matching using responses to all survey questions.
-All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
-The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelor's or more advanced degree.
-
-Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked.
-These items served as measures of personal success.
-They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of.
-This question was asked prior to any survey questions that could suggest particular reasons for disengaging.
-
-Two research assistants independently developed codebooks for the resulting 448 non-empty open responses.
-Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations.
-The codebook was updated in each iteration to reliably fit the open response data.
-Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement.
-In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification.
-In the second iteration, the remaining responses were coded in the same way.
-In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see !!!!!!![URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
-
-% Learners also rated the extent to which their progress was hindered by a number of obstacles. 
-Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adpated from \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
-
-\subsection{Results}
-
-We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control.
-In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
-
-% \subsubsection{Predicting Likely Dropouts}
-
-% SHERIF TODO: Add results of prediction.
-How well did the prediction work?
-And who took the survey; how sure was the model that those who took the survey would actually drop out, relative to those who didn't take the survey?
-
-\subsubsection{Psychological Factors}
-
 \begin{table*}[ht]
@@
 
+\subsubsection{Feedback Survey}
+
+Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently.
+Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement.
+And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
+Missing values were multiply imputated by predictive mean matching using responses to all survey questions.
+All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
+The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelor's or more advanced degree.
+
+Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked.
+These items served as measures of personal success.
+They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of.
+This question was asked prior to any survey questions that could suggest particular reasons for disengaging.
+
+Two research assistants independently developed codebooks for the resulting 448 non-empty open responses.
+Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations.
+The codebook was updated in each iteration to reliably fit the open response data.
+Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement.
+In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification.
+In the second iteration, the remaining responses were coded in the same way.
+In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see http://kizilcec.com/rsc/las2015attrition.pdf.} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
+
+Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
+
+\subsection{Results and Discussion}
+
+We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control.
+In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
+
+\subsubsection{Psychological Factors}
+
 Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
@@
 
+Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset.
+Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance.
+The current observational findings highlight the critical role of these factors in an online learning environment.
+Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$).
+An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging.
+These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues.
+Pschological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.
+
+The methodology used in this study can be extended to deliver timely and targeted interventions.
+The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course.
+The majority of them were unsuccessful in their own terms.
+All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners.
+Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).
+Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
+Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their social belonging.
+
 \subsubsection{Reasons for Dropout}
@@
 Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it.
+This list of reasons maps partially onto the factors identified in Study 1 (general time issues, difficulty, format \& content, and goals \& expectations).
 A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).
@@
 These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
+
+The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$).
+A learner needs to be highly self-regulated to be successful in an open learning environment, such as a MOOC.
+Very few learners explicitly indicated issues with volitional control in their open responses.
+They may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait.
+Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$).
+This could be a useful proxy for identifying issues with volitional control in future work.
+Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.
 
@@
 
-\subsection{Discussion}
-
-In this follow-up study, we examined the role of psychological factors in attrition and further investigated reasons for attrition.
-Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset.
-Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance.
-The current observational findings highlight the critical role of these factors in an online learning environment.
-Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$).
-An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging.
-These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues.
-Pschological factors could thereby, at least partially, account for individual differences in persistence and performance (Study 1).
-
-The methodology used in this study can be extended to deliver timely and targeted interventions.
-The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course.
-The majority of them were unsuccessful in their own terms.
-All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners.
-Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).
-Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
-Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their social belonging.
-
-Through iterative coding, we arrived at a list of challenges faced by a subset of learners in a particular MOOC (Table \ref{tab:s2reas}).
-The list maps onto the factors identified in Study 1 (general time issues, late start, difficulty, format \& content, expectations, and technical difficulties), with the additional challenge of a language barrier, which was not included as a response option in Study 1.
-The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$).
-A learner needs to be highly self-regulated to be successful in a MOOC.
-Very few learners explicitly indicated issues with volitional control in their open responses.
-Learners may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait.
-Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$).
-This could be a useful proxy for identifying issues with volitional control in future work.
-Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.
-
 \section{General Discussion}
 
-This study of over 60,000 online learners, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments.
-The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners with different demographic features and from different geographic locations; \emph{second}, the relation between personal success in online learning (by learners' own judgment) and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which was frequently related to learners' low capacity for volitional control.
+This study of over 100,000 online learners in 21 courses, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments.
+The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and from different geographic locations; \emph{second}, the relation between personal success in online learning (by learners' own judgment) and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which was frequently related to learners' low capacity for volitional control.
 Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf.
@@
 \section{Acknowledgments}
-We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for helping with the development of the codebook.
+We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for their assistance developing a codebook.
 The authors contributed equally to this work.

### 2015-01-24 paper_v2.tex
updated version

@@
 
-Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (cf.
-Table \ref{tab:s1sum}).
-For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
-% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
-On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course.
-On the feedback survey, learners reported their satisfaction with the course (6-item index, range from 1 to 7, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (Figure \ref{fig:s1reason} lists reasons).\footnote{Instead of a select-all-that-apply question, learners were encouraged to consider each reason in turn by using binary-choice items.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to mitigate acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
+Twenty MOOCs on topics in a variety of disciplines were selected for investigation.
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
+For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course.
+On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf.
+Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
+Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
 
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
-Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
-A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 17 MOOCs in our dataset.
-Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$.
-Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{Multiple imputation and subsequent pooled analyses were performed with the $R$ mice package.}
-% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
-% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
-% Information about age, gender, and education level were only available for learners who completed the initial course survey.
-% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior
+To analyze individual differences, we fit a logistic mixed-effects (hierarchical) model predicting binary persistence and performance milestones using learner characteristics, including demographic and geographic features.
+The analysis included 67,333 learners who completed the initial course survey, but excluded 381 learners who left all questions unanswered and learners in courses C6, C7, C19, C20 due to missing survey items.
+Any remaining missing survey responses were multiply imputed using other survey responses.
+We report pooled estimates that account for added uncertainty.\footnote{Multiple imputation ($m=10$) using the $R$ mice package.} Correlations between predictors were low ($|r|<0.15$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
+Courses were modeled as a random effect (instead of including dummy variables) to generalize to a larger population of similar MOOCs.
 
-\subsection{Results}
+\subsection{Results and Discussion}
 
-We address the research questions developed in the previous section in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across subsets of learners ($RQ2ab$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3ab$).
+We address the posed research questions in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across learner subgroups ($RQ2$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3$).
 
-\subsubsection{Individual Differences in Attrition}
+\subsubsection{Individual Differences in Attrition and Performance}
 
-We fit a logistic mixed-effects (hierarchical) model to predict multiple binary persistence and performance metrics (milestones) to estimate individual differences in learner characteristics, including demographic and geographic features.
-Correlations between predictors were low ($|r|<0.16$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
-The coefficient on education, for instance, reflects the association between having a bachelor's degree and achieving a milestone, partialing out all other predictors, including age.
-Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{Data for 3 of the 17 MOOCs was unavailable due to anonymization.}
-
-Figure \ref{fig:s1coefs} presents transformed regression coefficients with 95\% confidence intervals from each model.
+Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone.
 The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population.
-The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average aged male learners in North America without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
-In this sample of survey respondents, the baseline probability for each milestone was high: videos watched (27\% above 30\%, 16\% above 50\%, 8\% above 80\%) assessments attempted (78\% above 30\%, 55\% above 50\%, 33\% above 80\%), and grade (42\% above the 60$^\text{th}$, 31\% above the 80$^\text{th}$ percentile).
+The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
+The baseline probability for each milestone was relatively high: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).
 
 A number of significant and substantial individual differences emerged.
-Women were 10 to 15\% less likely than men to persist with lectures and assessments.
-Women, who constituted 41\% of learners in the sample, were also less likely to score a grade above the 60$^\text{th}$ percentile than men.
-Learners in Africa ($n = 2,097$), Asia ($n = 8,199$), and South America ($n = 2,574$) were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America ($n = 18,353$).
-In contrast, learners in Europe ($n = 11,992$) exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America.
-While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America.
-Learners in Oceania ($n = 1,166$) performed similarly to those in North America and Europe.
+Women were 12 to 20\% less likely than men to persist with lectures and assessments.
+Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.
+Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$).
+In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America.
+While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America.
+Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe.
+Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners.
+More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile.
+Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
+Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.
 
-Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance.
-More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to score highly on them.
-Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
-Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.
+The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning.
+Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America.
+There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates.
+The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model.
+The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings.
+The present work cannot move beyond presenting evidence for the existence of these gaps and discussing potential causes.
+Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps; but rather, gender and location are associated with countless individual characteristics, some of which impact academic achievement.
+The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}.
+More work is needed to identify potential causal mechanisms underlying these gaps.
 
@@
 
-\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Estimates from logistic mixed-effects model predicting persistence and performance milestones]{Estimates from logistic mixed-effects model predicting persistence and performance milestones.
+\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Individual differnces in persistence and course grades showing substantial achievement gaps]{Individual differnces in persistence and course grades showing substantial achievement gaps.
 An estimate of -10 indicates a 10\% lower chance of achieving a milestone.\label{fig:s1coefs}}
@@
 
-\subsubsection{Learner Satisfaction}
+\subsubsection{Satisfaction}
 % RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
@@
 
-Attrition can be a sign of success if a learner has achieved their personal goal.
-We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$).
-Satisfaction ratings were compared with a mixed-effects regression to model differences across courses and generalize to a larger population of MOOCs.
-Learners who reported that they stopped participating were 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$).
+Attrition can be a sign of success if a learner has achieved their personal learning objectives.
+We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who persisted to different extents ($RQ2b$).
+Satisfaction ratings were compared using mixed-effects regression ($N=23,390$).
+Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$).
 Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied.
-While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=14, p<.001$), the effect was relatively weak.
-Those who watched over 75\% of videos were only 5\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=4.97, M_{>\text{75\%}}=5.22$).
-
-% We also examined differences in satisfaction between learners grouped by certain behaviors.
-First, we classified learners into 4 groups based on the fraction of course videos they viewed.
-The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}.
-The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group.
-Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.
+The proportion of videos significantly predicted higher satisfaction ($t_{101}=20, p<.001$), such that those who watched over 75\% of videos were 8\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=3.80, M_{>\text{75\%}}=4.09$).
+As levels of self-reported satisfaction were generally high, it remains unclear if the estimated differences in satisfaction are consequential.
+Learners who disengaged early were only somewhat less satisfied than those who persisted in the course, which suggests that some learners arrived with modest course goals and achieved them.
+The following analysis of reasons for attrition can shed more light on this hypothesis.
 
@@
 
-% Say how many actually reported N=?
-Learners' reasons for disengaging from a MOOC were examined using survey responses ($RQ3a$).
-Across 17 MOOCs, learners selected reasons for disengaging from a pre-defined list.
-This list may not be collectively exhaustive, which is addressed in Study 2.
-The distribution of chosen reasons is illustrated in Figure \ref{fig:s1reason}.
+A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a pre-defined list ($RQ3a$).
+This list may not be collectively exhaustive, which is addressed in Study 2, but responses were collected from 1,698 learners in 20 different courses.
+Figure \ref{fig:s1reason} illustrates how frequently each reason was selected across courses.
 Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating.
-Learners faced issues keeping up with deadlines and found that the course demanded too much time.
+In a typical course (median proportion), 66\% faced issues keeping up with deadlines and for 46\% the course demanded too much time.
+According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gains.
+The shortage of time could be due to an abundance of higher-priority tasks, as enrolling in a MOOC neither guarantees that a learner can nor intends to allocate enough time to complete the course.
+Alternatively, a low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
+Learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.
+We investigate this hypothesis in Study 2.
+
 Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents.
-Notably, just under 20\% of respondents in a typical course (median proportion) stopped participating because they learned all they intended to learn.
+Notably, 17\% of respondents in a typical course stopped participating because they learned all they intended to learn.
+This finding resonates with some prior work on attrition in community colleges, where attrition has been interpreted as a sign of success.
+It also supports the observation that progress in the course and satisfaction were only weakly related.
+We investigated the underlying factor structure of the reported reasons for disengaging using an exploratory factor analysis.
+The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
+The following four-factor solution emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines);  second, {\em difficulty} (too advanced, exam too difficult); third, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well, not what I'm looking for, technical difficulties), and fourth, {\em goals \& expectations} (only exploring, learned all I wanted, not challenging, credential not valuable).
+The `late start' item failed to load onto any factor.
+General time issues and difficulty are very interpretable factors, while the other two factors comprise a complex set of reasons that should probably be further separated.
 
-% say how this was done (linear combination of correlation matrices; exluding any? if so, N=?)
-To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis.
-The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot.
-The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em difficulty} (too advanced, exam too difficult); fourth, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em goals \& expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable).
-The technical difficulties item failed to load onto any factor.
-
-% NEED TO SAY THAT IT'S LITTLE DATA HERE (n=?)
-We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course.
-A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations.
-Taking the course with friends or colleagues was correlated with disengaging because the course was not challenging enough, the material was too advanced for the learner's background, and the credential was not valuable enough to the learner ($r=.24, .20, .16$, respectively).
-Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$).
-Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).
-
-% other covariates: e.g. do we see geolocation difference for language and tech?
-% also gender differences?
+Reasons for disengaging from a course are likely to vary by learner characteristics ($RQ3b$).
+To explore if the substantial individual differences in attrition identified above could be associated with different reasons for disengaging, we investigated variation by geographic region ($N = 1,698$) and gender ($N = 1,039$ due to missing gender information).
+Learners in Africa, Asia, and Latin America were more likely to disengage due to technical difficulties (61\% more likely), too demanding deadlines, and starting late ($p\text{s} < .03$), while learners in Europe, Oceania, and Northern America were more likely to disengage because the course was not a good fit for them ($p=.02$).
+Women were more likely than men to report disengaging due to technical difficulties and because the course required too much time ($p<.01$).
+Technical difficulties were a critical individual difference that could be a source of the observed achievement gaps.
+The pre-defined list of reasons for attrition unfortunately did not include an item on language difficulties, which could be another plausible source of the global achievement gap.
 
@@
 
-\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
+\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Distribution of self-reported reasons for attrition across 20 MOOCs]{Distribution of self-reported reasons for attrition across 20 MOOCs.\label{fig:s1reason}}
 \end{figure*}
@@
 \end{knitrout}
-
-\subsection{Discussion}
-% need to adjust once the rest is finalized above
-A number of striking findings emerged from the preceeding analyses.
-We found strong evidence for gender inequalities in attrition and, to some extent, performance.
-Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship.
-The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women.
-Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched.
-We also identified a large achievement gap in terms of grades between learners from these continents.
-The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings.
-It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.
-
-Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end.
-Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations.
-General time issues were overall the most pronounced.
-Attrition in community colleges has been interpreted as a sign of success in prior work.
-This view resonates in the MOOC setting, where around 20\% of learners who disengaged reportedly had achieved their goal, and where progress in the course and learner satisfaction were only weakly related.
-
-Learners may disengage from the course when they become too occupied by higher-priority tasks.
-Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course.
-According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain.
-A low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task.
-This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.
 

### 2015-01-24 paper_v3.Rnw
updated version

--- 2015-01-24 paper_v2.Rnw
@@
 
-% Use this command to override the default ACM copyright statement (e.g. for preprints). 
-% Consult the conference website for the camera-ready copyright statement.
-
-%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-\toappear{
-Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+\toappear{\scriptsize Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
 Copyrights for components of this work owned by others than ACM must be honored.
@@
 \\
-{\emph{L@S'15}}, March 14--15, 2015, Vancouver, Canada. \\
-Copyright \copyright~2015 ACM ISBN/14/04...\$15.00. \\
-DOI string from ACM form confirmation}
-%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-
-% Arabic page numbers for submission. 
-% Remove this line to eliminate page numbers for the camera ready copy
-\pagenumbering{arabic}
+{\emph{L@S 2015}}, March 14--18, 2015, Vancouver, BC, Canada. \\
+Copyright is held by the owner/author(s). Publication rights licensed to ACM. \\
+ACM 978-1-4503-3411-2/15/03…\$15.00 \\
+http://dx.doi.org/10.1145/2724660.2724680}
 
@@
 \hypersetup{
-pdftitle={Attrition in Massive Open Online Courses},
+pdftitle={Attrition and Achievement Gaps in Online Learning},
 pdfauthor={Rene F. Kizilcec, Sherif Halawa},
-pdfkeywords={SIGCHI, proceedings, archival format},
+pdfkeywords={Online education; massive open online course; MOOC; persistence; dropout; individual differences; inequality; volition; social belonging; growth mindset; goal striving},
 bookmarksnumbered,
@@
 \newcommand\tabhead[1]{\small\textbf{#1}}
+
+\clubpenalty=9996
+\widowpenalty=8500
 
@@
 Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments.
-This has prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
-A systematic analysis of individual differences in attrition in 20 diverse massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap.
-Women and learners in Africa, Asia, and Latin America scored lower grades and persisted less than men and learners in Europe, Oceania, and Northern America.
-More persistent learners were, however, only marginally more satisfied with their achievement.
-Successful online learners who left the course early reported higher levels of goal striving, feelings of social belonging, and growth mindset than unsuccessful ones ($N=756$).
-The shortage of time was the primary obstacle for most learners, and many were hindered by insufficient volitional control.
-A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
+A systematic analysis of individual differences in attrition and grades in 20 massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap.
+Online learners in Africa, Asia, and Latin America scored substantially lower grades and were up to half as likely to persist than those in Europe, Oceania, and Northern America.
+Women also exhibited lower persistence and performance then men.
+Yet more persistent learners were only marginally more satisfied with their achievement.
+The primary obstacle for most learners was finding time for the course, which was partly related to low levels of volitional control.
+Self-ascribed successful learners reported higher levels of goal striving, growth mindset, and feelings of social belonging than unsuccessful ones ($N=756$).
+Insights into why learners leave online courses informs models of attrition and targeted interventions to support learners achieve their goals.
 \end{abstract}
 
-\keywords{Massive open online course, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}
+\keywords{Online education; massive open online course; MOOC; persistence; dropout; individual differences; self-regulation; volition; social belonging; growth mindset; goal striving}
 
+\category{K.3.1.}{Computers and Education}{Computer Uses in Education}
 \category{H.5.m.}{Information Interfaces and Presentation (e.g. HCI)}{Miscellaneous}
-\category{K.3.1.}{Computers and Education}{Computer Uses in Education}
 
@@
 
-Educational environments have become increasingly diverse.
-Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit.
+Educational environments have become increasingly diverse in format.
+Traditional schools and universities have a characteristically rigid structure, including instructor-defined syllabi, fixed time schedules, entry requirements, and material costs to enter and exit.
 Novel institutional structures have been developed to overcome particular constraints.
@@
 And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}.
-Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} resemble those on online media platforms.
-This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.
+Many of the prototypical behaviors observed in MOOCs resemble those on online media platforms \cite{kizilcec2013deconstructing,breslow2013studying}.
+This trend is also reflected in the large spectrum of learners' motivations for enrolling in MOOCs \cite{kizilcec2015motivation}.
 
@@
 Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested.
-Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized.
+Perspectives on attrition in open online courses ultimately depend on how these courses are viewed.
 For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}.
@@
 Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance.
-For Study 2, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback.
+For Study 2, we invited around 6,053 learners who were predicted to have disengaged from a MOOC to provide feedback.
 We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time.
@@
 \begin{description}
-  \item[RQ1] How do prior characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from the course?
+  \item[RQ1] How are prior characteristics (demographics, geographic location, intentions, and prior experience) related to learners' likelihood of disengaging from the course?
   \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
@@
 
-Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration.
-However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner.
+Distance education, relative to traditional in-person education, tends to attract a more mature student demographic and provide fewer opportunities for social integration.
+Students in distance learning programs are more likely to lead social lives outside of school, maybe working part-time and living with their partner.
 Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students.
@@
 \end{description}
-
-\section{Study 1: Attrition \& Satisfaction}
-
-The present study extends the existing literature on attrition with a large-scale quantitative treatment of attrition in MOOCs.
-We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners.
-This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggeling learners.
-
-\subsection{Methods}
-
-Twenty MOOCs on topics in a variety of disciplines were selected for investigation.
-Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
-For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
-On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course.
-On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
-Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf.
-Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately.
-The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
-Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
-
-To analyze individual differences, we fit a logistic mixed-effects (hierarchical) model predicting binary persistence and performance milestones using learner characteristics, including demographic and geographic features.
-The analysis included 67,333 learners who completed the initial course survey, but excluded 381 learners who left all questions unanswered and learners in courses C6, C7, C19, C20 due to missing survey items.
-Any remaining missing survey responses were multiply imputed using other survey responses.
-We report pooled estimates that account for added uncertainty.\footnote{Multiple imputation ($m=10$) using the $R$ mice package.} Correlations between predictors were low ($|r|<0.15$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
-Courses were modeled as a random effect (instead of including dummy variables) to generalize to a larger population of similar MOOCs.
-
-\subsection{Results and Discussion}
-
-We address the posed research questions in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across learner subgroups ($RQ2$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3$).
-
-\subsubsection{Individual Differences in Attrition and Performance}
-
-Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone.
-The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population.
-The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
-The baseline probability for each milestone was relatively high: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).
-
-A number of significant and substantial individual differences emerged.
-Women were 12 to 20\% less likely than men to persist with lectures and assessments.
-Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.
-Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$).
-In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America.
-While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America.
-Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe.
-Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners.
-More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile.
-Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
-Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.
-
-The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning.
-Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America.
-There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates.
-The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model.
-The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings.
-The present work cannot move beyond presenting evidence for the existence of these gaps and discussing potential causes.
-Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps; but rather, gender and location are associated with countless individual characteristics, some of which impact academic achievement.
-The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}.
-More work is needed to identify potential causal mechanisms underlying these gaps.
 
@@
 # 106 -1.51 > 80th percentile grade      22
+@
+
+\section{Study 1: Attrition \& Satisfaction}
+
+The present study extends the existing literature on attrition with a large-scale quantitative treatment of attrition in MOOCs.
+We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners.
+This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggeling learners.
+
+\subsection{Methods}
+
+Twenty MOOCs on topics in a variety of disciplines were selected for investigation.
+Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course.
+For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
+On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course.
+On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
+Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf.
+Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately.
+The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
+Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.
+
+To analyze individual differences, we fit a logistic mixed-effects (hierarchical) model predicting binary persistence and performance milestones using learner characteristics, including demographic and geographic features.
+The analysis included 67,333 learners who completed the initial course survey, but excluded 381 learners who left all questions unanswered and learners in courses C6, C7, C19, C20 due to missing survey items.
+Any remaining missing survey responses were multiply imputed using other survey responses.
+We report pooled estimates that account for added uncertainty.\footnote{Multiple imputation ($m=10$) using the $R$ mice package.} Correlations between predictors were low ($|r|<0.15$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded.
+Courses were modeled as a random effect (instead of including dummy variables) to generalize to a larger population of similar MOOCs.
+
+\subsection{Results and Discussion}
+
+We address the posed research questions in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across learner subgroups ($RQ2$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3$).
+
+\subsubsection{Individual Differences in Attrition and Performance}
+
+Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone.
+The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population.
+The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it.
+This choice notably corresponds to the dominant power structure narrative, but is solely statistically motivated.
+The baseline probability for each milestone was as follows: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).
+
+A number of significant and substantial individual differences emerged.
+Women were 12 to 20\% less likely than men to persist with lectures and assessments.
+Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.
+Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$).
+In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America.
+While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America.
+Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe.
+Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners.
+More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile.
+Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
+Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.
+
+The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning.
+Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America.
+There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates.
+The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model.
+The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings.
+The present work presents evidence for the existence of these gaps, but cannot explain why they occur.
+Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps.
+The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}.
+A geographical categorization to compare countries with a high and low propensity of English speakers could be misleading, due to confounding with psychological barriers that are less likely to hinder learners in English-speaking countries.
+More work is needed to identify potential causal mechanisms underlying these gaps.
+
+<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Distribution of self-reported reasons for attrition across 20 MOOCs.'>>=
+library(ggplot2)
+x = read.csv("drop_reasons.csv")
+ggplot(x, aes(reorder(variable, value, median), round(100*value))) + 
+  geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion of learners in the course who selected the reason",x="") + scale_y_continuous(breaks=c(0,10,20,30,40,50,60,70))
 @
@@
 
-<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Distribution of self-reported reasons for attrition across 20 MOOCs.'>>=
-library(ggplot2)
-x = read.csv("drop_reasons.csv")
-ggplot(x, aes(reorder(variable, value, median), round(100*value))) + 
-  geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion of learners in the course who selected the reason",x="") + scale_y_continuous(breaks=c(0,10,20,30,40,50,60,70))
-@
-
 \section{Study 2: Understanding Attrition}
@@
 Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors.
-Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}.
+Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learner's volitional control to engage in self-regulated learning \cite{corno2001volitional}.
 Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.).
+In contrast to attrition due to important obligations, there exist techniques to help learners become more self-regulated, which could boost their persistence.
 We constructed the following hypotheses about the role of volition in time-related attrition:
@@
 \begin{table*}[ht]
-\caption{Descriptive and inferrential statistics for psychological measures by learner success}
+\caption{Descriptive and inferrential statistics for psychological measures by self-ascribed success}
 \label{tab:psych}
-\small
 \center
@@
 
-Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).
+Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item on the extent to which lower-priority distractions have hindered the learner's course progress; $M=2.73, SD=1.29$).
 
@@
 In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
-
+\\
 \subsubsection{Psychological Factors}
 
-Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners.
-Success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
+Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-ascribed successful with unsuccessful learners.
+Self-ascribed success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
 The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
@@
 Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
-Social belonging was significantly higher for learners who were successful in terms of satisfaction, but those who were successful in terms of progress.
+Social belonging was significantly higher for learners who were successful in terms of satisfaction, but not those who were successful in terms of progress.
 The pattern was reversed for growth mindset, such that learners who were successful in terms of progress (but not satisfaction) exhibited a stronger growth mindset.
 This pattern could reflect a difference in the underlying process by which these psychological factors affect learner behavior and perception.
-The observed differences are expected to be larger in the full population of learners, compared to the small, homogenous sample in this study, consisting of learners unlikely to persist.
+The observed differences are expected to be larger in the full population of learners, compared to the homogenous sample in this study, consisting of learners who most likely left the course.
 
@@
 An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging.
-These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues.
+These psychological factors tend to vary across people based on prior experiences and they can change as a result of contextual cues.
 Pschological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.
@@
 All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners.
-Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).
+Recent work on psychological interventions showed lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).
 Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
-Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their social belonging.
+Struggeling learners can be identified by leveraging predictive models of persistence.
+However, it requires explanatory models to decide which type of intervention should be delivered to support learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their sense of belonging in the course.
+More research is needed to construct such explanatory models.
 
@@
 The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course.
-While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e.
-exhibit lower volitional control).
+While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (a sign of lower volitional control).
 These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).
@@
 Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
-We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e.
+We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} While the list of reasons is certainly not collectively exhaustive, it was based on frequently mentioned terms in the open responses.
+Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported significantly lower levels of volitional control, i.e.
 their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
 However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
-These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
+These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion appeared to be hindered by low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.
 
 The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$).
-A learner needs to be highly self-regulated to be successful in an open learning environment, such as a MOOC.
-Very few learners explicitly indicated issues with volitional control in their open responses.
-They may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait.
-Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$).
-This could be a useful proxy for identifying issues with volitional control in future work.
-Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.
+A learner needs to be highly self-regulated to be successful in open learning environment that provide little guidance, such as MOOCs.
+Very few learners explicitly indicated issues with self-regulation in their open responses.
+They may not possess the metacognitive skills to consciously notice the issue or they preferred not to admit to this undesirable trait.
+Learners who reported a shortage of time without providing specific reasons were more likely to also indicate insufficient volitional control ($H4b$).
+This could be a useful proxy for identifying issues of self-regulation in future work.
+Overall, the results suggest that additional support and guidance to build stronger self-regulatory skills could help many learners persist in the course for longer.
 
-\begin{table}[h!]
+\begin{table}
 \caption{Iteratively-Coded Reasons for Attrition ($n=399$)}
@@
 This study of over 100,000 online learners in 21 courses, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments.
-The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and from different geographic locations; \emph{second}, the relation between personal success in online learning (by learners' own judgment) and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which was frequently related to learners' low capacity for volitional control.
+The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and learners from different geographic locations; \emph{second}, the relation between self-ascribed success in online learning and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which appeared frequently related to learners' low capacity for volitional control.
 Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf.
@@
 Tinto's \cite{tinto1975dropout} classic model of attrition emphasizes the role of academic and social integration, and incorporates individual learner characteristics as contributors to learners' decision to persist in a course of study.
-Our findings provide strong support for the critical role of individual characteristics, such as gender, age, intentions, experiences, and geographic location.
-Furthermore, we found evidence for the importance of social integration in terms of perceived social belonging \cite{walton2007question}.
+The current work provides evidence for the critical role of individual characteristics, such as gender, age, intentions, prior experiences, and geographic location.
+Social integration also remains an important factor, but in terms of perceived social belonging \cite{walton2007question}.
 In contrast to Tinto's concept of social and academic integration, which characterizes a general state in an institution, social belonging is a subjective experience based on perceptions and interpretations of environmental cues.
-Yet, this subjective experience is typically shared by individuals with a common identity; for example, female learners in male-dominated fields or black minority learners in college.
-Models of attrition should be updated to acknowledge the potential impact of psychological threats, which can cause achievement gaps that would otherwise be overlooked or misinterpreted.
+Yet, this subjective experience is typically shared by individuals with a common identity, such as female learners in male-dominated fields.
+Models of attrition should be updated to account for the potential impact of psychological barriers, which can cause achievement gaps that would otherwise remain unnoticed or be misinterpreted.
 
-There are many practical implications of this research and the one we would like to emphasize is the need for and feasibility of targeted psychological interventions in online learning environments.
+Among a number of practical implications of this work, it provides new evidence for the need and feasibility of targeted interventions in online learning environments.
 While the current work does not provide explanations for the presence of substantial achievement gaps, we can speculate that similar psychological processes as in traditional educational environments are involved.
-Brief and targeted online interventions (mindset, social belonging, goal striving, volitional control, etc.) can potentially narrow existing achievement gaps in online learning.
-More research on prediction models that identify struggling learners and targeted interventions to support them can make a big difference in online learning.
+Brief and scalable online interventions (mindset, social belonging, goal striving, etc.) could potentially narrow existing achievement gaps in online learning.
+Advances in this area require more research on explanatory models to both identify struggling learners and predict what challenges they may face.
+The increasing digitization of education provides new opportunities to make learning environments more adaptive.
+It is time to abandon the one-size-fits-all model for education and focus on supporting individual learners achieve their goals.
 
@@
 
-% Balancing columns in a ref list is a bit of a pain because you
-% either use a hack like flushend or balance, or manually insert
-% a column break.  http://www.tex.ac.uk/cgi-bin/texfaq2html?label=balance
-% multicols doesn't work because we're already in two-column mode,
-% and flushend isn't awesome, so I choose balance.  See this
-% for more info: http://cs.brown.edu/system/software/latex/doc/balance.pdf
-%
-% Note that in a perfect world balance wants to be in the first
-% column of the last page.
-%
-% If balance doesn't work for you, you can remove that and
-% hard-code a column break into the bbl file right before you
-% submit:
-%
-% http://stackoverflow.com/questions/2149854/how-to-manually-equalize-columns-
-% in-an-ieee-paper-if-using-bibtex
-%
-% Or, just remove \balance and give up on balancing the last page.
-%
 \balance
@@
 \end{document}
-
-% Learners drop out at different times during the course.
-To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis.
-However, learners can develop risk factors at different times in the course.
-The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times.
-Such increased survey load can increase attrition or dissatisfaction in the course.
-
-% An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged.
-This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk).
-Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
-
-% \begin{table*}
-% \label{tab:s1reascor}
-% \caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
-% \small
-% \center
-% \begin{tabular}{lccccccccccccc}
-% \toprule
-%  & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
-% \midrule
-% 1 Assn or exam was too difficult    & 0.17 & 0.28 & 0.30 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.20 & 0.09 & 0.33 & 0.40 & 0.28 \\
-% 2 Couldn't keep up with deadlines   & 1 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
-% 3 Credentials not valuable to me    &   & 1 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
-% 4 Didn't enjoy online course format &   &  & 1 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
-% 5 I was only exploring              &   &  &  & 1 & 0.36 & 0.20 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
-% 6 I learned all I wanted            &   &  &  &  & 1 & 0.12 & 0.31 & 0.28 & 0.0? & 0.12 & 0.16 & 0.16 & 0.09 \\
-% 7 Material not explained properly   &   &  &  &  &  & 1 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
-% 8 Not challenging enough            &   &  &  &  &  &  & 1 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
-% 9 Not what I was looking for        &   &  &  &  &  &  &  & 1 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
-% 10 Course requires too much time    &   &  &  &  &  &  &  &  & 1 & 0.06 & 0.08 & 0.19 & 0.12 \\
-% 11 I joined too late                &   &  &  &  &  &  &  &  &  & 1 & 0.22 & 0.11 & 0.14 \\
-% 12 Technical difficulties           &   &  &  &  &  &  &  &  &  &  & 1 & 0.27 & 0.30 \\
-% 13 Too advanced for my background   &   &  &  &  &  &  &  &  &  &  &  & 1 & 0.33 \\
-% 14 Videos were confusing            &   &  &  &  &  &  &  &  &  &  &  &  & 1 \\
-% \bottomrule
-% \end{tabular}
-% \end{table*}
-
-% [.] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)
-
-% \begin{table}
-% \label{tab:rfd_ei_strongest_correlations}
-% \caption{Strongest correlations between enrollment intentions and reasons for disengaging ($|r| \geq 0.15$)}
-% \center
-% \small
-% \begin{tabular}{llr}
-% \toprule
-%  Enrollment Intention & Reason for Disengaging & $r$ \\
-% \midrule
-% Take with others & Course not challenging enough & 0.24 \\
-% Take with others & Too advanced for my background & 0.20 \\
-% Take with others & Credentials not valuable enough for me & 0.16 \\
-% Take with others & Technical difficulties & 0.15 \\
-% Experience an online course & I was only exploring the course & 0.17 \\
-% Improve my English & Technical difficulties & 0.17 \\
-% Personal growth & Could not keep up with deadlines & 0.16 \\
-% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
-% \bottomrule
-% \end{tabular}
-% \end{table}
-
-% \begin{table*}
-% \label{tab:study1_rfd_ei_corr}
-% \caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
-% \small
-% \center
-% \begin{tabular}{lrrrrrrrrrrrrrr}
-% \toprule
-%  & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
-% \midrule
-% Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
-% Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
-% Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
-% General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
-% Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
-% Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
-% Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
-% Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
-% Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
-% Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
-% Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
-% Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
-% Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
-% Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
-% \bottomrule
-% \end{tabular}
-% \end{table*}
-
-% The two most commonly reported reasons for dropout are related to one's ability to commit time to the course.
-The correlation between the two reasons is $r=0.32$.
-A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.
-Difficulty of an assignment or exam was the third most reported reason for dropout.
-It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).
-The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses.
-This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons.
-[What might we like to study more about respondents who selected this reason?] In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).
-Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting ....
-[TBC] The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''.
-it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$) Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$).
-Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}
-
-% 
-% <<loadcsv, echo=FALSE, warning=FALSE>>=
-% library(bitops)
-% library(RCurl)
-% data <- getURL("https://raw.githubusercontent.com/whynotyet/MOOC_dropout/master/compiled_data_all_courses.csv", ssl.verifypeer=0L, followlocation=1L)
-% a = read.csv(text=data)
-% 
-% # Shift scores for satisfaction index components up to 1-N
-% a$qoe = a$qoe + 4
-% a$satis_amount_learnt = a$satis_amount_learnt + 4
-% a$satis_vs_time_spent = a$satis_vs_time_spent + 4
-% a$satis_time_mgmt = a$satis_time_mgmt + 4
-% a$how_met_expectations = a$how_met_expectations + 3
-% a$how_engaging_material_presentation = a$how_engaging_material_presentation + 1
-% 
-% satM = subset(a, select=c('qoe', 'satis_amount_learnt', 'satis_vs_time_spent', 'how_met_expectations', 'how_engaging_material_presentation'))
-% library(psych)
-% 
-% a$satisfaction = rowMeans(satM, na.rm=TRUE)
-% a$satisfaction[!is.finite(a$satisfaction)] = NA
-% a$sat
-% d=subset(a, select=c('course_id', 'num_videos_watched', 'did_dropout', 'satisfaction'))
-% write.csv(d, "satisfaction.csv", row.names=F)
-% library(plyr)
-% a$did_dropout = revalue(factor(a$did_dropout), c("0"="No", "1"="Yes"))
-% @

### 2015-01-24 paper_v3.tex
updated version

--- 2015-01-24 paper_v3.Rnw
@@
 
-<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Individual differnces in persistence and course grades showing substantial achievement gaps.
-An estimate of -10 indicates a 10\\% lower chance of achieving a milestone.'>>=
-library(ggplot2)
-res = read.csv("pred_coefs2014-1-15.csv")
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
 
-res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
-res$y = factor(res$y, labels=c("> 30% lectures", "> 50% lectures", "> 80% lectures", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))
+\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Individual differnces in persistence and course grades showing substantial achievement gaps]{Individual differnces in persistence and course grades showing substantial achievement gaps.
+An estimate of -10 indicates a 10\% lower chance of achieving a milestone.\label{fig:s1coefs}}
+\end{figure*}
 
-res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:Latin America")))
-res$coef2 = factor(res$coef2, labels=rev(c("Intercept", "Gender: Female", "Age (normalized)", "Education: Bachelor+", "Prior Experience", "Intent to complete", "Hours: 2-4", "Hours: 4+ ", "Prior MOOCs: 1-3", "Prior MOOCs: 4+ ", "Location: Europe", "Location: Oceania", "Location: Africa", "Location: Asia", "Location: Latin America")))
-
-# This is what we submitted
-# ggplot(subset(res, coef2!="Intercept"), 
-#        aes(coef2, 100*est, ymin=100*(est-1.96*se), ymax=100*(est+1.96*se), color=factor(sign))) + 
-#   geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
-#   theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
-#   scale_color_manual(values=c('red','grey30','green3')) +
-#   labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
-
-# This is what it should be
-ggplot(subset(res, coef2!="Intercept"), 
-  aes(coef2, 100*(exp(est)-1), ymin=100*(exp(est-1.96*se)-1), ymax=100*(exp(est+1.96*se)-1), color=factor(sign))) + 
-  geom_hline(yintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
-  theme_minimal() + theme(legend.position="none",
-                          axis.title=element_text(vjust=0),
-                          panel.border=element_rect(fill="transparent",size=.5,color="grey50")) + 
-  scale_color_manual(values=c('black','grey50','black')) +
-  scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60,80)) +
-  labs(y="Predicted change in probability of achieving each milestone (95% CIs)", x="")
-coefs = subset(res, coef2=="Intercept", select=c("est","y"))  # should report these in the text of reference
-coefs$exp_est = round(100*exp(coefs$est)) # estimates seem high
-# 1   -1.63          > 30% lectures      20
-# 16  -2.20          > 50% lectures      11
-# 31  -3.13          > 80% lectures       4
-# 46  -1.00       > 30% assessments      37
-# 61  -1.38       > 50% assessments      25
-# 76  -1.73       > 80% assessments      18
-# 91  -1.21 > 60th percentile grade      30
-# 106 -1.51 > 80th percentile grade      22
-@
+\end{knitrout}
 
@@
 
-<<s1reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=10, fig.cap='Distribution of self-reported reasons for attrition across 20 MOOCs.'>>=
-library(ggplot2)
-x = read.csv("drop_reasons.csv")
-ggplot(x, aes(reorder(variable, value, median), round(100*value))) + 
-  geom_boxplot() + coord_flip() + theme_bw() + labs(y="Proportion of learners in the course who selected the reason",x="") + scale_y_continuous(breaks=c(0,10,20,30,40,50,60,70))
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]
+
+\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Distribution of self-reported reasons for attrition across 20 MOOCs]{Distribution of self-reported reasons for attrition across 20 MOOCs.\label{fig:s1reason}}
+\end{figure*}
+
+\end{knitrout}
 

### 2015-01-25 paper_v3.Rnw
final version

@@
 Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments.
-A systematic analysis of individual differences in attrition and grades in 20 massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap.
-Online learners in Africa, Asia, and Latin America scored substantially lower grades and were up to half as likely to persist than those in Europe, Oceania, and Northern America.
-Women also exhibited lower persistence and performance then men.
+A systematic analysis of individual differences in attrition and performance in 20 massive open online courses ($N>67,000$) revealed a geographic achievement gap and a gender achievement gap.
+Online learners in Africa, Asia, and Latin America scored substantially lower grades and were only half as likely to persist than those in Europe, Oceania, and Northern America.
+Women also exhibited lower persistence and performance than men.
 Yet more persistent learners were only marginally more satisfied with their achievement.
 The primary obstacle for most learners was finding time for the course, which was partly related to low levels of volitional control.
-Self-ascribed successful learners reported higher levels of goal striving, growth mindset, and feelings of social belonging than unsuccessful ones ($N=756$).
-Insights into why learners leave online courses informs models of attrition and targeted interventions to support learners achieve their goals.
+Self-ascribed successful learners reported higher levels of goal striving, growth mindset, and feelings of social belonging than unsuccessful ones.
+Insights into why learners leave online courses inform models of attrition and targeted interventions to support learners achieve their goals.
 \end{abstract}
@@
 Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}.
-Distance learning programs deliver education to remote parts of the world and to people who simply could not attend in-person classes.
+Distance learning programs have delivered education to remote parts of the world and to people who simply could not attend in-person classes.
 Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.
@@
 People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site.
-And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}.
+And among those who enroll and enter the site, a large proportion would only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}.
 Many of the prototypical behaviors observed in MOOCs resemble those on online media platforms \cite{kizilcec2013deconstructing,breslow2013studying}.
@@
 Perspectives on attrition in open online courses ultimately depend on how these courses are viewed.
-For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}.
+For instance, MOOCs can be viewed as bridging two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}.
 This view of MOOC participation clearly adds complexity to interpretations of attrition.
@@
 This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total.
-Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggeling to achieve their learning goals.
+Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggling to achieve their learning goals.
 These learners could benefit from changes to the course or targeted interventions.
@@
 
-The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting.
+The majority of early work on attrition centered around theoretical models of students' decision to persist or drop out of a traditional higher education setting.
 The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course.
-An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
-The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, the extent to which the student engages in goal-directed behavior in the face of distraction \cite{corno1993role}.
+An early model suggested that students' persistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}.
+The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, the extent to which the student engages in goal-directed behavior in the face of distraction \cite{corno2001volitional}.
 Motivation alone is necessary but not sufficient for persistence, because students may fail to sustain efforts in the absence of strong self-regulatory skills.
@@
 The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution.
-Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college.
+Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of individual students' characteristics, prior experiences, and experiences during college.
 While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration.
@@
 \begin{description}
-  \item[RQ1] How are prior characteristics (demographics, geographic location, intentions, and prior experience) related to learners' likelihood of disengaging from the course?
+  \item[RQ1] How are learners' characteristics (demographics, geographic location, intentions, prior experience) related to their persistence and performance in the course?
   \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
@@
 Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (instilling a growth mindset \cite{dweck2006mindset}) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}.
-This highlights the critical role of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.
+This highlights the critical role of mindset and cautions against overinterpreting non-persistence as an indicator of goal achievement.
 
 \begin{description}
-  \item[RQ2a] What proportion of learners who disengaged were satisfied with their achievements in the course?
-  \item[RQ2b] How does learners' level of satisfaction depend on their behavior in the course?
-  \item[H3] Successful learners have a more growth mindset than unsuccessful learners.
+  \item[RQ2] How is learners' satisfaction related to their persistence in the course?
+  \item[H3] Successful learners have a stronger growth mindset than unsuccessful learners.
 \end{description}
@@
 Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.).
-The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
+The significant change from Tinto's original model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
 % (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).
@@
 
-\section{Study 1: Attrition \& Satisfaction}
+\section{Study 1: Individual Differences, Satisfaction, \& Reasons for Attrition}
 
@@
 We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners.
-This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggeling learners.
+This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggling learners.
 
@@
 For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment.
-On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course.
+On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intended to spend on the course.
 On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'.
@@
 Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$).
-In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America.
+In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessments than learners in Northern America.
 While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America.
@@
 More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile.
-Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance.
+Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistence.
 Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.
 
-The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning.
+Leveraging the large scale of MOOCs, this study has revealed a global achievement gap and a gender achievement gap in online learning.
 Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America.
@@
 \subsubsection{Satisfaction}
-% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
-% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?
 
 Attrition can be a sign of success if a learner has achieved their personal learning objectives.
-We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who persisted to different extents ($RQ2b$).
-Satisfaction ratings were compared using mixed-effects regression ($N=23,390$).
+We investigated how satisfied learners were with their achievements for those who reported to have stopped participating and those who persisted to different extents ($RQ2$).
+Satisfaction ratings were compared using mixed-effects regression with multiple imputation of missing values ($N=23,390$).
 Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$).
@@
 \subsubsection{Reasons for Attrition}
-% RQ3a: What reasons do learners report for disengaging from a MOOC?
-% RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
 
-A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a pre-defined list ($RQ3a$).
+A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a predefined list ($RQ3a$).
 This list may not be collectively exhaustive, which is addressed in Study 2, but responses were collected from 1,698 learners in 20 different courses.
@@
 In a typical course (median proportion), 66\% faced issues keeping up with deadlines and for 46\% the course demanded too much time.
-According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gains.
+According to expectancy-value theory \cite{wigfield2000expectancy}, the learner is a rational agent who allocates time to different activities to maximize subjective gains.
 The shortage of time could be due to an abundance of higher-priority tasks, as enrolling in a MOOC neither guarantees that a learner can nor intends to allocate enough time to complete the course.
@@
 Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents.
-Notably, 17\% of respondents in a typical course stopped participating because they learned all they intended to learn.
+Notably, 17\% of respondents in a typical course stopped participating because they had learned all they intended to learn.
 This finding resonates with some prior work on attrition in community colleges, where attrition has been interpreted as a sign of success.
-It also supports the observation that progress in the course and satisfaction were only weakly related.
+It also supports the conclusion that progress in the course and satisfaction were only weakly related.
 We investigated the underlying factor structure of the reported reasons for disengaging using an exploratory factor analysis.
@@
 
-\section{Study 2: Understanding Attrition}
+\section{Study 2: Psychological Factors in Attrition}
 
@@
 A variety of factors could influence the amount of time learners have at their disposal.
-Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors.
+Higher priority obligations, such as professional, academic, and personal commitments are presumably large contributors.
 Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learner's volitional control to engage in self-regulated learning \cite{corno2001volitional}.
@@
 \begin{table*}[ht]
-\caption{Descriptive and inferrential statistics for psychological measures by self-ascribed success}
+\caption{Descriptive and inferential statistics for psychological measures by self-ascribed success}
 \label{tab:psych}
+\small
 \center
@@
 And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate).
-Missing values were multiply imputated by predictive mean matching using responses to all survey questions.
+Missing values were multiply imputed by predictive mean matching using responses to all survey questions.
 All estimates were pooled across ten imputations, with standard errors adjusted for added variance.
@@
 In the second iteration, the remaining responses were coded in the same way.
-In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see http://kizilcec.com/rsc/las2015attrition.pdf.} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
+In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).
+The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.
 
-Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item on the extent to which lower-priority distractions have hindered the learner's course progress; $M=2.73, SD=1.29$).
+Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and the degree of volitional control they exerted (one item on the extent to which lower-priority distractions have hindered the learner's course progress; $M=2.73, SD=1.29$).
 
@@
 
-We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control.
-In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
-\\
+First, we tested hypotheses about the role of psychological factors in the self-ascribed success of online learners.
+In a second set of analyses, we further investigated reasons for leaving a course ($RQ3a$) and the role of volition ($H4$).
+
 \subsubsection{Psychological Factors}
@@
 Self-ascribed success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied).
-The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress.
-Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics.
-Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures.
-Social belonging was significantly higher for learners who were successful in terms of satisfaction, but not those who were successful in terms of progress.
-The pattern was reversed for growth mindset, such that learners who were successful in terms of progress (but not satisfaction) exhibited a stronger growth mindset.
-This pattern could reflect a difference in the underlying process by which these psychological factors affect learner behavior and perception.
-The observed differences are expected to be larger in the full population of learners, compared to the homogenous sample in this study, consisting of learners who most likely left the course.
+The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful or unsuccessful in terms of their progress.
+Table \ref{tab:psych} provides summary statistics for each psychological measure by these two self-ascribed success metrics.
+Successful learners exhibited higher levels of goal striving in general.
+Social belonging was significantly higher only for learners who were also more satisfied.
+The pattern was reversed for growth mindset, which was stronger only for learners who were successful in terms of progress.
+This pattern reflects differences in the underlying process by which these psychological factors shape learner behavior and perception.
+Social belonging relates to feeling comfortable and connected in the learning environment, whereas mindset relates to one's ability to master demanding materials.
+The observed differences would presumably be larger in the full population of learners than in this homogeneous sample of at-risk learners.
 
@@
 These psychological factors tend to vary across people based on prior experiences and they can change as a result of contextual cues.
-Pschological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.
+Psychological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.
 
@@
 Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.
-Struggeling learners can be identified by leveraging predictive models of persistence.
+Struggling learners can be identified by leveraging predictive models of persistence.
 However, it requires explanatory models to decide which type of intervention should be delivered to support learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their sense of belonging in the course.
@@
 
-The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course.
-While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (a sign of lower volitional control).
+The majority of survey respondents (550 of 756; 73\%) reported being at most `somewhat satisfied' with their progress in the course.
+While many of them (66\%) were substantially held back by other commitments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (a sign of lower volitional control).
 These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).
@@
 This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders.
-Coders would have required additional training to avoid reading specific reasons into unspecific resposnes.
+Coders would have required additional training to avoid reading specific reasons into unspecific responses.
 We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} While the list of reasons is certainly not collectively exhaustive, it was based on frequently mentioned terms in the open responses.
 Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported significantly lower levels of volitional control, i.e.
-their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$).
+their progress was more severely hindered by lower-priority distractions ($t_{331}=2.9, p=.003, d=.32$).
 However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$).
@@
 They may not possess the metacognitive skills to consciously notice the issue or they preferred not to admit to this undesirable trait.
-Learners who reported a shortage of time without providing specific reasons were more likely to also indicate insufficient volitional control ($H4b$).
+Learners who reported a shortage of time without providing specific reasons were more likely to admit getting distracted by lower-priority tasks, hinting at insufficient volitional control ($H4b$).
 This could be a useful proxy for identifying issues of self-regulation in future work.
@@
 This study of over 100,000 online learners in 21 courses, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments.
-The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and learners from different geographic locations; \emph{second}, the relation between self-ascribed success in online learning and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which appeared frequently related to learners' low capacity for volitional control.
+The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and learners from different geographic locations; \emph{second}, the relation between self-ascribed success in online learning and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, most importantly having not enough time, which appeared frequently related to low levels of volitional control.
 Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf.
@@
 In line with early models of attrition that emphasized the importance of goal-directed behavior, we found goal striving to be higher in subjectively successful than unsuccessful learners.
-Moreover, volitional control was identified as a major contributor in reported time-related challenges with taking a MOOC.
+Moreover, volitional control was identified as a major contributor to reported time-related challenges with taking a MOOC.
 Tinto's \cite{tinto1975dropout} classic model of attrition emphasizes the role of academic and social integration, and incorporates individual learner characteristics as contributors to learners' decision to persist in a course of study.
@@
 \section{Acknowledgments}
-We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for their assistance developing a codebook.
+We thank Candace Thille, Octavia Zahrt, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for their assistance developing a codebook.
 The authors contributed equally to this work.
